---
title: "Advanced Empirical Research Methods)"
subtitle: MT25 Assignment 1 - Part 2
author: "Candidate Number: 1091806"
date: "2026-03-25"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
    toc_depth: 2
    number_sections: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
---

# Libraries
```{r}
# load libraries

library(mvtnorm)  # Question 1
library(knitr)    # Question 1
```

\newpage
# Question 1

**Model recap**:

$$Y = D + \beta Z + U$$

$$D = \mathbf{1}\{\alpha Z + \epsilon > 0\}$$

$$\begin{pmatrix} U \\ \epsilon \end{pmatrix} \sim N\left(\mathbf{0}, \Sigma\right), \quad \Sigma = \begin{pmatrix} 1 & 0.5 \\ 0.5 & 1 \end{pmatrix}$$

$$Z \sim \text{Bernoulli}(0.5), \quad \text{independent of } (U, \epsilon)$$

Construct Function 1: Data Generating Process (DGP) simulation 
```{r}
# DGP simulation function

simulate_data <- function(n, alpha, beta) {
  
  # Draw the correlated errors, (U, epsilon)
    
    # Construct sigma
    
      Sigma <- matrix(c(1, 0.5, 0.5, 1), nrow = 2)
    
    # Construct errors 
    
      errors <- rmvnorm(n, mean = c(0, 0), sigma = Sigma)
    
    U <- errors[, 1]
    eps <- errors[, 2]
  
  # Draw the instrument Z, independent of errors
 
    Z <- rbinom(n, size = 1, prob = 0.5)
  
  # Generate treatment D from the selection equation
       
    D <- as.integer(alpha * Z + eps > 0)
      
  # Generate outcome Y from the outcome equation    
      
    Y <- D + beta * Z + U
  
  # Return a data frame with named columns D, Z, and Y
      
    data.frame(D = D, Z = Z, Y = Y)
}
```

Construct Function 2: Wald (IV) estimator

$$\hat{\beta}_{IV} = \frac{\bar{Y}_{Z=1} - \bar{Y}_{Z=0}}{\bar{D}_{Z=1} - \bar{D}_{Z=0}}$$
```{r}
# Wald (IV) estimator

iv_estimate <- function(df) {
  
  # Numerator: reduced-form effect of Z on Y
    
    ey1 <- mean(df$Y[df$Z == 1])
    ey0 <- mean(df$Y[df$Z == 0])
  
  # Denominator: first-stage effect of Z on D
  
    ed1 <- mean(df$D[df$Z == 1])
    ed0 <- mean(df$D[df$Z == 0])
      
      
    (ey1 - ey0) / (ed1 - ed0)
    }
```

Test with beta = 0, exclusion restriction holds:
```{r}
# The Wald estimand should converge to the true LATE = 1

# Set seed
  
    set.seed(123)

# Run tests:

  results <- data.frame(
    n     = rep(10^5, 3),
    alpha = c(0.6, 1.5, 0.6),
    beta  = c(0, 0, 1),
    
    IV_estimate = c(
      
    # Test 1: beta=0 (exclusion holds) -> IV should be close to 1
      
      iv_estimate(simulate_data(10^5, alpha = 0.6, beta = 0)),
    
    # Test 2: beta=0 with stronger first stage -> IV still close to 1
    
      iv_estimate(simulate_data(10^5, alpha = 1.5, beta = 0)),
    
    # Test 3: beta=1 (exclusion violated) -> IV biased away from 1
      
      iv_estimate(simulate_data(10^5, alpha = 0.6, beta = 1))
  )
)

# Display results
  
  kable(results, digits = 4, caption = "IV estimates for different parameter values")
```

\newpage
# Question 2

## 2a)

Treatment is assigned by:

$$D = \mathbf{1}\{\alpha Z + \epsilon > 0\}$$
So, conditional on Z, individuals with larger $ϵ$ are more likely to be treated.

Since: 

$$Corr(U,ϵ) = 0.5 > 0$$
Those who select into treatment also tend to have higher values on the outcome error $U$. This means: 

$$ E[U∣D=1] > E[U∣D=0]$$ 

So, **D is endogenous**. OLS of Y on D would be biased upward due to positive selection.

## 2b) 

The individual treatment effect is $Y_1 - Y_0$. From the outcome equation, potential outcomes are:

$$Y_1 = 1 + \beta Z + U, \quad Y_0 = 0 + \beta Z + U$$

So for every individual:

$$Y_1 - Y_0 = 1$$

The treatment effect is **constant and equal to 1**, it does not vary across individuals. There is no distribution to speak of; it is degenerate at 1. Consequently:

$$\text{LATE} = \text{ATE} = \text{ATT} = 1$$


## 2c)

$\beta$ controls the **direct effect** of the instrument $Z$ on the outcome $Y$. 

When $\beta \neq 0$, $Z$ enters the outcome equation directly:

$$Y = D + \beta Z + U$$

This means $Z$ affects $Y$ through a channel other than $D$, which **violates the exclusion restriction**, a key assumption of the LATE model. 

When $\beta = 0$, the instrument only affects $Y$ through its effect on $D$, and the exclusion restriction holds.

## 2d)

$\alpha$ controls the **strength of the first stage**, ie. how much the instrument $Z$ shifts the probability of receiving treatment. 

In the selection equation:

$$D = \mathbf{1}\{\alpha Z + \epsilon > 0\}$$

If $\alpha = 0$, the instrument has no effect on $D$ and there are **no compliers**, so IV estimation breaks down.

A larger $\alpha$ means that $Z = 1$ pushes more individuals over the zero threshold, **increasing the share of compliers** (those whose treatment status changes with $Z$). 

## 2e)

The LATE model requires four assumptions. We check each in turn:

1. **Unconfounded Type:** $Z \sim \text{Bernoulli}(0.5)$, independent of $(U, \epsilon)$. Since compliance type is determined entirely by $\epsilon$, and $Z \perp \epsilon$, **this holds**.

2. **No Defiers / Monotonicity:** Potential treatment status is $D(z) = \mathbf{1}\{\alpha z + \epsilon > 0\}$. For $\alpha > 0$: $D(1) = \mathbf{1}\{\alpha + \epsilon > 0\} \geq \mathbf{1}\{\epsilon > 0\} = D(0)$ for all $\epsilon$, so $D(1) \geq D(0)$ always. No defiers exist, **this holds**.

3. **Existence of Compliers:** Compliers are individuals with $-\alpha < \epsilon \leq 0$, for whom $D(0) = 0$ but $D(1) = 1$. Since $\epsilon \sim N(0,1)$, this set has positive probability whenever $\alpha \neq 0$. **This holds**.

4. **Mean Exclusion:** Potential outcomes are $Y_d = d + \beta Z + U$, which depend on $Z$ directly when $\beta \neq 0$. So this assumption is **violated when $\beta \neq 0$**, and **satisfied only when $\beta = 0$**.

In summary, assumptions 1--3 always hold (for $\alpha > 0$). The exclusion restriction holds only when $\beta = 0$.

\newpage

# Question 3

The key difference between HM2015 and MW2017 lies in the **exclusion restriction**. 

HM2015 requires only **mean exclusion** (their Assumption 2, page 400):

$$E(Y^{d,1} \mid T = t, Z = 1) = E(Y^{d,0} \mid T = t, Z = 0] = E(Y^d \mid T = t) $$

This says the instrument does not affect the *mean* potential outcome within any subpopulation. 

MW2017 requires **full independence** (their Assumption 1, "LATE Independence - LI", page 306):

$$Z \perp (Y_1, Y_0, D_0, D_1)$$

This is strictly stronger: it implies the instrument does not affect the *entire distribution* of potential outcomes, not just the mean. Both papers share the remaining assumptions (unconfounded types and monotonicity).

Because MW2017 imposes the stronger assumption, its testable implications are **sharp**, they capture all observable violations of LI + LM. HM2015's inequalities are necessary but not sufficient conditions for instrument validity, so some violations may go undetected. 

On the other hand, HM2015's weaker requirement means its test remains valid in settings where only mean independence is credible.

Both tests are relevant because in just-identified IV models (one instrument for one endogenous variable), traditional overidentification tests like the Sargan test are unavailable. 













