---
title: "Advanced Empirical Research Methods)"
subtitle: MT25 Assignment 1 - Part 2
author: "Candidate Number: 1091806"
date: "2026-03-25"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
    toc_depth: 2
    number_sections: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
---

# Libraries
```{r}
# load libraries

library(mvtnorm)  # Question 1
library(knitr)    # Question 1, 5
library(modelsummary) # Question 5
library(kableExtra) # Question 5
```

\newpage
# Question 1

**Model recap**:

$$Y = D + \beta Z + U$$

$$D = \mathbf{1}\{\alpha Z + \epsilon > 0\}$$

$$\begin{pmatrix} U \\ \epsilon \end{pmatrix} \sim N\left(\mathbf{0}, \Sigma\right), \quad \Sigma = \begin{pmatrix} 1 & 0.5 \\ 0.5 & 1 \end{pmatrix}$$

$$Z \sim \text{Bernoulli}(0.5), \quad \text{independent of } (U, \epsilon)$$

Construct Function 1: Data Generating Process (DGP) simulation 
```{r}
# DGP simulation function

simulate_data <- function(n, alpha, beta) {
  
  # Draw the correlated errors, (U, epsilon)
    
    # Construct sigma
    
      Sigma <- matrix(c(1, 0.5, 0.5, 1), nrow = 2)
    
    # Construct errors 
    
      errors <- rmvnorm(n, mean = c(0, 0), sigma = Sigma)
    
    U <- errors[, 1]
    eps <- errors[, 2]
  
  # Draw the instrument Z, independent of errors
 
    Z <- rbinom(n, size = 1, prob = 0.5)
  
  # Generate treatment D from the selection equation
       
    D <- as.integer(alpha * Z + eps > 0)
      
  # Generate outcome Y from the outcome equation    
      
    Y <- D + beta * Z + U
  
  # Return a data frame with named columns D, Z, and Y
      
    data.frame(D = D, Z = Z, Y = Y)
}
```

Construct Function 2: Wald (IV) estimator

$$\hat{\beta}_{IV} = \frac{\bar{Y}_{Z=1} - \bar{Y}_{Z=0}}{\bar{D}_{Z=1} - \bar{D}_{Z=0}}$$
```{r}
# Wald (IV) estimator

iv_estimate <- function(df) {
  
  # Numerator: reduced-form effect of Z on Y
    
    ey1 <- mean(df$Y[df$Z == 1])
    ey0 <- mean(df$Y[df$Z == 0])
  
  # Denominator: first-stage effect of Z on D
  
    ed1 <- mean(df$D[df$Z == 1])
    ed0 <- mean(df$D[df$Z == 0])
      
      
    (ey1 - ey0) / (ed1 - ed0)
    }
```

Test with beta = 0, exclusion restriction holds:
```{r}
# The Wald estimand should converge to the true LATE = 1

# Set seed
  
    set.seed(123)

# Run tests:

  results <- data.frame(
    n     = rep(10^5, 3),
    alpha = c(0.6, 1.5, 0.6),
    beta  = c(0, 0, 1),
    
    IV_estimate = c(
      
    # Test 1: beta=0 (exclusion holds) -> IV should be close to 1
      
      iv_estimate(simulate_data(10^5, alpha = 0.6, beta = 0)),
    
    # Test 2: beta=0 with stronger first stage -> IV still close to 1
    
      iv_estimate(simulate_data(10^5, alpha = 1.5, beta = 0)),
    
    # Test 3: beta=1 (exclusion violated) -> IV biased away from 1
      
      iv_estimate(simulate_data(10^5, alpha = 0.6, beta = 1))
  )
)

# Display results
  
  kable(results, digits = 4, caption = "IV estimates for different parameter values")
```

\newpage
# Question 2

## 2a)

Treatment is assigned by:

$$D = \mathbf{1}\{\alpha Z + \epsilon > 0\}$$
So, conditional on Z, individuals with larger $ŽĄ$ are more likely to be treated.

Since: 

$$Corr(U,ŽĄ) = 0.5 > 0$$
Those who select into treatment also tend to have higher values on the outcome error $U$. This means: 

$$ E[UŌłŻD=1]ŅĆĀ > E[UŌłŻD=0]$$ 

So, **D is endogenous**. OLS of Y on D would be biased upward due to positive selection.

## 2b) 

The individual treatment effect is $Y_1 - Y_0$. From the outcome equation, potential outcomes are:

$$Y_1 = 1 + \beta Z + U, \quad Y_0 = 0 + \beta Z + U$$

So for every individual:

$$Y_1 - Y_0 = 1$$

The treatment effect is **constant and equal to 1**, it does not vary across individuals. There is no distribution to speak of; it is degenerate at 1. Consequently:

$$\text{LATE} = \text{ATE} = \text{ATT} = 1$$


## 2c)

$\beta$ controls the **direct effect** of the instrument $Z$ on the outcome $Y$. 

When $\beta \neq 0$, $Z$ enters the outcome equation directly:

$$Y = D + \beta Z + U$$

This means $Z$ affects $Y$ through a channel other than $D$, which **violates the exclusion restriction**, a key assumption of the LATE model. 

When $\beta = 0$, the instrument only affects $Y$ through its effect on $D$, and the exclusion restriction holds.

## 2d)

$\alpha$ controls the **strength of the first stage**, ie. how much the instrument $Z$ shifts the probability of receiving treatment. 

In the selection equation:

$$D = \mathbf{1}\{\alpha Z + \epsilon > 0\}$$

If $\alpha = 0$, the instrument has no effect on $D$ and there are **no compliers**, so IV estimation breaks down.

A larger $\alpha$ means that $Z = 1$ pushes more individuals over the zero threshold, **increasing the share of compliers** (those whose treatment status changes with $Z$). 

## 2e)

The LATE model requires four assumptions. We check each in turn:

1. **Unconfounded Type:** $Z \sim \text{Bernoulli}(0.5)$, independent of $(U, \epsilon)$. Since compliance type is determined entirely by $\epsilon$, and $Z \perp \epsilon$, **this holds**.

2. **No Defiers / Monotonicity:** Potential treatment status is $D(z) = \mathbf{1}\{\alpha z + \epsilon > 0\}$. For $\alpha > 0$: $D(1) = \mathbf{1}\{\alpha + \epsilon > 0\} \geq \mathbf{1}\{\epsilon > 0\} = D(0)$ for all $\epsilon$, so $D(1) \geq D(0)$ always. No defiers exist, **this holds**.

3. **Existence of Compliers:** Compliers are individuals with $-\alpha < \epsilon \leq 0$, for whom $D(0) = 0$ but $D(1) = 1$. Since $\epsilon \sim N(0,1)$, this set has positive probability whenever $\alpha \neq 0$. **This holds**.

4. **Mean Exclusion:** Potential outcomes are $Y_d = d + \beta Z + U$, which depend on $Z$ directly when $\beta \neq 0$. So this assumption is **violated when $\beta \neq 0$**, and **satisfied only when $\beta = 0$**.

In summary, assumptions 1--3 always hold (for $\alpha > 0$). The exclusion restriction holds only when $\beta = 0$.

\newpage

# Question 3

The key difference between HM2015 and MW2017 lies in the **exclusion restriction**. 

HM2015 requires only **mean exclusion** (their Assumption 2, page 400):

$$E(Y^{d,1} \mid T = t, Z = 1) = E(Y^{d,0} \mid T = t, Z = 0] = E(Y^d \mid T = t) $$

This says the instrument does not affect the *mean* potential outcome within any subpopulation. 

MW2017 requires **full independence** (their Assumption 1, "LATE Independence - LI", page 306):

$$Z \perp (Y_1, Y_0, D_0, D_1)$$

This is strictly stronger: it implies the instrument does not affect the *entire distribution* of potential outcomes, not just the mean. Both papers share the remaining assumptions (unconfounded types and monotonicity).

Because MW2017 imposes the stronger assumption, its testable implications are **sharp**, they capture all observable violations of LI + LM. HM2015's inequalities are necessary but not sufficient conditions for instrument validity, so some violations may go undetected. 

On the other hand, HM2015's weaker requirement means its test remains valid in settings where only mean independence is credible.

Both tests are relevant because in just-identified IV models (one instrument for one endogenous variable), traditional overidentification tests like the Sargan test are unavailable. 

\newpage
# Question 4

Recall from HM2015 Equation (7), the four testable inequality moment constraints:

$$\theta_1 = E[Y \mid D=1, Z=1, Y \leq y_q] - E[Y \mid D=1, Z=0]$$
$$\theta_2 = E[Y \mid D=1, Z=0] - E[Y \mid D=1, Z=1, Y \geq y_{1-q}]$$
$$\theta_3 = E[Y \mid D=0, Z=0, Y \leq y_r] - E[Y \mid D=0, Z=1]$$
$$\theta_4 = E[Y \mid D=0, Z=1] - E[Y \mid D=0, Z=0, Y \geq y_{1-r}]$$

where:

$$q = \frac{P(D=1 \mid Z=0)}{P(D=1 \mid Z=1)}$$

is the share of always takers among the treated when $Z=1$, and:

$$r = \frac{P(D=0 \mid Z=1)}{P(D=0 \mid Z=0)}$$

is the share of never takers among the untreated when $Z=0$.

Under IV validity, all four $\theta_j \leq 0$.


```{r}
get_theta_HM <- function(df) {
  
  # Subgroups
  
    Y_D1Z1 <- df$Y[df$D == 1 & df$Z == 1]
    
    Y_D1Z0 <- df$Y[df$D == 1 & df$Z == 0]
    
    Y_D0Z0 <- df$Y[df$D == 0 & df$Z == 0]
    
    Y_D0Z1 <- df$Y[df$D == 0 & df$Z == 1]
  
  # Proportions q and r
    
    q <- mean(df$D[df$Z == 0]) / mean(df$D[df$Z == 1])
    
    r <- mean(1 - df$D[df$Z == 1]) / mean(1 - df$D[df$Z == 0])
  
  # Quantile cutoffs
      
    yq     <- quantile(Y_D1Z1, probs = q)
    
    y1_q   <- quantile(Y_D1Z1, probs = 1 - q)
    
    yr     <- quantile(Y_D0Z0, probs = r)
    
    y1_r   <- quantile(Y_D0Z0, probs = 1 - r)
  
  # Compute the four theta values
      
    theta1 <- mean(Y_D1Z1[Y_D1Z1 <= yq])   - mean(Y_D1Z0)
    
    theta2 <- mean(Y_D1Z0)                  - mean(Y_D1Z1[Y_D1Z1 >= y1_q])
    
    theta3 <- mean(Y_D0Z0[Y_D0Z0 <= yr])   - mean(Y_D0Z1)
    
    theta4 <- mean(Y_D0Z1)                  - mean(Y_D0Z0[Y_D0Z0 >= y1_r])
  
    
  c(theta1 = theta1, theta2 = theta2, theta3 = theta3, theta4 = theta4)
}
```


\newpage
# Question 5

## Case 1: $\alpha = 0.6$, $\beta = 1$

Generate 100,000 observations with the exclusion restriction violated ($\beta = 1$):

```{r}
set.seed(123)
df_beta1 <- simulate_data(n = 100000, alpha = 0.6, beta = 1)
```

Compute the HM2015 inequality constraints:

```{r}
theta_beta1 <- get_theta_HM(df_beta1)
kable(
  data.frame(
    Theta = names(theta_beta1),
    Value = round(theta_beta1, 4),
    Violated = theta_beta1 > 0
  ),
  caption = "HM2015 Inequalities (alpha = 0.6, beta = 1)",
  booktabs = TRUE,
  row.names = FALSE
) %>% kable_styling(latex_options = "hold_position")
```

We detect a violation: at least one $\theta_j > 0$, meaning the point-identified mean outcome of the always takers (or never takers) falls outside its bounds. This is expected since $\beta = 1$ means $Z$ directly enters the outcome equation, violating the exclusion restriction.

Now compute the Wald estimand and compare to the true LATE. Recall the Function 2 we constructed in Question 1:

$$\hat{\beta}_{IV} = \frac{\bar{Y}_{Z=1} - \bar{Y}_{Z=0}}{\bar{D}_{Z=1} - \bar{D}_{Z=0}}$$

```{r, results='hide'}
wald_beta1 <- iv_estimate(df_beta1)
cat("Wald estimand:", round(wald_beta1, 4), "\n")
cat("True LATE:     1\n")
cat("Bias:         ", round(wald_beta1 - 1, 4), "\n")
```

| $\beta$ | Wald Estimand | Bias |
| --- | --- | --- |
| 1 | `r round(wald_beta1, 4)` | `r round(wald_beta1 - 1, 4)` |

The Wald estimand is substantially larger than 1. The bias arises because the numerator of the Wald estimand captures both the treatment effect of $D$ and the direct effect $\beta Z$, inflating the estimate.

## Case 2: $\alpha = 0.6$, $\beta = 0$

Repeat with the exclusion restriction satisfied ($\beta = 0$):

```{r}
set.seed(123)
df_beta0 <- simulate_data(n = 100000, alpha = 0.6, beta = 0)
```

Compute the HM2015 inequality constraints:

```{r}
theta_beta0 <- get_theta_HM(df_beta0)
kable(
  data.frame(
    Theta = names(theta_beta0),
    Value = round(theta_beta0, 4),
    Violated = theta_beta0 > 0
  ),
  caption = "HM2015 Inequalities (alpha = 0.6, beta = 0)",
  row.names = FALSE
) %>% kable_styling(latex_options = "hold_position")
```

All four $\theta_j \leq 0$: no violation detected, which is consistent with the fact that all LATE assumptions hold when $\beta = 0$.

Compute the Wald estimand:

```{r, results='hide'}
wald_beta0 <- iv_estimate(df_beta0)
cat("Wald estimand:", round(wald_beta0, 4), "\n")
cat("True LATE:     1\n")
cat("Bias:         ", round(wald_beta0 - 1, 4), "\n")
```

The Wald estimand is close to 1 with negligible bias, confirming it correctly identifies the LATE when the exclusion restriction holds. Combining with the results from Q2e, all four LATE assumptions are satisfied when $\beta = 0$, so the Wald estimand can be interpreted as the LATE.


| $\beta$ | Wald Estimand | Bias |
| --- | --- | --- |
| 0 | `r round(wald_beta0, 4)` | `r round(wald_beta0 - 1, 4)` |


\newpage

# Question 6

The exclusion restriction is violated for any $\beta \neq 0$. However, the HM2015 inequalities are only *necessary* conditions for instrument validity ŌĆö they may fail to detect small violations. We now search for a value of $\beta$ where all four $\theta_j \leq 0$ despite $\beta \neq 0$.

We loop over a grid of $\beta$ values from 0.05 to 1. For each value, we simulate 100,000 observations, compute the HM2015 inequalities, and record whether any violation is detected:

```{r}

# Set seed

    set.seed(123)

# Define a grid of beta values to test
  
    beta_vals <- seq(0.05, 1, by = 0.05)

# Prepare a data frame to store results

    results <- data.frame(beta = beta_vals, 
                          max_theta = NA, 
                          wald = NA,
                          violation_detected = NA)
                          

# Loop over each beta value
    for (i in seq_along(beta_vals)) {
     
      # Simulate data with exclusion restriction violated (beta != 0)
      
          df <- simulate_data(n = 100000, alpha = 0.6, beta = beta_vals[i])
      
      # Compute HM2015 inequalities
      
          theta <- get_theta_HM(df)
      
      # Record the largest theta (if > 0, violation is detected)
      
          results$max_theta[i] <- round(max(theta), 4)
      
          results$violation_detected[i] <- max(theta) > 0
      
      # Record the Wald estimand (should be biased away from 1)
     
           results$wald[i] <- round(iv_estimate(df), 4)
    }
```

Display the results:

```{r}
kable(results,
      col.names = c("$\\beta$", "max($\\theta_j$)", "Wald Estimand","Violation Detected"),
      caption = "HM2015 detection across values of $\\beta$",
      booktabs = TRUE,
      row.names = FALSE,
      escape = FALSE,
      format = "latex"
) %>% kable_styling(latex_options = "hold_position")
```
For small values of $\beta$ (e.g. around 0.1--0.2), the exclusion restriction is violated ($\beta \neq 0$) and the Wald estimand is biased away from 1, yet all four $\theta_j \leq 0$, no violation is detected by the HM2015 test.

This is possible because the HM2015 inequalities are **necessary but not sufficient** conditions for instrument validity. The test works by checking whether the point-identified mean potential outcome of the always takers (or never takers) falls outside bounded intervals in the mixed population. These bounds have positive width, which is driven by the share of compliers. When the direct effect of $Z$ on $Y$ is small, the point-identified mean can still land within the bounds despite the exclusion restriction being violated. 

In other words, small violations are "absorbed" within the width of the bounds.

\newpage

# Question 7 

Recall from MW2017 Equation (5), the testable implications of $LI + LM$ are, for any $y \in \mathcal{Y}$

$$\theta(y, 1) \equiv E[p_1 D(1-Z) - p_0 DZ \mid Y = y] \leq 0$$
$$\theta(y, 0) \equiv E[p_0 (1-D)Z - p_1 (1-D)(1-Z) \mid Y = y] \leq 0$$

where $p_k = \Pr(Z = k)$.

These are **conditional** moment inequalities, they must hold at every value of $Y$. For each observation $i$, we compute:

$$L_i^1 = p_1 D_i(1 - Z_i) - p_0 D_i Z_i$$
$$L_i^0 = p_0 (1 - D_i)Z_i - p_1 (1 - D_i)(1 - Z_i)$$

Then $\theta(y, 1) = E[L^1 \mid Y = y]$ and $\theta(y, 0) = E[L^0 \mid Y = y]$. Since $Y$ is continuous, for a given scalar $y$ we evaluate the conditional expectation at observations where $Y = y$:

```{r}
get_theta_MW <- function(data, y) {
  
  p1 <- mean(data$Z)
  p0 <- 1 - p1
  n  <- nrow(data)
  
  # Bandwidth (Silverman rule of thumb)
    
    h <- 1.06 * sd(data$Y) * n^(-1/5)
  
  # Kernel weights and variables
  
    u <- (data$Y - y) / h
  
    K <- ifelse(abs(u) <= 1, 0.75 * (1 - u^2), 0)
  
  # Construct G_1 and G_0
  
    G1 <- p1 * data$D * (1 - data$Z) - p0 * data$D * data$Z
  
    G0 <- p0 * (1 - data$D) * data$Z - p1 * (1 - data$D) * (1 - data$Z)
  
  # Local linear regression: regress G on (Y - y), weighted by K
  
    Y_c <- data$Y - y
  
    thetay0 <- coef(lm(G0 ~ Y_c, weights = K))[1]
  
    thetay1 <- coef(lm(G1 ~ Y_c, weights = K))[1]
  
  c(thetay1 = thetay1, thetay0 = thetay0)
}
```


Since $Y$ is continuous, each observed $Y_i$ is essentially unique, so evaluating at a single $y$ returns the value of $L^0_i$ and $L^1_i$ for that observation.

# Question 8









