---
title: "AERM_part2"
subtitle: MT25 Assignment 1 - Part 2
author: "Candidate Number: 1091806"
date: "2026-03-25"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
    toc_depth: 2
    number_sections: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
---

# Libraries
```{r}
# load libraries

library(mvtnorm)  # Question 1
library(knitr)    # Question 1
```

\newpage
# Question 1

**Model recap**:

$$Y = D + \beta Z + U$$

$$D = \mathbf{1}\{\alpha Z + \epsilon > 0\}$$

$$\begin{pmatrix} U \\ \epsilon \end{pmatrix} \sim N\left(\mathbf{0}, \Sigma\right), \quad \Sigma = \begin{pmatrix} 1 & 0.5 \\ 0.5 & 1 \end{pmatrix}$$

$$Z \sim \text{Bernoulli}(0.5), \quad \text{independent of } (U, \epsilon)$$

Construct Function 1: Data Generating Process (DGP) simulation 
```{r}
# DGP simulation function

simulate_data <- function(n, alpha, beta) {
  
  # Draw the correlated errors, (U, epsilon)
    
    # Construct sigma
    
      Sigma <- matrix(c(1, 0.5, 0.5, 1), nrow = 2)
    
    # Construct errors 
    
      errors <- rmvnorm(n, mean = c(0, 0), sigma = Sigma)
    
    U <- errors[, 1]
    eps <- errors[, 2]
  
  # Draw the instrument Z, independent of errors
 
    Z <- rbinom(n, size = 1, prob = 0.5)
  
  # Generate treatment D from the selection equation
       
    D <- as.integer(alpha * Z + eps > 0)
      
  # Generate outcome Y from the outcome equation    
      
    Y <- D + beta * Z + U
  
  # Return a data frame with named columns D, Z, and Y
      
    data.frame(D = D, Z = Z, Y = Y)
}
```

Construct Function 2: Wald (IV) estimator

$$\hat{\beta}_{IV} = \frac{\bar{Y}_{Z=1} - \bar{Y}_{Z=0}}{\bar{D}_{Z=1} - \bar{D}_{Z=0}}$$
```{r}
# Wald (IV) estimator

iv_estimate <- function(df) {
  
  # Numerator: reduced-form effect of Z on Y
    
    ey1 <- mean(df$Y[df$Z == 1])
    ey0 <- mean(df$Y[df$Z == 0])
  
  # Denominator: first-stage effect of Z on D
  
    ed1 <- mean(df$D[df$Z == 1])
    ed0 <- mean(df$D[df$Z == 0])
      
      
    (ey1 - ey0) / (ed1 - ed0)
    }
```

Test with beta = 0, exclusion restriction holds:
```{r}
# The Wald estimand should converge to the true LATE = 1

# Set seed
  
    set.seed(123)

# Run tests:

  results <- data.frame(
    n     = rep(10^5, 3),
    alpha = c(0.6, 1.5, 0.6),
    beta  = c(0, 0, 1),
    
    IV_estimate = c(
      
    # Test 1: beta=0 (exclusion holds) -> IV should be close to 1
      
      iv_estimate(simulate_data(10^5, alpha = 0.6, beta = 0)),
    
    # Test 2: beta=0 with stronger first stage -> IV still close to 1
    
      iv_estimate(simulate_data(10^5, alpha = 1.5, beta = 0)),
    
    # Test 3: beta=1 (exclusion violated) -> IV biased away from 1
      
      iv_estimate(simulate_data(10^5, alpha = 0.6, beta = 1))
  )
)

# Display results
  
  kable(results, digits = 4, caption = "IV estimates for different parameter values")
```

\newpage
# Question 2

## 2a)

Treatment is assigned by:

$$D = \mathbf{1}\{\alpha Z + \epsilon > 0\}$$
So, conditional on Z, individuals with larger $ϵ$ are more likely to be treated.

Since: 

$$Corr(U,ϵ) = 0.5 > 0$$
Those who select into treatment also tend to have higher values on the outcome error $U$. This means: 

$$ E[U∣D=1] > E[U∣D=0]$$ 

So, **D is endogenous**. OLS of Y on D would be biased upward due to positive selection.







