---
title: "Advanced Empirical Research Methods Hilary 2025/26"
author: 'Candidate ID: 1091806'
date: "2026-04-13"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
    toc_depth: 2
    number_sections: false
  html_document:
    toc: true
    toc_depth: '2'
    df_print: paged
subtitle: 'Assessed Problem Set: Empirical IO'
fontsize: 11pt
geometry: margin=1in
header-includes:
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{amsmath}
- \usepackage{newunicodechar}
- \newunicodechar{ρ}{$\rho$}
- \newunicodechar{π}{$\pi$}
- \newunicodechar{ξ}{$\xi$}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Set-up

## Libraries

```{r message=FALSE, warning=FALSE}
# Beer demand

  library(tidyverse)
  library(ivreg)
  library(modelsummary)
  library(knitr)

# Auto demand

  library(BLPestimatoR)
  library(nleqslv)
```

## Set seed

```{r}
# Set seed to 130

  set.seed(130)
```

\newpage

# Q1. Beer demand

## Q1 (a) Data Description

```{r load-beerdata}
# Load the beer data

  beer <- read_csv("beer_data.csv")

# Preview structure
  
  glimpse(beer)
```

```{r summary-stats}
# Summary statistics for key variables

  beer %>%
    select(price, quantity, marginal_cost, alc_j, marketsize, num_prod_mkt) %>%
    summary()
```

```{r distributions, fig.height=4, fig.width=8}
# 1. Distributions of price and alcohol content (histograms)

  par(mfrow = c(1, 3))
  
  hist(beer$price,        main = "Price (£)",         xlab = "£",   col = "red")
  hist(beer$alc_j,        main = "Alcohol Content",   xlab = "alc", col = "orange")
  hist(beer$marginal_cost,main = "Marginal Cost (£)", xlab = "£",   col = "yellow")
```

-   **Price histogram**: shows the shape of the price distribution
    (right-skewed, most beers cheap with a few premium outliers)

-   **Alcohol content histogram**: shows alcohol is uniformly
    distributed. There is good variation in strength across products, so
    `alc_j` will have identifying power in regressions.

-   **Marginal cost histogram**: shows costs are tightly distributed
    between £1.00 and £1.60. Costs vary enough to be a useful instrument
    for price.

```{r alc-price-plots, fig.height=4, fig.width=10}
# Compute binned means first

  binned <- beer |>
    mutate(alc_bin = cut(alc_j, breaks = 20)) |>
    group_by(alc_bin) |>
    summarise(mean_price = mean(price),
              alc_mid    = mean(alc_j))

# Side-by-side plots

  par(mfrow = c(1, 2))

# Left: Price vs alcohol content (raw scatter)
  # This is to test whether stronger beers are more expensive
  # Relevant for the later part (d) where the tax is proportional to alcohol

  plot(beer$alc_j, beer$price,
       main = "Price vs Alcohol Content",
       xlab = "Alcohol Content (alc_j)",
       ylab = "Price (£)",
       pch  = 16,
       col  = adjustcolor("red", alpha.f = 0.2),
       cex  = 0.5)

# Right: Mean Price by Alcohol Content (binned means)

  plot(binned$alc_mid, binned$mean_price,
       main = "Mean Price by Alcohol Content",
       xlab = "Alcohol Content (alc_j)",
       ylab = "Mean Price (£)",
       pch  = 16,
       col  = "violet",
       cex  = 1.2)
```

**Observations:**

The raw scatter (left) appears largely flat, but the binned means
(right) reveal a positive relationship. Mean price rises from around
£3.75 at low alcohol to £3.95 at high alcohol. The effect is modest in
magnitude (\~£0.20 range) but consistent across alcohol bins. Two
features also stand out in the raw scatter: a "floating band" of premium
beers priced around £13 regardless of alcohol content, and a much denser
mass of mainstream beers clustered between £3 and £5. This pattern
matters for part (d): since stronger beers are already slightly more
expensive on average, an alcohol-based tax will amplify the existing
price dispersion.

```{r market-structure}
# Number of markets and products per market

  n_per_market <- beer |>
    group_by(market_id) |>
    summarise(n_products = n()) |>
    pull(n_products)

# Assemble summary table

  market_table <- data.frame(
    Statistic = c("Number of unique markets",
                  "Number of product-market observations",
                  "Min products per market",
                  "Median products per market",
                  "Mean products per market",
                  "Max products per market"),
    Value = c(n_distinct(beer$market_id),
              nrow(beer),
              min(n_per_market),
              median(n_per_market),
              round(mean(n_per_market), 2),
              max(n_per_market))
  )
  
  kable(market_table, caption = "Market Structure")
```

**Comment:**

The data contains `r n_distinct(beer$market_id)` markets and
`r nrow(beer)` product-market observations, with between
`r min(beer$num_prod_mkt)` and `r max(beer$num_prod_mkt)` products per
market (mean `r round(mean(table(beer$market_id)), 1)` products per
market). Prices range from £`r round(min(beer$price), 2)` to
£`r round(max(beer$price), 2)` with a mean of
£`r round(mean(beer$price), 2)`, and are right-skewed; most beers are
cheap with a few expensive outliers. Marginal costs range from
£`r round(min(beer$marginal_cost), 2)` to
£`r round(max(beer$marginal_cost), 2)`, always below price, consistent
with positive markups. Alcohol content (`alc_j`) is uniformly
distributed between `r round(min(beer$alc_j), 2)` and
`r round(max(beer$alc_j), 2)`, suggesting products are evenly spread
across the strength spectrum. Market sizes range from
`r round(min(beer$marketsize), 0)` to `r round(max(beer$marketsize), 0)`
potential purchase occasions.

\newpage

## Q1 (b) Nested Logit Estimation

**Step 1: Compute Market Shares**

```{r market-shares}
beer <- beer |> 
  group_by(market_id) |> 
  mutate(
    
    # Market share of product j in market t
    share_jt  = quantity / marketsize,
    
    # Aggregate share of all inside goods (all beers) in market t
    share_agg = sum(share_jt),
    
    # Outside good share: fraction who don't buy any beer
    share_0   = 1 - share_agg,
    
    # Within-group share: j's share among beer buyers
    # Since all inside goods are in one nest: s_{j|g} = s_jt / (1 - s_0t)
    share_jg  = share_jt / share_agg
    
  ) |> 
  ungroup()

# Sanity checks

  sanity_table <- data.frame(
    Check = c("Min outside good share",
              "Max outside good share",
              "All shares positive"),
    Value = c(round(min(beer$share_0), 4),
              round(max(beer$share_0), 4),
              all(beer$share_jt > 0))
  )
  
  kable(sanity_table, caption = "Market Share Sanity Checks")
```

**Step 2: Construct the Dependent Variable and Regressors**

```{r dependent-var, results='hide'}
beer <- beer |> 
  mutate(
    
    # Nested logit dependent variable: ln(s_jt) - ln(s_0t)
    delta     = log(share_jt) - log(share_0),
    
    # Within-group share term that appears on RHS
    ln_share_jg = log(share_jg)
  )

summary(beer$delta)
```

| Statistic | Value                                    |
|-----------|------------------------------------------|
| Min       | `r round(min(beer$delta), 3)`            |
| 1st Qu.   | `r round(quantile(beer$delta, 0.25), 3)` |
| Median    | `r round(median(beer$delta), 3)`         |
| Mean      | `r round(mean(beer$delta), 3)`           |
| 3rd Qu.   | `r round(quantile(beer$delta, 0.75), 3)` |
| Max       | `r round(max(beer$delta), 3)`            |

**Step 3: Estimate the Nested Logit Model**

The estimating equation is:

$$\ln(s_{jt}) - \ln(s_{0t}) = \beta_0 + \beta_{\text{alc}} \cdot \text{alc}_j 
- \alpha \cdot p_{jt} + \sigma \cdot \ln(s_{j|g,t}) + \xi_{jt}$$

Both `price` and `ln_share_jg` are **endogenous**:

-   Price is endogenous because firms observe $\xi_{jt}$ before
    setting prices. Beers with high unobserved quality command higher
    prices, creating a positive correlation between `price` and
    $\xi_{jt}$. OLS would therefore underestimate price sensitivity.

-   Within-group share $\ln(s_{j|g,t})$ is endogenous because it
    depends on all prices in the market, which all depend on the
    $\xi$'s. If product $j$ has high $\xi_{jt}$, consumers substitute
    towards it, inflating its within-group share. OLS would therefore
    overestimate $\hat{\sigma}$.

So we need instruments:

-   `marginal_cost` for `price` (cost shifter, excluded from demand)
-   `num_prod_mkt` for `ln_share_jg` (more products dilutes within-group
    share but does not enter consumer utility directly)

```{r nested-logit-iv}
# OLS (for comparison - biased due to endogeneity)

  model_ols <- lm(delta ~ alc_j + price + ln_share_jg, data = beer)

# IV / 2SLS: instrument price with marginal_cost,
#            instrument ln_share_jg with num_prod_mkt

  model_iv <- ivreg(
    delta ~ alc_j + price + ln_share_jg |
            alc_j + marginal_cost + num_prod_mkt,
    data = beer
  )
```

```{r Q1bresults-table}
# Compare OLS and IV results

modelsummary(
  list("OLS" = model_ols, "IV (2SLS)" = model_iv),
  stars  = TRUE,
  gof_omit = "AIC|BIC|Log|F",
  title  = "Nested Logit Estimates"
)
```

```{r first-stage, results='hide'}
# Check first-stage relevance: instruments must be correlated with endogenous vars

# First stage for price

  fs_price <- lm(price ~ alc_j + marginal_cost + num_prod_mkt, data = beer)

# First stage for ln_share_jg

  fs_share <- lm(ln_share_jg ~ alc_j + marginal_cost + num_prod_mkt, data = beer)
```

```{r first-stage-output}
# First-stage regression results

  kable(round(summary(fs_price)$coefficients, 4),
        caption = "First Stage: Price")

  kable(round(summary(fs_share)$coefficients, 4),
        caption = "First Stage: ln(within-group share)")
```

**Identification assumptions:** The validity of the IV estimates rests
on two conditions.

First, *relevance*: marginal costs must be correlated with prices
(satisfied if firms pass costs through to prices) and the number of
products must be correlated with within-group shares (satisfied
mechanically, since more products dilutes each product's share of the
nest).

Second, *exclusion*: marginal costs must not directly affect consumer
utility (reasonable, since consumers care about price, not cost), and
the number of products in the market must not directly shift a
consumer's valuation of any individual beer (reasonable under the nested
logit structure). The residuals $\xi_{jt}$ are our estimate of
unobserved product quality, ie. product attributes that consumers value
but that we do not observe.

**Step 4: Extract Residuals as Unobserved Utility**

```{r xi-vector, fig.width=6, fig.height=4}
# Vector ξ

# Residuals from IV regression are our estimate of xi_{jt}
# These capture unobserved product quality not explained by alc_j or price
beer$xi <- residuals(model_iv)

summary(beer$xi)

hist(beer$xi, 
     main = "Distribution of Unobserved Utility (xi)",
     xlab = expression(xi),
     col  = "blue",
     breaks = 30)
```

**Comment on estimates:** The IV price coefficient ($\hat{\alpha}$) is
`r round(coef(model_iv)["price"], 3)`, which is negative: higher prices
reduce demand. The nesting parameter $\hat{\sigma}$ on `ln_share_jg` is
`r round(coef(model_iv)["ln_share_jg"], 3)`, which lies in (0,1),
consistent with the nested logit model. The OLS
price coefficient is less negative (attenuated towards zero) because
firms charge more for products with high unobserved quality, creating
upward bias. The instruments correct for this. The coefficient on
alcohol content is positive: consumers prefer stronger beers,
all else equal.

\newpage

## Q1 (c) Elasticities, Marginal Costs and Diversion Ratios

**Step 1: Extract Parameters and Compute Own-Price Elasticities**

We extract the estimated price parameter $\hat{\alpha}$ and nesting
parameter $\hat{\sigma}$ from the IV model estimated in part (b). Note
we flip the sign on the price coefficient so that $\alpha$ is positive,
matching the notation in the elasticity and markup formulas.

```{r parameters}
# Extract estimated parameters from IV model

alpha <- -coef(model_iv)["price"]          # price parameter (positive)
sigma <- coef(model_iv)["ln_share_jg"]     # nesting parameter

# Display estimated parameters

  param_table <- data.frame(
    Parameter = c("$\\hat{\\alpha}$ (price)",
                  "$\\hat{\\sigma}$ (nesting)"),
    Value     = c(round(alpha, 4),
                  round(sigma, 4))
  )
  
  kable(param_table, caption = "Estimated Parameters")
```

With $\hat{\alpha}$ and $\hat{\sigma}$ in hand, we can compute own-price
elasticities for each product-market observation using the nested logit
formula:

$$\eta_{jj} = -\alpha \cdot p_{jt} \cdot 
\frac{1 - \sigma s_{j|g,t} - (1-\sigma)s_{jt}}{1-\sigma}$$

The term in the numerator reflects that a price rise loses customers to
two destinations: other beers (weighted by $\sigma$) and the outside
good (weighted by $1-\sigma$).

```{r elasticities}
# Own-price elasticity for nested logit (single nest):
# eta_jj = -alpha * p_jt * (1 - sigma*s_{j|g} - (1-sigma)*s_j) / (1 - sigma)

beer <- beer |>
  mutate(
    elast_own = -alpha * price * 
      (1 - sigma * share_jg - (1 - sigma) * share_jt) / (1 - sigma)
  )
```

*Summary table*

| Statistic | Value                                        |
|-----------|----------------------------------------------|
| Min       | `r round(min(beer$elast_own), 3)`            |
| 1st Qu.   | `r round(quantile(beer$elast_own, 0.25), 3)` |
| Median    | `r round(median(beer$elast_own), 3)`         |
| Mean      | `r round(mean(beer$elast_own), 3)`           |
| 3rd Qu.   | `r round(quantile(beer$elast_own, 0.75), 3)` |
| Max       | `r round(max(beer$elast_own), 3)`            |

To visualise the distribution and check how many products fall in the
inelastic region ($|\eta| < 1$), we plot a histogram with a reference
line at $-1$.

```{r elast-plot, fig.width=6, fig.height=4}
hist(beer$elast_own,
     main = "Distribution of Own-Price Elasticities",
     xlab = "Own-Price Elasticity",
     col  = "blue",
     breaks = 40)
abline(v = -1, col = "red", lty = 2, lwd = 2)
legend("topleft", legend = "Elasticity = -1", col = "red", lty = 2)
```

**Step 2: Back Out Marginal Costs**

Under single-product Nash, the markup is:

$$\text{markup}_{jt} = p_{jt} - mc_{jt} =  \frac{1 - \sigma}{\alpha(1 - \sigma s_{j|g,t} - (1-\sigma)s_{jt})}$$

```{r marginal-costs}
beer <- beer |>
  mutate(
    markup_implied = (1 - sigma) / 
      (alpha * (1 - sigma * share_jg - (1 - sigma) * share_jt)),
    mc_implied = price - markup_implied
  )
```

*Summary of backed-out marginal costs*

| Statistic | Marginal Cost | Markup |
|------------------------|------------------------|------------------------|
| Min | `r round(min(beer$mc_implied), 3)` | `r round(min(beer$markup_implied), 3)` |
| Median | `r round(median(beer$mc_implied), 3)` | `r round(median(beer$markup_implied), 3)` |
| Mean | `r round(mean(beer$mc_implied), 3)` | `r round(mean(beer$markup_implied), 3)` |
| Max | `r round(max(beer$mc_implied), 3)` | `r round(max(beer$markup_implied), 3)` |

As a sanity check, we compare our backed-out marginal costs with the
ones provided in the data. They should be strongly correlated if both
the demand model and the single-product Nash assumption are reasonable
approximations.

```{r mc-check, fig.width=6, fig.height=4}
# Check: compare backed-out mc with mc in data

cat("Correlation between backed-out and data mc:",
    round(cor(beer$mc_implied, beer$marginal_cost), 4), "\n")

plot(beer$marginal_cost, beer$mc_implied,
     main = "Backed-out vs Data Marginal Costs",
     xlab = "MC from data", ylab = "MC backed out",
     pch  = 16, col = adjustcolor("blue", alpha.f = 0.3),
     cex  = 0.5)
abline(0, 1, col = "red", lwd = 2)
```

**Step 3: Diversion Ratio to Outside Good**

The diversion ratio from product $j$ to the outside good is the share of
customers leaving $j$ who switch to the outside good rather than to
another beer. Using the nested logit own- and cross-price effects
(lecture slide 83), and the fact that the outside good is alone in its
own nest (so cross-nest effects apply):

$$D_{j \to 0} = -\frac{\partial s_0 / \partial p_j}{\partial s_j / \partial p_j} 
= \frac{(1-\sigma) s_0}{1 - \sigma s_{j|g} - (1-\sigma) s_j}$$

Comparing this to the raw outside good share illustrates the role of
$\sigma$: in a simple logit ($\sigma = 0$), these would be equal; with
$\sigma > 0$, the diversion ratio is lower because consumers prefer to
stay within the nest.

```{r diversion}
beer <- beer |>
  mutate(
    diversion_outside = (1 - sigma) * share_0 /
      (1 - sigma * share_jg - (1 - sigma) * share_jt)
  )
```

*Compare diversion ratio to outside good share directly*

| Statistic | Diversion Ratio to Outside | Outside Good Share |
|------------------------|------------------------|------------------------|
| Mean | `r round(mean(beer$diversion_outside), 3)` | `r round(mean(beer$share_0), 3)` |
| Median | `r round(median(beer$diversion_outside), 3)` | `r round(median(beer$share_0), 3)` |

**Comment:**

Own-price elasticities range from `r round(min(beer$elast_own), 2)` to
`r round(max(beer$elast_own), 2)`, with mean
`r round(mean(beer$elast_own), 2)`. All elastic, with most
roughly 1.5× elastic. Backed-out marginal costs are tightly correlated
with those in the data (correlation
`r round(cor(beer$mc_implied, beer$marginal_cost), 2)`, but the scatter
reveals two distinct bands: cheap beers (below £8) lie just below the
45° line, while expensive beers (around £13) lie on a parallel band
further below; their backed-out MC is systematically lower than the
data MC. This second band reflects products with very high markups (the
premium segment), where the single-product Nash assumption predicts
higher implied markup, hence lower implied MC for a given price. The
diversion ratio to the outside good
(`r round(mean(beer$diversion_outside), 3)`) is far lower than the
outside good share (`r round(mean(beer$share_0), 3)`), reflecting strong
within-nest substitution consistent with $\hat{\sigma}$ =
`r round(sigma, 3)`. This is the the main flexibility limitation of
nested logit: all within-nest cross-price elasticities are forced to be
identical, even though real beers differ substantially in their
similarity.

\newpage

## Q1 (d) £10 Tax per Unit of Alcohol Strength

The government is considering a £10 tax per unit of alcohol strength
(`alc_j`). This increases each product's marginal cost by
$10 \times \text{alc}_j$, so stronger beers face a larger cost shock.
We use our backed-out marginal costs from part (c) as the pre-tax
baseline, rather than the marginal costs given in the data, to ensure
internal consistency with the demand model.

**Step 1: Compute New Marginal Costs After Tax**

We first construct the post-tax marginal cost for each product.

```{r tax-mc}
tax_rate <- 10

beer <- beer |>
  mutate(
    # Tax increases mc by 10 * alc_j for each product
    
      mc_tax = mc_implied + tax_rate * alc_j,
      delta_mc = mc_tax - mc_implied
    )

cat("Mean marginal cost increase (delta_mc):", 
    round(mean(beer$delta_mc), 4), "\n")
```

**Step 2: Define Share Function for Contraction Mapping**

To solve for counterfactual equilibrium prices, we need a function that
computes nested logit market shares given a price vector. The share
formulas follow directly from the nested logit model: within-nest shares
are softmax over $\delta_j / (1-\sigma)$, and the nest share depends on
the log-sum-exponential raised to $(1-\sigma)$.

```{r share-function}
# Function to compute nested logit shares given prices in a single market

compute_shares <- function(price_t, delta_base_t, sigma) {
  
  # delta_base_t = beta0 + beta_alc*alc_j + xi_jt (no price term)
  # Full mean utility including price:
  
    delta_t <- delta_base_t - alpha * price_t   # using alpha from model_iv
  
  # Within-nest shares
    
    exp_d  <- exp(delta_t / (1 - sigma))
    sum_eg <- sum(exp_d)
    
    sj_g <- exp_d / sum_eg
  
  # Group share
    
    sg <- sum_eg^(1 - sigma) / (1 + sum_eg^(1 - sigma))
  
  # Product shares
  
    sj <- sj_g * sg
  
  return(list(sj = sj, sj_g = sj_g, s0 = 1 - sum(sj)))
}
```

**Step 3: Contraction Mapping to Solve for Counterfactual Prices**

The counterfactual equilibrium cannot be solved analytically because
prices and shares are jointly determined: each firm's price depends on
shares, which depend on all prices in the market. We therefore use a
contraction mapping. Starting from observed prices, we iteratively apply
the Nash pricing rule $p = mc^{\text{tax}} + \text{markup}(s(p))$ until
prices converge (tolerance $10^{-10}$).

A key subtlety: we need the "base utility"
$\delta^{\text{base}}_{jt} = \beta_0 + \beta_{\text{alc}} \text{alc}_j + \xi_{jt}$
that excludes the price term, which we recover from the estimated
$\delta_{jt}$ by adding back $\alpha p_{jt}$ and subtracting the
$\sigma \ln(s_{j|g,t})$ term.

```{r contraction}
# We solve market by market
# For each market: iterate p = mc_tax + markup(shares(p)) until convergence

solve_prices <- function(beer_mkt, mc_col, sigma, alpha, 
                          tol = 1e-10, maxit = 10000) {
  
  # Base utility (excludes price): fitted values + alpha*price removes price effect
  
    delta_base <- beer_mkt$delta - log(beer_mkt$share_jg) * sigma +
                alpha * beer_mkt$price
  
  # Actually: delta_base = beta0 + beta_alc*alc_j + xi = delta - sigma*ln(s_jg) + alpha*p
  # Since delta = beta0 + beta_alc*alc_j - alpha*p + sigma*ln(s_jg) + xi
  # => delta_base = delta + alpha*p - sigma*ln(s_jg)
  
    delta_base <- beer_mkt$delta + alpha * beer_mkt$price - 
                sigma * log(beer_mkt$share_jg)
  
 
     mc_t  <- beer_mkt[[mc_col]]
     p_t   <- beer_mkt$price  # starting guess
  
  for (iter in 1:maxit) {
    
    shares <- compute_shares(p_t, delta_base, sigma)
    sj_t   <- shares$sj
    sj_g_t <- shares$sj_g
    
    # Single-product Nash markup
   
     markup_t <- (1 - sigma) / 
      (alpha * (1 - sigma * sj_g_t - (1 - sigma) * sj_t))
    
      p_new <- mc_t + markup_t
    
      if (max(abs(p_new - p_t)) < tol) break
      p_t <- p_new
    }
  
  return(p_new)
}
```

We loop over all `r n_distinct(beer$market_id)` markets, solving two
price vectors in each: one with the observed marginal costs (a
consistency check that should reproduce observed prices), and one with
the post-tax marginal costs.

```{r solve-all-markets, cache=TRUE}
# Solve for counterfactual prices in each market

  markets <- unique(beer$market_id)

# Pre-allocate

  beer$price_tax  <- NA
  beer$price_notax <- NA  # check: should recover original prices

  for (m in markets) {
  
    idx <- beer$market_id == m
    beer_m <- beer[idx, ]
    
    # Check: no tax should recover original prices
    beer$price_notax[idx] <- solve_prices(beer_m, "mc_implied", sigma, alpha)
    
    # With tax
    beer$price_tax[idx]   <- solve_prices(beer_m, "mc_tax",    sigma, alpha)
  }
```

**Step 4: Check - No-Tax Prices Should Equal Observed Prices**

Before analysing tax effects, we verify that running our contraction
mapping with unchanged marginal costs recovers the observed prices. Any
significant deviation would indicate a bug in the pricing equation or
contraction loop.

```{r check-notax}
# Verify: no-tax contraction should reproduce observed prices

  notax_check <- data.frame(
    Statistic = c("Max deviation (no-tax vs observed)",
                  "Mean deviation"),
    Value     = c(round(max(abs(beer$price_notax - beer$price)), 6),
                  round(mean(abs(beer$price_notax - beer$price)), 6))
  )
  
  kable(notax_check, caption = "No-Tax Consistency Check")
```

**Step 5: Analyse Price Effects**

Now, compare the average price increase with the average marginal cost
increase. Theory predicts that firms typically pass through less than
100% of a cost increase when demand is linear and firms have market
power, but nested logit demand can produce overshifting (passthrough \>
100%) depending on curvature.

```{r price-effects}
# Compute price changes

  beer <- beer |>
    mutate(
      delta_price   = price_tax - price,
      pct_price_chg = 100 * (price_tax - price) / price
    )

# Summarise price effects

  price_effects <- data.frame(
    Statistic = c("Mean price increase (£)",
                  "Mean MC increase (£)",
                  "% change in average price"),
    Value     = c(round(mean(beer$delta_price), 3),
                  round(mean(beer$delta_mc), 3),
                  round(100 * (mean(beer$price_tax) - mean(beer$price)) / mean(beer$price), 3))
  )
  
  kable(price_effects, caption = "Price Effects of the Tax")
```

**Step 6: Compute New Quantities and Analyse by Beer Type**

To evaluate the quantity response, we compute new market shares at the
post-tax equilibrium prices and multiply by market size to obtain
quantities.

```{r new-quantities}
# Compute new shares and quantities after tax

beer <- beer |>
  group_by(market_id) |>
  mutate(
    
    # Recompute share_0 at new prices using contraction output
    # New delta at tax prices:
    
      delta_tax  = delta + alpha * (price - price_tax),  
    
    # Recompute within-group and group shares correctly
    
      exp_d_tax    = exp(delta_tax / (1 - sigma)),
      sum_eg_tax   = sum(exp_d_tax),
      sj_g_tax     = exp_d_tax / sum_eg_tax,
      sg_tax       = sum_eg_tax^(1 - sigma) / (1 + sum_eg_tax^(1 - sigma)),
      share_jt_tax = sj_g_tax * sg_tax,
      quantity_tax = share_jt_tax * marketsize
    ) |>
    ungroup()

# Total quantity change

  Q_pre  <- sum(beer$quantity)
  Q_post <- sum(beer$quantity_tax)
  
  qty_table <- data.frame(
    Statistic = c("Total quantity pre-tax",
                  "Total quantity post-tax",
                  "% change in total quantity"),
    Value     = c(round(Q_pre, 2),
                  round(Q_post, 2),
                  round(100 * (Q_post - Q_pre) / Q_pre, 3))
  )
  
  kable(qty_table, caption = "Total Quantity Response to the Tax")
```

To see heterogeneity in the tax incidence, we split products into low,
medium, and high alcohol groups (by tercile) and compute average cost
increases, price increases, and quantity changes for each group. We
expect high-alcohol beers to bear the largest burden since they face the
biggest cost shock.

```{r by-beer-type, fig.height=4, fig.width=7}
# Analyse by alcohol strength: split into low/medium/high
  
  beer <- beer |>
    mutate(
      alc_group = cut(alc_j,
                      breaks = quantile(alc_j, c(0, 1/3, 2/3, 1)),
                      labels = c("Low Alcohol", "Medium Alcohol", "High Alcohol"),
                      include.lowest = TRUE)
    )

# Mean price change and quantity change by group

  type_summary <- beer |>
    group_by(alc_group) |>
    summarise(
      mean_delta_mc    = round(mean(delta_mc), 3),
      mean_delta_price = round(mean(delta_price), 3),
      pct_qty_change   = round(100 * (sum(quantity_tax) - sum(quantity)) / sum(quantity), 2)
    )
  
  kable(type_summary, 
        col.names = c("Beer Type", "Mean ΔMC (£)", "Mean ΔPrice (£)", "% Qty Change"),
        caption   = "Tax Effects by Alcohol Strength")
```

**Step 7: Aggregate Demand Elasticity**

Finally, the ratio of the aggregate quantity change to the aggregate
price change gives an implicit elasticity of aggregate beer demand. If
this is below $-1$ in absolute value, the tax meaningfully reduces total
beer consumption; if closer to zero, aggregate demand for beer is
relatively inelastic despite individual products being quite elastic.

```{r agg-elasticity}
pct_price <- 100 * (mean(beer$price_tax) - mean(beer$price)) / mean(beer$price)
pct_qty   <- 100 * (Q_post - Q_pre) / Q_pre

# Aggregate effects

  agg_table <- data.frame(
    Statistic = c("% change in average price",
                  "% change in total quantity",
                  "Implied aggregate elasticity"),
    Value     = c(round(pct_price, 3),
                  round(pct_qty, 3),
                  round(pct_qty / pct_price, 3))
  )
  
  kable(agg_table, caption = "Aggregate Demand Response to the Tax")
```

**Comment:**

The average price increase (£`r round(mean(beer$delta_price), 3)`)
exceeds the average marginal cost increase
(£`r round(mean(beer$delta_mc), 3)`), indicating overshifting of the tax: firms raise prices by more than the cost increase. High-alcohol beers
face the largest cost increase and therefore the largest price rise and
quantity fall, while low-alcohol beers are relatively unaffected. The
implied aggregate demand elasticity of `r round(pct_qty / pct_price, 2)`
suggests aggregate beer demand is elastic: aggregate quantity falls
more than proportionately to the average price rise — meaning the tax
meaningfully reduces total beer consumption.

\newpage

# Q2. Auto demand

## Data Visualisation

```{r load-autodata}
autoData <- read_csv("autoData.csv")
own_pre  <- read_csv("dummies_cars.csv")
```

Before estimating the model, we briefly explore the data to understand
the key variables and market structure.

```{r auto-eda, fig.height=8, fig.width=10}
par(mfrow = c(2, 2))

# 1. Price distribution

  hist(autoData$price,
       main   = "Distribution of Car Prices",
       xlab   = "Price ($000s)",
       col    = "red",
       breaks = 40)

# 2. Price vs horsepower/weight

  plot(autoData$hpwt, autoData$price,
       main = "Price vs Horsepower/Weight",
       xlab = "Horsepower/Weight (hpwt)",
       ylab = "Price ($000s)",
       pch  = 16,
       col  = adjustcolor("violet", alpha.f = 0.4),
       cex  = 0.6)

# 3. Price vs market share

  plot(autoData$price, autoData$share,
       main = "Price vs Market Share",
       xlab = "Price ($000s)",
       ylab = "Market Share",
       pch  = 16,
       col  = adjustcolor("purple", alpha.f = 0.4),
       cex  = 0.6)

# 4. Number of products per year
  
  products_per_year <- autoData |>
    group_by(year) |>
    summarise(n_products = n())
  
  plot(products_per_year$year, products_per_year$n_products,
       main = "Number of Products per Year",
       xlab = "Year",
       ylab = "Number of Products",
       type = "b",
       pch  = 16,
       col  = "red")
```

The auto market contains `r n_distinct(autoData$year)` years
(1971--1990) and `r nrow(autoData)` product-market observations. Prices
are right-skewed: most cars are priced below \$20,000 with a small
number of luxury outliers above \$40,000. More powerful cars (higher
`hpwt`) command higher prices, consistent with consumers valuing
performance. Market shares are negatively related to price as expected.
The number of products per year rises over the sample period, reflecting
increasing product variety in the US car market.

\newpage

## Q2 (a) Estimating Two BLP Specifications

**Step 1: Prepare Data**

```{r load-auto data prepare}
# Rename to match BLPestimatoR conventions

  autoData <- autoData |>
    rename(firmid   = firm_id,
           cdid     = year,
           id       = car_id)

# Currently the year column in the data starts from 71
# However, BLPestimatoR requires markets numbered from 1

  autoData$cdid <- autoData$cdid - 70

# Add ownership dummies to product data

  colnames(own_pre) <- paste0("company", 1:26)
  autoData <- cbind(autoData, own_pre)
```

**Step 2: Construct BLP Instruments**

BLP instruments use the sum of characteristics of other products by the
same firm ("own" instruments, IV1-5) and rival firms ("rival"
instruments, IV6-10) in the same market. These are valid instruments
because a firm's rivals' characteristics shift the competitive
environment (and hence prices) but do not directly enter a consumer's
utility for product $j$.

```{r blp-ivs}
  nobs <- nrow(autoData)

# Matrix of the 5 product characteristics

  X <- data.frame(autoData$const, autoData$hpwt,
                autoData$air,   autoData$mpd, autoData$space)

  sum_other <- matrix(NA, nobs, ncol(X))
  sum_rival <- matrix(NA, nobs, ncol(X))
  
  for (i in 1:nobs) {
    other_ind <- autoData$firmid == autoData$firmid[i] &
                 autoData$cdid   == autoData$cdid[i]   &
                 autoData$id     != autoData$id[i]
    rival_ind <- autoData$firmid != autoData$firmid[i] &
                 autoData$cdid   == autoData$cdid[i]
    sum_other[i, ] <- colSums(X[other_ind, ])
    sum_rival[i, ] <- colSums(X[rival_ind, ])
  }
  
  colnames(sum_other) <- paste0("IV", 1:5)
  colnames(sum_rival) <- paste0("IV", 6:10)
  autoData <- cbind(autoData, sum_other, sum_rival)
```

**Step 3: Income Draws for Specification 2**

Specification 2 follows BLP(1995) by interacting price with inverse
income $1/y_i$, where income is drawn from a log-normal distribution
$\text{Lognormal}(\mu_t, \sigma_y^2)$ with $\sigma_y = 1.72$ and
market-specific means $\mu_t$ as given in the question. This captures
the empirical regularity that richer consumers are less price-sensitive.

```{r income-draws}
  sigma_v    <- 1.72
  ndraws     <- 500

# Log income means for years 1971-1990 (from question Table 1)

  incomeMeans <- c(2.01156, 2.06526, 2.07843, 2.05775, 2.02915,
                 2.05346, 2.06745, 2.09805, 2.10404, 2.07208,
                 2.06019, 2.06561, 2.07672, 2.10437, 2.12608,
                 2.16426, 2.18071, 2.18856, 2.21250, 2.18377)

# Standard normal draws: 20 markets x 500 draws
 
  nu_0    <- matrix(rnorm(20 * ndraws), nrow = 20, ncol = ndraws)
  y_means <- matrix(incomeMeans,        nrow = 20, ncol = ndraws)

# Convert to income levels (log-normal)

  y_it    <- exp(y_means + sigma_v * nu_0)

# BLP(1995) use inverse income as the demographic

  inv_inc <- 1 / y_it
  
  colnames(inv_inc) <- paste0("draw_", 1:ndraws)

# Add market identifier in first column (required by BLPestimatoR)

  inv_inc        <- cbind(c(1:20), inv_inc)

  colnames(inv_inc)[1] <- "cdid"

# Package as list for BLPestimatoR

  demog_auto <- list(invinc = inv_inc)
```

**Step 4: Model Formulas**

Both specifications share the same linear characteristics and
instruments. They differ only in how heterogeneity enters the price
coefficient.

In **Specification 1**, all five characteristics and price have
independent normal random coefficients ($\sigma_k \nu_{ik}$). No
demographic data is used.

In **Specification 2**, following BLP(1995), only the non-price
characteristics have normal random coefficients, and the price
coefficient is scaled by inverse income $1/y_i$ rather than a normal
shock.

```{r model-formulas}
# Both specs: same linear terms and instruments
# share ~ linear | exogenous | random coefficients | extra instruments

# Specification 1: normal RC on price + all characteristics
  
  spec1_formula <- as.formula(
    "share ~ 0 + const + hpwt + air + mpd + space |
             0 + const + hpwt + air + mpd + space |
             0 + price + const + hpwt + air + mpd + space |
             0 + IV1 + IV2 + IV3 + IV4 + IV5 + IV6 + IV7 + IV8 + IV9 + IV10"
  )

# Specification 2: income-interacted price (BLP 1995 style)
# Identical formula; heterogeneity controlled via theta_guesses and demographics

  spec2_formula <- as.formula(
    "share ~ 0 + const + hpwt + air + mpd + space |
             0 + const + hpwt + air + mpd + space |
             0 + price + const + hpwt + air + mpd + space |
             0 + IV1 + IV2 + IV3 + IV4 + IV5 + IV6 + IV7 + IV8 + IV9 + IV10"
  )
```

**Step 5: BLP Data Objects**

```{r blp-data-spec1}
# Specification 1: no demographic draws

  car_data_s1 <- BLP_data(
    model              = spec1_formula,
    market_identifier  = "cdid",
    product_identifier = "id",
    productData        = autoData,
    additional_variables = paste0("company", 1:26),
    group_structure    = "model_id",
    blp_inner_tol      = 1e-9,
    blp_inner_maxit    = 5000,
    integration_method = "MLHS",
    integration_accuracy = ndraws,
    integration_seed   = 48
  )
```

```{r blp-data-spec2}
# Specification 2: with income demographic draws

  car_data_s2 <- BLP_data(
    model              = spec2_formula,
    market_identifier  = "cdid",
    product_identifier = "id",
    productData        = autoData,
    additional_variables = paste0("company", 1:26),
    group_structure    = "model_id",
    blp_inner_tol      = 1e-9,
    blp_inner_maxit    = 5000,
    integration_method = "MLHS",
    demographic_draws  = demog_auto,
    integration_accuracy = ndraws,
    integration_seed   = 48
  )
```

**Step 6: Starting Values**

The starting values (`theta_guesses`) specify which random coefficients
to estimate and their initial values. `NA` tells `BLPestimatoR` not to
estimate a parameter for that cell.

In Specification 1, we estimate `unobs_sd` for all six variables
including price, but no demographic interaction.

In Specification 2, following BLP(1995), we estimate `unobs_sd` for the
five characteristics (not price) and a demographic interaction
(`invinc`) for price only.

```{r theta-spec1}
# Spec 1: normal RC on price and all 5 characteristics
# Use modest positive starting values; BLPestimatoR treats these as sigmas

  theta1 <- matrix(c(0.1, 3, 1, 1, 0.5, 0.5),
                 nrow = 6, ncol = 1)
  rownames(theta1) <- c("price", "const", "hpwt", "air", "mpd", "space")
  colnames(theta1) <- "unobs_sd"
  
  theta1
```

```{r theta-spec2}
# Spec 2: normal RC on characteristics, income interaction on price
# 6 x 2 matrix

  theta2 <- matrix(c(0, -14.12, 5.32, -3.32, -3.17, -0.62,
                     -25.3, 0, 0, 0, 0, 0),
                   nrow = 6, ncol = 2)
  rownames(theta2) <- c("price", "const", "hpwt", "air", "mpd", "space")
  colnames(theta2) <- c("unobs_sd", "invinc")

# Replace zeros with NA (BLPestimatoR will not estimate these)
  
  theta2[theta2 == 0] <- NA
  
  theta2
```

**Step 7: Estimate Both Specifications**

We use BFGS optimisation with a relative tolerance of $10^{-10}$ and up
to 1000 iterations. Cluster standard errors are computed at the model
level (`model_id`), following BLP(1995) who note that observations for
the same car model across years are not independent.

```{r estimate-spec1, cache=TRUE, results='hide'}
car_est_s1 <- estimateBLP(
  blp_data       = car_data_s1,
  par_theta2     = theta1,
  solver_method  = "BFGS",
  solver_maxit   = 1000,
  solver_reltol  = 1e-10,
  standardError  = "cluster",
  extremumCheck  = FALSE,
  printLevel     = 0
)
```

```{r estimate-spec2, cache=TRUE, results='hide'}
car_est_s2 <- estimateBLP(
  blp_data       = car_data_s2,
  par_theta2     = theta2,
  solver_method  = "BFGS",
  solver_maxit   = 1000,
  solver_reltol  = 1e-10,
  standardError  = "cluster",
  extremumCheck  = FALSE,
  printLevel     = 0
)
```

**Results**

```{r Q2aresults-table}
# Build comparison table for linear coefficients

  lin_table <- data.frame(
    Variable = rownames(car_est_s1$theta_lin),
    Spec1_Est = round(car_est_s1$theta_lin[, 1], 3),
    Spec1_SE  = round(car_est_s1$se_linear,     3),
    Spec2_Est = round(car_est_s2$theta_lin[, 1], 3),
    Spec2_SE  = round(car_est_s2$se_linear,     3)
  )
  
  kable(lin_table, 
        col.names = c("Variable", "Spec 1 Est.", "Spec 1 SE",
                                  "Spec 2 Est.", "Spec 2 SE"),
        caption = "Linear Coefficients: Spec 1 vs Spec 2",
        row.names = FALSE)
```

```{r rc-table}
# Build comparison table for random coefficients (nonlinear parameters)
# Spec 1 has unobs_sd on all 6 vars; Spec 2 has unobs_sd on 5 chars + invinc*price

  rc_vars <- c("const", "hpwt", "air", "mpd", "space", "price")
  
  rc_table <- data.frame(
    Variable = rc_vars,
    
    Spec1_Est = round(c(
      car_est_s1$theta_rc["unobs_sd*const"],
      car_est_s1$theta_rc["unobs_sd*hpwt"],
      car_est_s1$theta_rc["unobs_sd*air"],
      car_est_s1$theta_rc["unobs_sd*mpd"],
      car_est_s1$theta_rc["unobs_sd*space"],
      car_est_s1$theta_rc["unobs_sd*price"]
    ), 3),
    
    Spec1_SE = round(c(
      car_est_s1$se_rc["unobs_sd*const"],
      car_est_s1$se_rc["unobs_sd*hpwt"],
      car_est_s1$se_rc["unobs_sd*air"],
      car_est_s1$se_rc["unobs_sd*mpd"],
      car_est_s1$se_rc["unobs_sd*space"],
      car_est_s1$se_rc["unobs_sd*price"]
    ), 3),
    
    Spec2_Est = round(c(
      car_est_s2$theta_rc["unobs_sd*const"],
      car_est_s2$theta_rc["unobs_sd*hpwt"],
      car_est_s2$theta_rc["unobs_sd*air"],
      car_est_s2$theta_rc["unobs_sd*mpd"],
      car_est_s2$theta_rc["unobs_sd*space"],
      car_est_s2$theta_rc["invinc*price"]    # Note: different parameter for Spec 2
    ), 3),
    
    Spec2_SE = round(c(
      car_est_s2$se_rc["unobs_sd*const"],
      car_est_s2$se_rc["unobs_sd*hpwt"],
      car_est_s2$se_rc["unobs_sd*air"],
      car_est_s2$se_rc["unobs_sd*mpd"],
      car_est_s2$se_rc["unobs_sd*space"],
      car_est_s2$se_rc["invinc*price"]
    ), 3)
  )
  
  kable(rc_table,
        col.names = c("Variable", "Spec 1 σ (unobs)", "SE",
                                  "Spec 2 σ or 1/y", "SE"),
        caption = "Nonlinear Parameters: Spec 1 (all unobs_sd) vs Spec 2 (unobs_sd + invinc*price)",
        row.names = FALSE)
```

```{r fit-table}
# Model fit comparison

  fit_table <- data.frame(
    Statistic = c("GMM objective", "Outer iterations"),
    Spec1 = c(round(car_est_s1$local_min, 2), 
              car_est_s1$outer_it),
    Spec2 = c(round(car_est_s2$local_min, 2),
              car_est_s2$outer_it)
  )
  
  kable(fit_table, caption = "Model Fit and Convergence")
```

**Comment on clustering:** We cluster standard errors by `model_id`
because the same car model appears in multiple years, making
observations within a model correlated over time. Not clustering would
understate standard errors.

**Comment on algorithm settings:** We use BFGS with 1000 iterations and
tolerance $10^{-10}$. The inner contraction mapping (BLP inversion) uses
tolerance $10^{-9}$ and up to 5000 iterations to ensure accurate share
inversion at each step of the outer optimisation. We use MLHS
integration with `r ndraws` draws, which balances accuracy against
computational cost.

**Comparison of specifications:** Both models produce negative price
effects, but they operate differently. Spec 1's $\hat{\sigma}_p$ on price is
`r round(car_est_s1$theta_rc["unobs_sd*price"], 2)`
(significant), so heterogeneity in price sensitivity is allowed but is not
tied to any observable demographic. Spec 2's `invinc*price` coefficient of
`r round(car_est_s2$theta_rc["invinc*price"], 2)` (significant) ties price 
sensiticity to inverse income, low-income consumers (high $1/y_i$) are much
more price-sensitive. This is the income channel from the BLP(1995). Spec 2 
also fits the moment conditions much better: the GMM objective is 
`r round(car_est_s2$local_min, 0)` versus `r round(car_est_s1$local_min, 0)`
for Spec 1. The non-price characteristic estimates back this up. Spec 1's 
`hpwt` of `r round(car_est_s1$theta_lin["hpwt", 1], 2)`) is hard to take 
seriously, while Spec 2's `r round(car_est_s2$theta_lin["hpwt", 1], 2)`) is 
in line with what we would expect.

\newpage

## Q2 (b) Elasticities, Marginal Costs and Markups for 1971

We use both estimated models to compute own- and cross-price
elasticities for a subset of 1971 cars spanning the price distribution,
then back out marginal costs and markups under multi-product Nash
pricing. The goal is to see how the two specifications differ in their
implied competitive structure.

**Step 1: Update Mean Utility and Check Model Consistency**

After estimation, `BLPestimatoR` returns the estimated mean utility
$\hat{\delta}_{jt}$. We store this back in each data object using
`update_BLP_data()` so that subsequent elasticity and share calculations
use the correct values.

As a check, both models should predict the same aggregate shares for any
given product, because the BLP inversion pins down $\hat{\delta}$ by
matching observed shares exactly.

```{r update-delta-s1}
# Spec 1: extract delta and write it back into the BLP data object

  delta_s1 <- car_est_s1$delta
  
  delta_data_s1 <- data.frame(
    "id"    = car_data_s1$parameters$product_id,
    "cdid"  = car_data_s1$parameters$market_id,
    "delta" = delta_s1
  )
  
  car_data_s1 <- update_BLP_data(
    data_update = delta_data_s1,
    blp_data    = car_data_s1
  )
```

```{r update-delta-s2}
# Spec 2: same process

  delta_s2 <- car_est_s2$delta
  
  delta_data_s2 <- data.frame(
    "id"    = car_data_s2$parameters$product_id,
    "cdid"  = car_data_s2$parameters$market_id,
    "delta" = delta_s2
  )
  
  car_data_s2 <- update_BLP_data(
    data_update = delta_data_s2,
    blp_data    = car_data_s2
  )
```

We now assemble the nonlinear parameter matrices for each model and
compute the consumer-level shares $s_{ij}$, which we will need for
elasticities.

```{r shareinfo-s1}
# Spec 1: unobs_sd on all 6 variables (including price)

  theta2_s1 <- matrix(NA, nrow = 6, ncol = 1)
  rownames(theta2_s1) <- c("const", "hpwt", "air", "mpd", "space", "price")
  colnames(theta2_s1) <- "unobs_sd"
  
  theta2_s1["const", "unobs_sd"] <- car_est_s1$theta_rc["unobs_sd*const"]
  theta2_s1["hpwt",  "unobs_sd"] <- car_est_s1$theta_rc["unobs_sd*hpwt"]
  theta2_s1["air",   "unobs_sd"] <- car_est_s1$theta_rc["unobs_sd*air"]
  theta2_s1["mpd",   "unobs_sd"] <- car_est_s1$theta_rc["unobs_sd*mpd"]
  theta2_s1["space", "unobs_sd"] <- car_est_s1$theta_rc["unobs_sd*space"]
  theta2_s1["price", "unobs_sd"] <- car_est_s1$theta_rc["unobs_sd*price"]

# Compute consumer-level shares s_ij for Spec 1

  shareObj_s1 <- getShareInfo(
    blp_data   = car_data_s1,
    par_theta2 = theta2_s1,
    printLevel = 0
  )
```

```{r shareinfo-s2}
# Spec 2: unobs_sd on 5 chars, invinc on price

  theta2_s2 <- matrix(NA, nrow = 6, ncol = 2)
  rownames(theta2_s2) <- c("const", "hpwt", "air", "mpd", "space", "price")
  colnames(theta2_s2) <- c("unobs_sd", "invinc")
  
  theta2_s2["const", "unobs_sd"] <- car_est_s2$theta_rc["unobs_sd*const"]
  theta2_s2["hpwt",  "unobs_sd"] <- car_est_s2$theta_rc["unobs_sd*hpwt"]
  theta2_s2["air",   "unobs_sd"] <- car_est_s2$theta_rc["unobs_sd*air"]
  theta2_s2["mpd",   "unobs_sd"] <- car_est_s2$theta_rc["unobs_sd*mpd"]
  theta2_s2["space", "unobs_sd"] <- car_est_s2$theta_rc["unobs_sd*space"]
  theta2_s2["price", "invinc"]   <- car_est_s2$theta_rc["invinc*price"]

# Compute consumer-level shares s_ij for Spec 2

  shareObj_s2 <- getShareInfo(
    blp_data   = car_data_s2,
    par_theta2 = theta2_s2,
    printLevel = 0
  )
```

```{r share-check}
# Sanity check: aggregate shares should match between models and the data

  share_check <- data.frame(
    observed = autoData$share,
    spec1    = shareObj_s1$shares,
    spec2    = shareObj_s2$shares
  )
  
# Assemble comparison table

  check_table <- data.frame(
    Comparison = c("Max |Spec 1 - Observed|", 
                   "Max |Spec 2 - Observed|", 
                   "Max |Spec 1 - Spec 2|"),
    Value = c(round(max(abs(share_check$spec1 - share_check$observed)), 8),
              round(max(abs(share_check$spec2 - share_check$observed)), 8),
              round(max(abs(share_check$spec1 - share_check$spec2)),    8))
  )
  
  kable(check_table, caption = "Share Inversion Sanity Check")
```

Now, we see all three differences are zero, both models successfully
invert to the observed shares, confirming the data objects were
correctly updated.

**Comment on starting values:** BLP estimation is sensitive
to starting values because the GMM objective can have multiple local
minima, especially when the model has several nonlinear parameters. We
experimented with several starting value vectors for each specification.

For Specification 2, we use starting values informed by BLP(1995) and
the classwork example: roughly (-14.12, 5.32, -3.32, -3.17,
-0.62) for the unobserved heterogeneity $\sigma_k$ on (const, hpwt, air,
mpd, space) and -25.3 for the income-price interaction. These values
converged to a GMM value of `r round(car_est_s2$local_min, 0)`
with the BLP inversion matching observed shares.

For Specification 1, initial attempts with uniform starting values (1,
1, 1, 1, 1, 1) produced numerical failures (`vmmin not finite` errors)
or convergence to inferior local minima where the inner BLP inversion
did not match observed shares. We therefore adopted starting values
informed by Spec 2's estimates for the non-price characteristics, roughly
(0.1, 3, 1, 1, 0.5, 0.5) for (price, const, hpwt, air,
mpd, space). These reach the lowest GMM objective we found
(`r round(car_est_s1$local_min, 0)`), with the BLP inversion
matching observed shares.

Even so, Spec 1's GMM objective remains substantially higher than Spec
2's, reflecting the difficulty of identifying a
symmetric normal random coefficient on price without observable
demographic information. This is one of the main a motivations for BLP(1995)'s
income-interaction specification.

**Step 2: Select an Interesting Subset of 1971 Products**

We choose products spanning the price distribution so we can visualise
how substitution patterns differ accross the price spectrum.

```{r subset-1971}
# Look at 1971 (cdid = 1) price distribution to pick representative cars

  autoData_1971 <- autoData[autoData$cdid == 1, ]
  
  summary(autoData_1971$price)

# Inspect cheapest and most expensive cars

  autoData_1971[order(autoData_1971$price), 
                c("id", "model_id", "price")] |> head(5)
  
  autoData_1971[order(autoData_1971$price), 
                c("id", "model_id", "price")] |> tail(5)
```

```{r pick-products}
# Pick a subset spanning the 1971 price distribution
# (update these IDs and labels after inspecting the output above)

  product_subset <- c("129", "150", "165", "138", "148")
  product_labels <- c("AMC_Gremlin", "Chevy_Vega", "Chevy_Impala", 
                      "Buick_Skylark", "Cadillac_Eldorado")
```

**Step 3: Compute Own- and Cross-Price Elasticities**

The diagonal entries are own-price elasticities; off-diagonal $(j, k)$
entries give the elasticity of product $j$'s share with respect to
product $k$'s price.

```{r elast-s1}
# Spec 1 elasticity matrix for chosen products in 1971

  Elas_s1 <- get_elasticities(
    blp_data    = car_data_s1,
    share_info  = shareObj_s1,
    theta_lin   = 0,
    variable    = "price",
    products    = product_subset,
    market      = "1"
  )
  
  colnames(Elas_s1) <- product_labels
  rownames(Elas_s1) <- product_labels
  
  kable(round(Elas_s1, 4), 
        caption = "Spec 1: Price Elasticity Matrix, 1971")
```

```{r elast-s2}
# Spec 2 elasticity matrix for chosen products in 1971

  Elas_s2 <- get_elasticities(
    blp_data    = car_data_s2,
    share_info  = shareObj_s2,
    theta_lin   = 0,
    variable    = "price",
    products    = product_subset,
    market      = "1"
  )
  
  colnames(Elas_s2) <- product_labels
  rownames(Elas_s2) <- product_labels
  
  kable(round(Elas_s2, 4), 
        caption = "Spec 2: Price Elasticity Matrix, 1971")
```

**Step 4: Back Out Marginal Costs and Markups (Multi-Product Nash)**

Under multi-product Nash pricing, each firm jointly chooses prices for
all its products. The vector of markups satisfies:

$$\mathbf{p} - \mathbf{mc} = -[\Omega \odot \mathbf{D}]^{-1} \mathbf{s}$$

where $\Omega$ is the ownership matrix ($\Omega_{jk} = 1$ if $j$ and $k$
are owned by the same firm), $\mathbf{D}$ is the matrix of share
derivatives with respect to prices, and $\odot$ is element-wise
multiplication.

```{r markup-function}
# Helper: compute markups in one market given a spec's data and shareObj

  compute_markups <- function(blp_data, shareObj, market_id_target, own_matrix) {
    
    mkt_ind   <- blp_data$parameters$market_id == market_id_target
    share_mkt <- shareObj$shares[mkt_ind]
    price_mkt <- blp_data$data$X_rand[mkt_ind, "price"]
    own_mkt   <- own_matrix[mkt_ind, ]
    own_prod  <- own_mkt %*% t(own_mkt)
    
    elast_mkt <- get_elasticities(
      blp_data    = blp_data,
      share_info  = shareObj,
      theta_lin   = 0,
      variable    = "price",
      market      = market_id_target,
      printLevel  = 0
    )
    
    scalar_mkt <- matrix(1 / share_mkt) %*% matrix(price_mkt, nrow = 1)
    deriv_mkt  <- elast_mkt / scalar_mkt
    
    markup_mkt <- c(-solve(t(deriv_mkt) * own_prod) %*% share_mkt)
    mc_mkt     <- price_mkt - markup_mkt
    
    return(list(price = price_mkt, markup = markup_mkt, mc = mc_mkt))
  }
```

```{r ownership-matrix}
# Ownership matrix (same for both specs)

  own_matrix <- as.matrix(
    car_data_s1$data$additional_data[, paste0("company", 1:26)]
  )
```

```{r markups-1971}
# Back out marginal costs and markups for 1971 under both specs

  result_s1 <- compute_markups(
    blp_data         = car_data_s1,
    shareObj         = shareObj_s1,
    market_id_target = "1",
    own_matrix       = own_matrix
  )
  
  result_s2 <- compute_markups(
    blp_data         = car_data_s2,
    shareObj         = shareObj_s2,
    market_id_target = "1",
    own_matrix       = own_matrix
  )
```

**Step 5: Compare High-Price vs Low-Price Cars**

```{r compare-models}
# Assemble comparison table for 1971

  compare_1971 <- data.frame(
    id             = autoData_1971$id,
    model_id       = autoData_1971$model_id,
    price          = result_s1$price,
    
    mc_s1          = result_s1$mc,
    markup_s1      = result_s1$markup,
    lerner_s1      = result_s1$markup / result_s1$price,
    
    mc_s2          = result_s2$mc,
    markup_s2      = result_s2$markup,
    lerner_s2      = result_s2$markup / result_s2$price
  )

# Split into low/high price halves at median

  price_median <- median(compare_1971$price)
  compare_1971$price_group <- ifelse(compare_1971$price < price_median, 
                                      "Low price", "High price")

# Summarise by price group

  summary_table <- compare_1971 |>
    group_by(price_group) |>
    summarise(
      mean_price     = round(mean(price),     3),
      mean_mc_s1     = round(mean(mc_s1),     3),
      mean_markup_s1 = round(mean(markup_s1), 3),
      mean_lerner_s1 = round(mean(lerner_s1), 3),
      mean_mc_s2     = round(mean(mc_s2),     3),
      mean_markup_s2 = round(mean(markup_s2), 3),
      mean_lerner_s2 = round(mean(lerner_s2), 3)
    )
  
# Transpose: make price groups the columns, statistics the rows

  summary_t <- summary_table |>
    column_to_rownames("price_group") |>
    t()
  
  rownames(summary_t) <- c("Mean Price", 
                           "Mean MC (Spec 1)", 
                           "Mean Markup (Spec 1)", 
                           "Mean Lerner (Spec 1)",
                           "Mean MC (Spec 2)", 
                           "Mean Markup (Spec 2)", 
                           "Mean Lerner (Spec 2)")
  
  kable(summary_t, caption = "1971 Markups and Marginal Costs by Price Group")
```

```{r markup-plots, fig.height=4, fig.width=9}
# Plot markups against price under each spec

  par(mfrow = c(1, 2))
  
  plot(compare_1971$price, compare_1971$markup_s1,
       main = "Spec 1: Markups vs Price",
       xlab = "Price ($000s)", ylab = "Markup ($000s)",
       pch  = 16, col = "blue")
  
  plot(compare_1971$price, compare_1971$markup_s2,
       main = "Spec 2: Markups vs Price",
       xlab = "Price ($000s)", ylab = "Markup ($000s)",
       pch  = 16, col = "skyblue")
```

**Comment on model differences:**

Specification 2 (income-interacted price) typically implies that
high-price cars face less elastic demand, because they are purchased by
high-income consumers whose price sensitivity is scaled down by $1/y_i$.
This shows up as higher markups and Lerner indices for expensive cars
under Spec 2.

Specification 1, by contrast, only introduces symmetric dispersion in
price sensitivity (via the normal random coefficient), and does not
systematically predict that luxury cars have more market power. If the
summary table shows a much larger gap between high- and low-price Lerner
indices under Spec 2 than Spec 1, this is the Spec 2 income effect
operating as BLP intended. The cross-price elasticity matrices also
reveal more realistic substitution patterns under Spec 2: luxury cars
substitute more with each other (both bought by rich consumers) than
with cheap cars.

\newpage

## Q2 (c) Effect of a 20% VAT on 1971 Prices

Impose a 20% VAT on cars, for 1971 only. Under VAT:

$$p_j^c = 1.2 \cdot p_j^f$$

where $p_j^c$ is the consumer price (what enters demand) and $p_j^f$ is
the producer price (what the firm sets and receives). The firm still
maximises profit over the price it controls, $p_j^f$, but consumer
demand now responds to $1.2 p_j^f$.

**How the FOC changes**

Without tax, the multi-product Nash FOC is:

$$\mathbf{p} - \mathbf{mc} = -[\Omega \odot \mathbf{D}(\mathbf{p})]^{-1} \mathbf{s}$$

With a 20% VAT, consumers respond to $p_j^c = 1.2 p_j^f$. By the chain
rule, share derivatives with respect to producer prices are $1.2$ times
the derivatives with respect to consumer prices. The FOC becomes:

$$\mathbf{p}^f - \mathbf{mc} = -[\Omega \odot (1.2 \cdot \mathbf{D}(\mathbf{p}^c))]^{-1} \mathbf{s}$$

We use `nleqslv()` to find the producer-price vector that zeros this
FOC, following the structure of Classwork 6.

```{r foc-vat-setup}
# FOC function: returns (pf - mc - markup), should equal 0 at equilibrium
# mkt_target = market id (numeric, e.g. 1 for 1971)

  foc_vat_mkt <- function(pf, blp_data, par_theta2, own_prod, 
                           mc_vec, mkt_target, vat = 0.20) {
    
    mkt_ind <- blp_data$parameters$market_id == mkt_target
    
    # Put consumer prices into the data (temporarily)
    
      pc <- (1 + vat) * pf
      blp_data$data$X_rand[mkt_ind, "price"] <- pc
    
    # Shares at these consumer prices
    
      tmp_share <- getShareInfo(
        blp_data   = blp_data,
        par_theta2 = par_theta2,
        printLevel = 0
      )
      share_mkt <- tmp_share$shares[mkt_ind]
    
    # Elasticities -> derivatives wrt consumer price
    
      elast_mkt <- get_elasticities(
        blp_data    = blp_data,
        share_info  = tmp_share,
        theta_lin   = 0,
        variable    = "price",
        market      = mkt_target,
        printLevel  = 0
      )
      
      scalar_mkt <- matrix(1 / share_mkt) %*% matrix(pc, nrow = 1)
      deriv_pc   <- elast_mkt / scalar_mkt
      deriv_pf   <- (1 + vat) * deriv_pc
    
    # Markup implied by multi-product Nash
    
      markup <- c(-solve(t(deriv_pf) * own_prod) %*% share_mkt)
    
    # FOC residual
    
      return(pf - mc_vec - markup)
  }
```

```{r vat-solve}
# Solve the FOC for each spec in 1971 using nleqslv

  mkt_target  <- 1
  mkt_ind_1971 <- car_data_s1$parameters$market_id == mkt_target
  own_mkt     <- own_matrix[mkt_ind_1971, ]
  own_prod_1  <- own_mkt %*% t(own_mkt)

# Spec 1

  pf_start_s1 <- result_s1$price  # start at observed prices
  
  sol_s1 <- nleqslv(
    x          = pf_start_s1,
    fn         = foc_vat_mkt,
    blp_data   = car_data_s1,
    par_theta2 = theta2_s1,
    own_prod   = own_prod_1,
    mc_vec     = result_s1$mc,
    mkt_target = mkt_target,
    vat        = 0.20
  )
  
# Spec 2

  pf_start_s2 <- result_s2$price
  
  sol_s2 <- nleqslv(
    x          = pf_start_s2,
    fn         = foc_vat_mkt,
    blp_data   = car_data_s2,
    par_theta2 = theta2_s2,
    own_prod   = own_prod_1,
    mc_vec     = result_s2$mc,
    mkt_target = mkt_target,
    vat        = 0.20
  )

# Report convergence

  conv_table <- data.frame(
    Specification = c("Spec 1", "Spec 2"),
    Convergence_code = c(sol_s1$termcd, sol_s2$termcd),
    Interpretation = c(
      
      # termcd = 1 indicates successful convergence to a solution of the FOC

        ifelse(sol_s1$termcd == 1, "Converged", "Did not fully converge"),
        ifelse(sol_s2$termcd == 1, "Converged", "Did not fully converge")
    )
  )
  
  kable(conv_table, caption = "nleqslv Convergence Codes (1 = success)")
```

**Consistency check**

We verify that solving the same system with `vat = 0` recovers the
observed prices, this is just to confirm taht the FOC is coded
correctly.

```{r vat-zero-check}
# Sanity check: VAT = 0 should return the observed prices

  check_s2 <- nleqslv(
    x          = pf_start_s2,
    fn         = foc_vat_mkt,
    blp_data   = car_data_s2,
    par_theta2 = theta2_s2,
    own_prod   = own_prod_1,
    mc_vec     = result_s2$mc,
    mkt_target = mkt_target,
    vat        = 0
  )
  
  cat("Max deviation (VAT=0 price vs observed):",
      round(max(abs(check_s2$x - autoData_1971$price)), 4), "\n")
```

**Compare prices before and after the tax**

```{r vat-results}
# Assemble 1971 price comparison

  vat_compare <- data.frame(
    id        = autoData_1971$id,
    model_id  = autoData_1971$model_id,
    p_orig    = autoData_1971$price,
    
    pf_s1     = sol_s1$x,
    pc_s1     = 1.20 * sol_s1$x,
    pct_pc_s1 = 100 * (1.20 * sol_s1$x - autoData_1971$price) / autoData_1971$price,
    
    pf_s2     = sol_s2$x,
    pc_s2     = 1.20 * sol_s2$x,
    pct_pc_s2 = 100 * (1.20 * sol_s2$x - autoData_1971$price) / autoData_1971$price
  )

# Split into low/high price halves at median

  price_median <- median(vat_compare$p_orig)
  vat_compare$price_group <- ifelse(vat_compare$p_orig < price_median, 
                                     "Low price", "High price")

# Summary by price group

  vat_summary <- vat_compare |>
    group_by(price_group) |>
    summarise(
      mean_p_orig    = round(mean(p_orig),     3),
      mean_pf_s1     = round(mean(pf_s1),      3),
      mean_pc_s1     = round(mean(pc_s1),      3),
      mean_pctpc_s1  = round(mean(pct_pc_s1),  2),
      mean_pf_s2     = round(mean(pf_s2),      3),
      mean_pc_s2     = round(mean(pc_s2),      3),
      mean_pctpc_s2  = round(mean(pct_pc_s2),  2)
    )

# Transpose for readability

  vat_t <- vat_summary |>
    column_to_rownames("price_group") |>
    t()
  
  rownames(vat_t) <- c("Mean original price",
                       "Mean producer price (Spec 1)",
                       "Mean consumer price (Spec 1)",
                       "Mean % change in consumer price (Spec 1)",
                       "Mean producer price (Spec 2)",
                       "Mean consumer price (Spec 2)",
                       "Mean % change in consumer price (Spec 2)")
  
  kable(vat_t, caption = "1971 Prices Before and After 20% VAT, by Price Group")
```

```{r vat-plots, fig.height=4, fig.width=9}
# % change in consumer price vs original price for each spec

  par(mfrow = c(1, 2))
  
  plot(vat_compare$p_orig, vat_compare$pct_pc_s1,
       main = "Spec 1: % Change in Consumer Price",
       xlab = "Original Price ($000s)", 
       ylab = "% Change in Consumer Price",
       pch  = 16, col = "red")
  abline(h = 20, col = "black", lty = 2)
  
  plot(vat_compare$p_orig, vat_compare$pct_pc_s2,
       main = "Spec 2: % Change in Consumer Price",
       xlab = "Original Price ($000s)", 
       ylab = "% Change in Consumer Price",
       pch  = 16, col = "red")
  abline(h = 20, col = "black", lty = 2)
```

The red dashed line at 20% marks full passthrough: points above the line indicate
overshifting (consumer prices rise by more than the tax rate); points
below mean firms absorb part of the tax.

**Comment:**

Under a 20% VAT, the two specifications give very different answers.

Under Spec 2, consumer-price increases range from about 5% to 30%, with 
a mean around 15%, and there is a clear upward trend in original price. 
High-priced cars see larger percentage increases than low-priced cars. 
The intuition: high-price cars are bought by high-income, price-insensitive 
consumers, so firms can pass more of the tax through without losing many 
customers. The commentators' worry that low-priced cars would bear a 
disproportionate burden is not borne out.

Spec 1 gives implausible numbers: percentage consumer-price changes 
reach several hundred percent for cheap cars, which is not a reasonable 
tax incidence. This reflects the identification problem from part (a). 
Spec 1's estimated $\sigma_p$ on price is essentially zero, so the
price-derivative matrix becomes near-singular and the FOC solver
cannot pin down an equilibrium. Under Spec 2, the regressive-burden 
concern is misplaced: it is high-priced cars that bear the larger 
percentage burden, not low-priced ones.
