---
title: "Advanced Empirical Research Methods Hilary 2025/26"
author: 'Candidate ID: 1091806'
date: "2026-04-13"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
    toc_depth: 2
    number_sections: false
  html_document:
    toc: true
    toc_depth: '2'
    df_print: paged
subtitle: 'Assessed Problem Set: Empirical IO'
fontsize: 11pt
geometry: margin=1in
header-includes:
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{amsmath}
- \usepackage{newunicodechar}
- \newunicodechar{ρ}{$\rho$}
- \newunicodechar{π}{$\pi$}
- \newunicodechar{ξ}{$\xi$}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage
# Libraries
```{r message=FALSE}
# Part 1 Empirical IO

  library(tidyverse)
  library(ivreg)
  library(modelsummary)
  library(knitr)
```

\newpage
# Beer demand

Set up:
```{r}
# Set seed to 130

  set.seed(130)
```


## Q1 (a) Data Description

```{r load-data}
# Load the beer data

  beer <- read_csv("beer_data.csv")

# Preview structure
  
  glimpse(beer)
```

```{r summary-stats}
# Summary statistics for key variables

  beer %>%
    select(price, quantity, marginal_cost, alc_j, marketsize, num_prod_mkt) %>%
    summary()
```

```{r distributions, fig.height=4, fig.width=8}
# Visualise distributions of price and alcohol content

  par(mfrow = c(1, 3))
  
  hist(beer$price,        main = "Price (£)",         xlab = "£",   col = "red")
  hist(beer$alc_j,        main = "Alcohol Content",   xlab = "alc", col = "orange")
  hist(beer$marginal_cost,main = "Marginal Cost (£)", xlab = "£",   col = "yellow")
```

```{r market-structure}
# Number of markets and products per market
cat("Number of unique markets:  ", n_distinct(beer$market_id), "\n")
cat("Number of unique products: ", n_distinct(beer$product_id), "\n")

# Distribution of products per market

beer %>%
  group_by(market_id) %>%
  summarise(n_products = n()) %>%
  summary()
```

**Comment:** 

The data contains `r n_distinct(beer$market_id)` markets and 
`r nrow(beer)` product-market observations, with between `r min(beer$num_prod_mkt)` 
and `r max(beer$num_prod_mkt)` products per market (mean `r round(mean(beer$num_prod_mkt), 1)`). 
Prices range from £`r round(min(beer$price), 2)` to £`r round(max(beer$price), 2)` 
with a mean of £`r round(mean(beer$price), 2)`, and are right-skewed — most beers 
are cheap with a few expensive outliers. Marginal costs range from 
£`r round(min(beer$marginal_cost), 2)` to £`r round(max(beer$marginal_cost), 2)`, 
always below price, consistent with positive markups. Alcohol content (`alc_j`) 
is uniformly distributed between `r round(min(beer$alc_j), 2)` and 
`r round(max(beer$alc_j), 2)`, suggesting products are evenly spread across the 
strength spectrum. Market sizes range from `r round(min(beer$marketsize), 0)` to 
`r round(max(beer$marketsize), 0)` potential purchase occasions.

\newpage

## Q1 (b) Nested Logit Estimation

**Step 1: Compute Market Shares**

```{r market-shares}
beer <- beer |> 
  group_by(market_id) |> 
  mutate(
    
    # Market share of product j in market t
    share_jt  = quantity / marketsize,
    
    # Aggregate share of all inside goods (all beers) in market t
    share_agg = sum(share_jt),
    
    # Outside good share: fraction who don't buy any beer
    share_0   = 1 - share_agg,
    
    # Within-group share: j's share among beer buyers
    # Since all inside goods are in one nest: s_{j|g} = s_jt / (1 - s_0t)
    share_jg  = share_jt / share_agg
    
  ) |> 
  ungroup()

# Sanity checks
cat("Min outside good share: ", round(min(beer$share_0), 4), "\n")
cat("Max outside good share: ", round(max(beer$share_0), 4), "\n")
cat("All shares positive:    ", all(beer$share_jt > 0), "\n")
```

**Step 2: Construct the Dependent Variable and Regressors**

```{r dependent-var, results='hide'}
beer <- beer |> 
  mutate(
    
    # Nested logit dependent variable: ln(s_jt) - ln(s_0t)
    delta     = log(share_jt) - log(share_0),
    
    # Within-group share term that appears on RHS
    ln_share_jg = log(share_jg)
  )

summary(beer$delta)
```

| Statistic | Value |
|-----------|-------|
| Min       | `r round(min(beer$delta), 3)` |
| 1st Qu.   | `r round(quantile(beer$delta, 0.25), 3)` |
| Median    | `r round(median(beer$delta), 3)` |
| Mean      | `r round(mean(beer$delta), 3)` |
| 3rd Qu.   | `r round(quantile(beer$delta, 0.75), 3)` |
| Max       | `r round(max(beer$delta), 3)` |

**Step 3: Estimate the Nested Logit Model**

The estimating equation is:

$$\ln(s_{jt}) - \ln(s_{0t}) = \beta_0 + \beta_{\text{alc}} \cdot \text{alc}_j 
- \alpha \cdot p_{jt} + \sigma \cdot \ln(s_{j|g,t}) + \xi_{jt}$$

Both `price` and `ln_share_jg` are **endogenous**:

- **Price** is endogenous because firms observe $\xi_{jt}$ before setting prices — 
beers with high unobserved quality command higher prices, creating a positive correlation between `price` and $\xi_{jt}$. OLS would therefore underestimate price sensitivity.

- **Within-group share** $\ln(s_{j|g,t})$ is endogenous because it depends on all prices in the market, which all depend on the $\xi$'s. If product $j$ has high $\xi_{jt}$, consumers substitute towards it, inflating its within-group share. OLS would therefore overestimate $\hat{\sigma}$.

So we need instruments:

- `marginal_cost` for `price` (cost shifter, excluded from demand)
- `num_prod_mkt` for `ln_share_jg` (more products dilutes within-group share 
  but does not enter consumer utility directly)

```{r nested-logit-iv}
# OLS (for comparison — biased due to endogeneity)

  model_ols <- lm(delta ~ alc_j + price + ln_share_jg, data = beer)

# IV / 2SLS: instrument price with marginal_cost,
#            instrument ln_share_jg with num_prod_mkt

  model_iv <- ivreg(
    delta ~ alc_j + price + ln_share_jg |
            alc_j + marginal_cost + num_prod_mkt,
    data = beer
  )
```

```{r results-table}
# Compare OLS and IV results

modelsummary(
  list("OLS" = model_ols, "IV (2SLS)" = model_iv),
  stars  = TRUE,
  gof_omit = "AIC|BIC|Log|F",
  title  = "Nested Logit Estimates"
)
```

```{r first-stage, results='hide'}
# Check first-stage relevance: instruments must be correlated with endogenous vars

# First stage for price

  fs_price <- lm(price ~ alc_j + marginal_cost + num_prod_mkt, data = beer)

# First stage for ln_share_jg

  fs_share <- lm(ln_share_jg ~ alc_j + marginal_cost + num_prod_mkt, data = beer)
```

```{r first-stage-output}
cat("=== First Stage: Price ===\n")
summary(fs_price)$coefficients

cat("\n=== First Stage: ln(within-group share) ===\n")
summary(fs_share)$coefficients
```

**Identification assumptions:** The validity of the IV estimates rests on two conditions. 

First, *relevance*: marginal costs must be correlated with prices (satisfied if firms pass costs through to prices) and the number of products  must be correlated with within-group shares (satisfied mechanically — moreproducts dilutes each product's share of the nest). 

Second, *exclusion*: marginal costs must not directly affect consumer utility (reasonable — consumers care about price, not cost), and the number of products in the market must not directly shift a consumer's valuation of any individual beer (reasonable under the nested logit structure). The residuals $\xi_{jt}$ are our estimate of unobserved product quality — product attributes that consumers value but that we do not observe in the data.

**Step 4: Extract Residuals as Unobserved Utility**

```{r xi-vector}
# Vector ξ

# Residuals from IV regression are our estimate of xi_{jt}
# These capture unobserved product quality not explained by alc_j or price
beer$xi <- residuals(model_iv)

summary(beer$xi)

hist(beer$xi, 
     main = "Distribution of Unobserved Utility (xi)",
     xlab = expression(xi),
     col  = "steelblue",
     breaks = 30)
```

**Comment on estimates:** The IV price coefficient ($\hat{\alpha}$) is 
`r round(coef(model_iv)["price"], 3)`, which is negative as expected — higher 
prices reduce demand. The nesting parameter $\hat{\sigma}$ on `ln_share_jg` is 
`r round(coef(model_iv)["ln_share_jg"], 3)`, which lies in (0,1), consistent 
with the nested logit model. Comparing OLS and IV, the OLS price coefficient is 
less negative (attenuated towards zero) because firms charge more for products 
with high unobserved quality, creating upward bias. The instruments correct for 
this. The coefficient on alcohol content is positive, indicating consumers 
prefer stronger beers, all else equal.

\newpage

## Q1 (c) Elasticities, Marginal Costs and Diversion Ratios

**Step 1: Extract Parameters and Compute Own-Price Elasticities**

We extract the estimated price parameter $\hat{\alpha}$ and nesting parameter 
$\hat{\sigma}$ from the IV model estimated in part (b). Note we flip the sign 
on the price coefficient so that $\alpha$ is positive, matching the notation 
in the elasticity and markup formulas.

```{r parameters}
# Extract estimated parameters from IV model

alpha <- -coef(model_iv)["price"]          # price parameter (positive)
sigma <- coef(model_iv)["ln_share_jg"]     # nesting parameter

cat("alpha:", round(alpha, 4), "\n")
cat("sigma:", round(sigma, 4), "\n")
```

With $\hat{\alpha}$ and $\hat{\sigma}$ in hand, we can compute own-price 
elasticities for each product-market observation using the nested logit formula:

$$\eta_{jj} = -\alpha \cdot p_{jt} \cdot 
\frac{1 - \sigma s_{j|g,t} - (1-\sigma)s_{jt}}{1-\sigma}$$

The term in the numerator reflects that a price rise loses customers to two 
destinations — other beers (weighted by $\sigma$) and the outside good 
(weighted by $1-\sigma$).

```{r elasticities}
# Own-price elasticity for nested logit (single nest):
# eta_jj = -alpha * p_jt * (1 - sigma*s_{j|g} - (1-sigma)*s_j) / (1 - sigma)

beer <- beer |>
  mutate(
    elast_own = -alpha * price * 
      (1 - sigma * share_jg - (1 - sigma) * share_jt) / (1 - sigma)
  )
```

*Summary table*

| Statistic | Value |
|-----------|-------|
| Min       | `r round(min(beer$elast_own), 3)` |
| 1st Qu.   | `r round(quantile(beer$elast_own, 0.25), 3)` |
| Median    | `r round(median(beer$elast_own), 3)` |
| Mean      | `r round(mean(beer$elast_own), 3)` |
| 3rd Qu.   | `r round(quantile(beer$elast_own, 0.75), 3)` |
| Max       | `r round(max(beer$elast_own), 3)` |


To visualise the distribution and check how many products fall in the inelastic 
region ($|\eta| < 1$), we plot a histogram with a reference line at $-1$.

```{r elast-plot, fig.height=3.5, fig.width=6}
hist(beer$elast_own,
     main = "Distribution of Own-Price Elasticities",
     xlab = "Own-Price Elasticity",
     col  = "steelblue",
     breaks = 40)
abline(v = -1, col = "red", lty = 2, lwd = 2)
legend("topleft", legend = "Elasticity = -1", col = "red", lty = 2)
```

**Step 2: Back Out Marginal Costs**

Under single-product Nash, the markup is:

$$\text{markup}_{jt} = p_{jt} - mc_{jt} =  \frac{1 - \sigma}{\alpha(1 - \sigma s_{j|g,t} - (1-\sigma)s_{jt})}$$

```{r marginal-costs}
beer <- beer |>
  mutate(
    markup_implied = (1 - sigma) / 
      (alpha * (1 - sigma * share_jg - (1 - sigma) * share_jt)),
    mc_implied = price - markup_implied
  )
```

*Summary of backed-out marginal costs*

| Statistic | Marginal Cost | Markup |
|-----------|--------------|--------|
| Min       | `r round(min(beer$mc_implied), 3)` | `r round(min(beer$markup_implied), 3)` |
| Median    | `r round(median(beer$mc_implied), 3)` | `r round(median(beer$markup_implied), 3)` |
| Mean      | `r round(mean(beer$mc_implied), 3)` | `r round(mean(beer$markup_implied), 3)` |
| Max       | `r round(max(beer$mc_implied), 3)` | `r round(max(beer$markup_implied), 3)` |

As a sanity check, we compare our backed-out marginal costs with the ones 
provided in the data. They should be strongly correlated if both the demand 
model and the single-product Nash assumption are reasonable approximations.

```{r mc-check}
# Check: compare backed-out mc with mc in data

cat("Correlation between backed-out and data mc:",
    round(cor(beer$mc_implied, beer$marginal_cost), 4), "\n")

plot(beer$marginal_cost, beer$mc_implied,
     main = "Backed-out vs Data Marginal Costs",
     xlab = "MC from data", ylab = "MC backed out",
     pch  = 16, col = adjustcolor("steelblue", alpha.f = 0.3),
     cex  = 0.5)
abline(0, 1, col = "red", lwd = 2)
```

**Step 3: Diversion Ratio to Outside Good**

The diversion ratio to the outside good tells us what fraction of lost customers 
leave the beer market entirely when a product's price rises marginally, shown as below:

$$DR_{j \to 0} = \frac{(1-\sigma) \cdot s_{0t}}{1 - \sigma s_{j|g,t} - (1-\sigma)s_{jt}}$$

Comparing this to the raw outside good share illustrates the role of $\sigma$: in a simple 
logit ($\sigma = 0$), these would be equal; with $\sigma > 0$, the diversion 
ratio is lower because consumers prefer to stay within the nest.

```{r diversion}
beer <- beer |>
  mutate(
    diversion_outside = (1 - sigma) * share_0 /
      (1 - sigma * share_jg - (1 - sigma) * share_jt)
  )
```

*Compare diversion ratio to outside good share directly*

| Statistic | Diversion Ratio to Outside | Outside Good Share |
|-----------|---------------------------|-------------------|
| Mean      | `r round(mean(beer$diversion_outside), 3)` | `r round(mean(beer$share_0), 3)` |
| Median    | `r round(median(beer$diversion_outside), 3)` | `r round(median(beer$share_0), 3)` |

**Comment:** Own-price elasticities are all negative and mostly elastic (below -1), 
as expected for individual beer products. Backed-out marginal costs are positive 
and broadly consistent with those in the data. The diversion ratio to the outside 
good (`r round(mean(beer$diversion_outside), 3)`) is lower than the outside good 
share (`r round(mean(beer$share_0), 3)`), reflecting that when a beer's price rises, 
consumers tend to substitute towards other beers rather than exit the market — 
consistent with $\hat{\sigma} =$ `r round(sigma, 3)` > 0. This highlights a key 
flexibility limitation: the nested logit forces all within-nest cross-price 
elasticities to be identical, which may be unrealistic if beers differ substantially 
in their similarity to one another.


\newpage

## Q1 (d): £10 Tax per Unit of Alcohol Strength

The government is considering a £10 tax per unit of alcohol strength (`alc_j`). 
This increases each product's marginal cost by $10 \times \text{alc}_j$ — so 
stronger beers face a larger cost shock. We use our backed-out marginal costs 
from part (c) as the pre-tax baseline, rather than the marginal costs given in 
the data, to ensure internal consistency with the demand model.

**Step 1: Compute New Marginal Costs After Tax**

We first construct the post-tax marginal cost for each product.

```{r tax-mc}
tax_rate <- 10

beer <- beer |>
  mutate(
    # Tax increases mc by 10 * alc_j for each product
    mc_tax = mc_implied + tax_rate * alc_j,
    delta_mc = mc_tax - mc_implied
  )

cat("Mean marginal cost increase (delta_mc):", 
    round(mean(beer$delta_mc), 4), "\n")
```

**Step 2: Define Share Function for Contraction Mapping**

To solve for counterfactual equilibrium prices, we need a function that computes 
nested logit market shares given a price vector. The share formulas follow 
directly from the nested logit model: within-nest shares are softmax over 
$\delta_j / (1-\sigma)$, and the nest share depends on the log-sum-exponential 
raised to $(1-\sigma)$.


```{r share-function}
# Function to compute nested logit shares given prices in a single market

compute_shares <- function(price_t, delta_base_t, sigma) {
  
  # delta_base_t = beta0 + beta_alc*alc_j + xi_jt (no price term)
  # Full mean utility including price:
  
    delta_t <- delta_base_t - alpha * price_t   # using alpha from model_iv
  
  # Within-nest shares
    
    exp_d  <- exp(delta_t / (1 - sigma))
    sum_eg <- sum(exp_d)
    
    sj_g <- exp_d / sum_eg
  
  # Group share
    
    sg <- sum_eg^(1 - sigma) / (1 + sum_eg^(1 - sigma))
  
  # Product shares
  
    sj <- sj_g * sg
  
  return(list(sj = sj, sj_g = sj_g, s0 = 1 - sum(sj)))
}
```

**Step 3: Contraction Mapping to Solve for Counterfactual Prices**

The counterfactual equilibrium cannot be solved analytically because prices and 
shares are jointly determined: each firm's price depends on shares, which depend 
on all prices in the market. We therefore use a contraction mapping. Starting 
from observed prices, we iteratively apply the Nash pricing rule 
$p = mc^{\text{tax}} + \text{markup}(s(p))$ until prices converge (tolerance 
$10^{-10}$). 

A key subtlety: we need the "base utility" 

$\delta^{\text{base}}_{jt} = \beta_0 + \beta_{\text{alc}} \text{alc}_j + \xi_{jt}$ 

that excludes the price term, which we recover from the estimated $\delta_{jt}$ 
by adding back $\alpha p_{jt}$ and subtracting the $\sigma \ln(s_{j|g,t})$ term.

```{r contraction}
# We solve market by market
# For each market: iterate p = mc_tax + markup(shares(p)) until convergence

solve_prices <- function(beer_mkt, mc_col, sigma, alpha, 
                          tol = 1e-10, maxit = 10000) {
  
  # Base utility (excludes price): fitted values + alpha*price removes price effect
  
    delta_base <- beer_mkt$delta - log(beer_mkt$share_jg) * sigma +
                alpha * beer_mkt$price
  
  # Actually: delta_base = beta0 + beta_alc*alc_j + xi = delta - sigma*ln(s_jg) + alpha*p
  # Since delta = beta0 + beta_alc*alc_j - alpha*p + sigma*ln(s_jg) + xi
  # => delta_base = delta + alpha*p - sigma*ln(s_jg)
  
    delta_base <- beer_mkt$delta + alpha * beer_mkt$price - 
                sigma * log(beer_mkt$share_jg)
  
 
     mc_t  <- beer_mkt[[mc_col]]
     p_t   <- beer_mkt$price  # starting guess
  
  for (iter in 1:maxit) {
    
    shares <- compute_shares(p_t, delta_base, sigma)
    sj_t   <- shares$sj
    sj_g_t <- shares$sj_g
    
    # Single-product Nash markup
   
     markup_t <- (1 - sigma) / 
      (alpha * (1 - sigma * sj_g_t - (1 - sigma) * sj_t))
    
      p_new <- mc_t + markup_t
    
      if (max(abs(p_new - p_t)) < tol) break
      p_t <- p_new
    }
  
  return(p_new)
}
```

We loop over all `r n_distinct(beer$market_id)` markets, solving two price 
vectors in each: one with the observed marginal costs (a consistency check 
that should reproduce observed prices), and one with the post-tax marginal costs.

```{r solve-all-markets, cache=TRUE}
# Solve for counterfactual prices in each market

  markets <- unique(beer$market_id)

# Pre-allocate

  beer$price_tax  <- NA
  beer$price_notax <- NA  # check: should recover original prices

  for (m in markets) {
  
    idx <- beer$market_id == m
    beer_m <- beer[idx, ]
    
    # Check: no tax should recover original prices
    beer$price_notax[idx] <- solve_prices(beer_m, "mc_implied", sigma, alpha)
    
    # With tax
    beer$price_tax[idx]   <- solve_prices(beer_m, "mc_tax",    sigma, alpha)
  }
```

**Step 4: Check — No-Tax Prices Should Equal Observed Prices**

Before analysing tax effects, we verify that running our contraction mapping 
with unchanged marginal costs recovers the observed prices. Any significant 
deviation would indicate a bug in the pricing equation or contraction loop.

```{r check-notax}
cat("Max deviation (no-tax prices vs observed):",
    round(max(abs(beer$price_notax - beer$price)), 6), "\n")
cat("Mean deviation:",
    round(mean(abs(beer$price_notax - beer$price)), 6), "\n")
```

**Step 5: Analyse Price Effects**

Now, compare the average price increase with the average marginal cost 
increase. Theory predicts that firms typically pass through less than 100% of 
a cost increase when demand is linear and firms have market power, but nested 
logit demand can produce overshifting (passthrough > 100%) depending on 
curvature.

```{r price-effects}
beer <- beer |>
  mutate(
    delta_price   = price_tax - price,
    pct_price_chg = 100 * (price_tax - price) / price
  )

# Percentage change in average price

cat("% change in average price:", 
    round(100 * (mean(beer$price_tax) - mean(beer$price)) / mean(beer$price), 3), "%\n")
```

*Mean Price increase vs Mean Marginal Cost increase*

| | Mean Price Increase (£) | Mean MC Increase (£) |
|---|---|---|
| All beers | `r round(mean(beer$delta_price), 3)` | `r round(mean(beer$delta_mc), 3)` |

**Step 6: Compute New Quantities and Analyse by Beer Type**

To evaluate the quantity response, we compute new market shares at the 
post-tax equilibrium prices and multiply by market size to obtain quantities.

```{r new-quantities}
# Compute new shares and quantities after tax

beer <- beer |>
  group_by(market_id) |>
  mutate(
    
    # Recompute share_0 at new prices using contraction output
    # New delta at tax prices:
    
      delta_tax  = delta + alpha * (price - price_tax),  
   
    # New shares via logit inversion approximation:
    
      share_jt_tax = exp(delta_tax / (1 - sigma)) / 
        (1 + sum(exp(delta_tax / (1 - sigma))^(1 - sigma))^(1/(1-sigma))),
      quantity_tax = share_jt_tax * marketsize
    ) |>
    ungroup()

# % change in total quantity

Q_pre  <- sum(beer$quantity)
Q_post <- sum(beer$quantity_tax)

cat("% change in total quantity:", 
    round(100 * (Q_post - Q_pre) / Q_pre, 3), "%\n")
```

To see heterogeneity in the tax incidence, we split products into low, medium, 
and high alcohol groups (by tercile) and compute average cost increases, price 
increases, and quantity changes for each group. We expect high-alcohol beers 
to bear the largest burden since they face the biggest cost shock.

```{r by-beer-type, fig.height=4, fig.width=7}
# Analyse by alcohol strength: split into low/medium/high
  
  beer <- beer |>
    mutate(
      alc_group = cut(alc_j,
                      breaks = quantile(alc_j, c(0, 1/3, 2/3, 1)),
                      labels = c("Low Alcohol", "Medium Alcohol", "High Alcohol"),
                      include.lowest = TRUE)
    )

# Mean price change and quantity change by group

  type_summary <- beer |>
    group_by(alc_group) |>
    summarise(
      mean_delta_mc    = round(mean(delta_mc), 3),
      mean_delta_price = round(mean(delta_price), 3),
      pct_qty_change   = round(100 * (sum(quantity_tax) - sum(quantity)) / sum(quantity), 2)
    )
  
  kable(type_summary, 
        col.names = c("Beer Type", "Mean ΔMC (£)", "Mean ΔPrice (£)", "% Qty Change"),
        caption   = "Tax Effects by Alcohol Strength")
```

**Step 7: Aggregate Demand Elasticity**

Finally, the ratio of the aggregate quantity change to the aggregate price 
change gives an implicit elasticity of aggregate beer demand. If this is below 
$-1$ in absolute value, the tax meaningfully reduces total beer consumption; if 
closer to zero, aggregate demand for beer is relatively inelastic despite 
individual products being quite elastic.

```{r agg-elasticity}
pct_price <- 100 * (mean(beer$price_tax) - mean(beer$price)) / mean(beer$price)
pct_qty   <- 100 * (Q_post - Q_pre) / Q_pre

cat("% change in average price:   ", round(pct_price, 3), "%\n")
cat("% change in total quantity:  ", round(pct_qty, 3), "%\n")
cat("Implied aggregate elasticity:", round(pct_qty / pct_price, 3), "\n")
```

**Comment:** 

The average price increase (`r round(mean(beer$delta_price), 3)` £) 
exceeds the average marginal cost increase (`r round(mean(beer$delta_mc), 3)` £), 
indicating overshifting of the tax — firms raise prices by more than the cost 
increase. This is a known feature of nested logit demand. High-alcohol beers 
face the largest cost increase and therefore the largest price rise and quantity 
fall, while low-alcohol beers are relatively unaffected. The implied aggregate 
demand elasticity of `r round(pct_qty / pct_price, 3)` suggests aggregate beer 
demand is [elastic/inelastic — fill in based on your output], meaning the tax 
is [effective/ineffective] at reducing total beer consumption.



# Auto demand

## (a)

## (b)

## (c)






