---
title: "AERM Exam Outputs and Results"
output:
  pdf_document: default
  html_document: default
date: "2026-03-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

Use Julia code


# Exercise 1: Solve Standard Consumption-Saving Problem for Different Values of $\rho$

## Q1: Stationary Income Distribution

Income follows a two-state Markov chain with symmetric persistence $\rho_L = \rho_H = \rho$ and transition matrix:

$$P = \begin{pmatrix} \rho & 1-\rho \\ 1-\rho & \rho \end{pmatrix}$$

Starting from $P(y = y_L) = 0.2$, the figure below shows the share of low-income households over a 100-year horizon for each value of $\rho$.

```{r, echo=FALSE, fig.align='center', out.width='85%'}
knitr::include_graphics("/Users/caisuyan/Documents/GitHub/AERM/Advanced ERM Problem Set/MT Examed PS/results/fig_ex1_q1.pdf")
```

*Explanation*:

For all $\rho \in (0,1)$, the share of low-income households converges to 0.5, which is the unique stationary distribution of this symmetric Markov chain, but higher persistence slows the rate of convergence.

The two edge cases behave differently: $\rho = 0$ causes full state-switching each period (so the share oscillates before settling), while $\rho = 1$ makes both states absorbing, so the initial distribution $P(y_L) = 0.2$ is permanent.

## Q2: Volatility vs. Risk

*Explanation*:

In the stationary distribution, the variance of $\Delta \log y$ equals $(1-\rho) \cdot (\log 2)^2$, which is maximized at $\rho = 0$ where every household switches state each period. 

However, $\rho = 0$ does not generate the most risk for households: with i.i.d. income, bad shocks are transitory and easily self-insured through savings, whereas high persistence (e.g. $\rho = 0.95$) means adverse shocks last many periods, making them far harder to smooth.

## Q3: Simulation vs. Continuum

```{r, echo=FALSE, fig.align='center', out.width='85%'}
knitr::include_graphics("../results/fig_ex1_q3.pdf")
```

The simulated share of high-income types closely tracks the continuum prediction but does NOT match it exactly, because a finite sample of 1000 households is subject to sampling variation whereas the continuum result holds by the law of large numbers over a unit mass of agents.


## Q4: OLS vs. Fixed Effects

Using the simulated panel data, I regress current income on lagged income via OLS and fixed effects (implemented by demeaning), using only linear algebra as instructed.

| Model | $\hat{\beta}$ |
|-------|--------------|
| OLS   | 0.8983       |
| FE    | 0.5690       |
| True $\rho$ | 0.95  |

OLS is closer to the true $\rho = 0.95$ because the data-generating process contains no individual fixed effects — all households face the same Markov chain — so OLS is consistent. The FE estimator suffers from Nickell bias: with only $T = 8$ periods, demeaning mechanically creates a negative correlation between the demeaned lagged variable and the demeaned error, biasing $\hat{\beta}_{FE}$ substantially downward. This bias is systematic and would shrink only as $T \to \infty$.




