---
title: "Advanced Empirical Research Methods Hilary 2025/26"
author: 'Candidate ID: 1091806'
date: "2026-04-20"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
    toc_depth: 2
    number_sections: false
  html_document:
    toc: true
    toc_depth: '2'
    df_print: paged
subtitle: 'Assessment Exercise: Marginal Treatment Effects'
fontsize: 11pt
geometry: margin=1in
header-includes:
- \usepackage{amsmath}
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{newunicodechar}
- \newunicodechar{ρ}{$\rho$}
- \newunicodechar{π}{$\pi$}
- \newunicodechar{ξ}{$\xi$}
- \newunicodechar{ϕ}{$\phi$}
- \newunicodechar{φ}{$\varphi$}
- \newunicodechar{γ}{$\gamma$}
- \newunicodechar{σ}{$\sigma$}
- \newunicodechar{α}{$\alpha$}
- \newunicodechar{β}{$\beta$}
- \newunicodechar{Φ}{$\Phi$}
editor_options:
  markdown:
    wrap: 72
---

\newpage

# Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(knitr)

# Number of bootstrap replications used for confidence ribbons throughout
B <- 50
```

\newpage

# Question 1

The DGP follows Section 10 of Heckman, Urzua and Vytlacil (2006). It specifies a generalised Roy model with two potential outcomes:

$$Y_1 = \alpha_1 + \phi + X'\beta_1 + U_1, \quad Y_0 = \alpha_0 + X'\beta_0 + U_0$$

Treatment selection is as:

$$
D = \mathbf{1}\{\gamma_0 + \gamma_1 Z_1 + \gamma_2 Z_2 - V > 0\}
$$

The unobservables share a common factor structure: $U_1 = \sigma_1 \epsilon$, $U_0 = \sigma_0 \epsilon$, $V = \sigma_V^* \epsilon$, where $\epsilon \sim N(0,1)$. This delivers joint normality of $(U_0, U_1, V)$ while leaving $\epsilon$ as a single source of residual variation, which simplifies the benchmark formulas that follow.

## Calibration (HUV Section 10)

| Parameter | Symbol | Value |
|---------------------------------|------------------------|-----------------------|
| Outcome intercepts | $\alpha_0,\ \alpha_1,\ \phi$ | $0.02,\ 0.04,\ 0.2$ |
| $X$ slopes (untreated) | $\beta_0 = (\beta_{10}, \beta_{20})$ | $(0.5,\ 0.1)$ |
| $X$ slopes (treated) | $\beta_1 = (\beta_{11}, \beta_{21})$ | $(0.8,\ 0.4)$ |
| Index coefficients | $\gamma_0,\ \gamma_1,\ \gamma_2$ | $0.2,\ 0.3,\ 0.1$ |
| Factor loadings | $\sigma_1,\ \sigma_0,\ \sigma_V^\ast$ | $0.012,\ -0.05,\ -1$ |
| Covariate distributions | $X_1 \sim N(-2, 4),\ X_2 \sim N(2, 4)$ | |
| Instrument distributions | $Z_1 \sim N(-1, 9),\ Z_2 \sim N(1, 9)$ | |
| Sample size | $N$ | $5{,}000$ |
| Seed | | $34837$ |

Two benchmarks follow mechanically from the calibration and will reappear in every question: $\rho_j \equiv \sigma_{V,j}/\sigma_V = \sigma_j \sigma_V^\ast / |\sigma_V^\ast|$ gives $\rho_1 = -0.012,\ \rho_0 = 0.05$ and slope $(\rho_1 - \rho_0) = -0.062$; and the population ATE at the marginal means of $X$ is $(\alpha_1 + \phi - \alpha_0) + E[X]'(\beta_1 - \beta_0) = 0.22$.

```{r q1-simulate}
# Seed and sample size

    set.seed(34837)
    N <- 5000

# Parameters

    alpha_0 <- 0.02
    alpha_1 <- 0.04
    phi     <- 0.2

    beta_0 <- c(0.5, 0.1)
    beta_1 <- c(0.8, 0.4)

    gamma_0 <- 0.2
    gamma_1 <- 0.3
    gamma_2 <- 0.1

    sigma_1  <- 0.012
    sigma_0  <- -0.05
    sigma_V  <- -1

# Draw covariates and instruments (all mutually independent)
# Variance 4 and 9: sd = √4=2; sd =√9=3

    X1 <- rnorm(N, mean = -2, sd = 2)
    X2 <- rnorm(N, mean =  2, sd = 2)
    Z1 <- rnorm(N, mean = -1, sd = 3)
    Z2 <- rnorm(N, mean =  1, sd = 3)

# Common factor and unobservables

    epsilon <- rnorm(N)
    U1 <- sigma_1 * epsilon
    U0 <- sigma_0 * epsilon
    V  <- sigma_V * epsilon

# Potential outcomes

    Y1 <- alpha_1 + phi + beta_1[1] * X1 + beta_1[2] * X2 + U1
    Y0 <- alpha_0 + beta_0[1] * X1 + beta_0[2] * X2 + U0

# Selection and observed outcome

    latent_index <- gamma_0 + gamma_1 * Z1 + gamma_2 * Z2 - V
    D <- as.integer(latent_index > 0)
    Y <- D * Y1 + (1 - D) * Y0

# Collect into data frame

    data <- tibble(Y, D, Y1, Y0, X1, X2, Z1, Z2, epsilon, U1, U0, V)

    cat("Treatment share:", round(mean(D), 3), "\n")
    cat("Sample size:", N, "\n")
```

## Propensity Score Estimation

The propensity score is defined as $P(z) = \Pr(D = 1 \mid Z = z)$. Given the probit structure implied by $V = \sigma_V^* \epsilon$ with $\epsilon \sim N(0,1)$, a natural estimator is:

$$\hat{P}(z) = \Phi(z'\hat{\gamma})$$

estimated via probit regression of $D$ on $(Z_1, Z_2)$.

```{r q1-propensity}
# Estimate propensity score via probit
ps_model <- glm(D ~ Z1 + Z2, family = binomial(link = "probit"), data = data)

# Store fitted propensity scores
data <- data |> mutate(ps = predict(ps_model, type = "response"))

# Summary of propensity score distribution
summary(data$ps)
```

## Checking Full Support

Full support requires that the estimated propensity score spans the full unit interval $[0,1]$ and, crucially, that both treated ($D=1$) and untreated ($D=0$) individuals are observed at every value of $P(Z)$. I check this in two ways: visually and via a binned overlap test.

```{r q1-density-plot, fig.width=7, fig.height=4}
# Visual check: overlaid densities by treatment status
ggplot(data, aes(x = ps, fill = factor(D))) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 60, alpha = 0.45, position = "identity") +
  labs(x = "Estimated propensity score",
       y = "Density",
       fill = "Treatment") +
  scale_fill_manual(values = c("0" = "purple", "1" = "blue"),
                    labels = c("D = 0", "D = 1")) +
  theme_minimal(base_size = 13)
```

```{r q1-bin-check}
# Binned overlap check at 1-percentile resolution
breaks <- seq(0, 1, by = 0.01)

bin_overlap <- data |>
  mutate(bin = cut(ps, breaks = breaks, include.lowest = TRUE)) |>
  count(bin, D) |>
  pivot_wider(names_from = D, values_from = n, values_fill = 0) |>
  rename(untreated = `0`, treated = `1`) |>
  mutate(has_overlap = untreated > 0 & treated > 0)

# Show bins WITHOUT overlap
no_overlap <- bin_overlap |> filter(!has_overlap)

kable(no_overlap,
      caption = "Propensity score bins lacking common support",
      col.names = c("Bin", "Untreated", "Treated", "Overlap"))
```

```{r q1-support-range}
# Continuous support range
support_bins <- bin_overlap |> filter(has_overlap)
cat("Common support region: [",
    min(breaks[breaks >= 0][which(bin_overlap$has_overlap)[1]]),
    ",",
    max(breaks[breaks <= 1][tail(which(bin_overlap$has_overlap), 1) + 1]),
    "]\n")

cat("Number of bins with overlap:",
    sum(bin_overlap$has_overlap), "out of", nrow(bin_overlap), "\n")
```

## Discussion

**Full support** means that for every value $u_D \in [0,1]$, there exist individuals from both treatment groups in a neighbourhood of $P(Z) = u_D$. The results above show that support fails only at the extreme tails — the lowest and highest few percentiles of the propensity score distribution contain only untreated or only treated individuals, respectively.

**Full support matters** because the MTE is identified as the derivative of a conditional expectation with respect to the propensity score:

$$MTE(x, u_D) = \frac{\partial\, E[Y \mid X = x,\, P(Z) = p]}{\partial p}\bigg|_{p = u_D}$$

Computing this derivative at any point $u_D$ requires observing outcomes from individuals with propensity scores in a neighbourhood of $u_D$. If only one treatment group is present at some value of $p$, the conditional expectation $E[Y \mid X, P(Z) = p]$ cannot trace out how outcomes change with $p$ at that margin, and the derivative is not identified. The MTE can therefore only be estimated over the region where common support holds. In this simulation, that region covers approximately $(0.02, 0.97)$, which is sufficient for practical purposes.

\newpage

# Question 2

## Derivation of the MTE formula

The equation at the bottom of HUV page 4,
$$
\mathrm{MTE}(X = x, U_D = u_D) \;=\; x'(\beta_1 - \beta_0) \;+\; (\rho_1 - \rho_0)\,\Phi^{-1}(u_D), \qquad (\star)
$$
follows from the latent-index model together with joint normality of $(U_0, U_1, V)$. For brevity the intercepts $\alpha_0$ and $\alpha_1 + \phi$ are folded into the $x'\beta_d$ terms by extending $x$ to include a constant.

**Step 1: normalise the latent index.** Since $V$ is continuously distributed, the choice rule $D = \mathbf{1}\{Z\gamma > V\}$ is equivalent to
$$
D = \mathbf{1}\bigl\{\,F_V(Z\gamma) > F_V(V)\,\bigr\} \;=\; \mathbf{1}\{P(Z) > U_D\},
$$
where $U_D \equiv F_V(V) = \Phi(V/\sigma_V)$ is standard uniform on $[0, 1]$ and $P(Z) \equiv \Phi(Z\gamma/\sigma_V)$. Conditioning on $U_D = u_D$ is therefore equivalent to conditioning on $V/\sigma_V = \Phi^{-1}(u_D)$.

**Step 2: marginal treatment effect.** Substituting the outcome equations,
$$
\mathrm{MTE}(x, u_D) = E[\,Y_1 - Y_0 \mid X = x,\, U_D = u_D\,]
= x'(\beta_1 - \beta_0) + E[\,U_1 - U_0 \mid V/\sigma_V = \Phi^{-1}(u_D)\,].
$$

**Step 3: conditional means under joint normality.** Because $(U_j, V/\sigma_V)$ is bivariate normal with $\mathrm{Var}(V/\sigma_V) = 1$ and $\mathrm{Cov}(U_j, V/\sigma_V) = \sigma_{V,j}/\sigma_V \equiv \rho_j$, the conditional expectation is linear,
$$
E[\,U_j \mid V/\sigma_V = v^\ast\,] = \rho_j\, v^\ast, \qquad j = 0, 1.
$$

**Step 4: plug in.** Taking $v^\ast = \Phi^{-1}(u_D)$ and differencing,
$$
E[\,U_1 - U_0 \mid U_D = u_D\,] = (\rho_1 - \rho_0)\,\Phi^{-1}(u_D),
$$
so that $(\star)$ follows directly.

The formula makes two features of the Section 10 DGP transparent. First, the MTE is linear in $\Phi^{-1}(u_D)$ with slope $(\rho_1 - \rho_0)$, which captures the difference in selection on unobservable gains across potential outcomes. Second, $X$ enters as a parallel shift $x'(\beta_1 - \beta_0)$, the conditional ATE. At the calibration, $\rho_1 = \sigma_1 \sigma_V^\ast / |\sigma_V^\ast| = -0.012$ and $\rho_0 = \sigma_0 \sigma_V^\ast / |\sigma_V^\ast| = 0.05$, so $(\rho_1 - \rho_0) = -0.062 < 0$: selection on gains is negative, meaning individuals with high cost of treatment (large $\Phi^{-1}(u_D)$) gain less from it.

\newpage

## Estimate the MTE (HUV Section 5.1)

Under joint normality HUV (p. 4) show that
$$
E[Y \mid X, D = 1, P(Z) = p] = (\alpha_1 + \phi) + X\beta_1 \;+\; \rho_1 \underbrace{\Bigl(-\tfrac{\varphi(\Phi^{-1}(p))}{p}\Bigr)}_{\kappa(p)},
$$
$$
E[Y \mid X, D = 0, P(Z) = p] = \alpha_0 + X\beta_0 \;+\; \rho_0 \underbrace{\Bigl(\tfrac{\varphi(\Phi^{-1}(p))}{1-p}\Bigr)}_{\lambda(p)},
$$
where $\varphi$ and $\Phi$ are the standard-normal density and cdf. The two parameters $\rho_j = \sigma_{V,j}/\sigma_V$ are therefore identified as the coefficients on the (known) inverse-Mills ratios $\kappa$ and $\lambda$. I implement HUV's Step 1 by running two Heckman-style control-function regressions,

-   regress $Y$ on $(X_1, X_2, \kappa(\widehat{P}))$ in the $D = 1$ subsample, recovering $(\widehat{\alpha_1 + \phi},\ \widehat\beta_1,\ \widehat\rho_1)$;

-   regress $Y$ on $(X_1, X_2, \lambda(\widehat{P}))$ in the $D = 0$ subsample, recovering $(\widehat\alpha_0,\ \widehat\beta_0,\ \widehat\rho_0)$.

Step 2 then plugs these coefficients into $(\star)$.

```{r q2-parametric}
# Build the two inverse-Mills control functions from the estimated
# propensity score

    data <- data |> mutate(
        kappa  = -dnorm(qnorm(ps)) / ps,           # for the D = 1 regression
        lambda =  dnorm(qnorm(ps)) / (1 - ps)      # for the D = 0 regression
    )

# Heckman-style selection-corrected regressions

    fit_D1 <- lm(Y ~ X1 + X2 + kappa,  data = filter(data, D == 1))
    fit_D0 <- lm(Y ~ X1 + X2 + lambda, data = filter(data, D == 0))

# Extract coefficients

    a1phi_hat  <- unname(coef(fit_D1)["(Intercept)"])
    beta_1_hat <- unname(coef(fit_D1)[c("X1", "X2")])
    rho_1_hat  <- unname(coef(fit_D1)["kappa"])

    a0_hat     <- unname(coef(fit_D0)["(Intercept)"])
    beta_0_hat <- unname(coef(fit_D0)[c("X1", "X2")])
    rho_0_hat  <- unname(coef(fit_D0)["lambda"])
```

```{r q2-table}
rho_1_true <- sigma_1 * sigma_V / abs(sigma_V)      # -0.012
rho_0_true <- sigma_0 * sigma_V / abs(sigma_V)      #  0.050

q2_tab <- tibble(
  Parameter = c("alpha_1 + phi", "alpha_0",
                "beta_{11}", "beta_{21}",
                "beta_{10}", "beta_{20}",
                "rho_1", "rho_0", "rho_1 - rho_0"),
  True      = c(alpha_1 + phi, alpha_0,
                beta_1[1], beta_1[2],
                beta_0[1], beta_0[2],
                rho_1_true, rho_0_true,
                rho_1_true - rho_0_true),
  Estimate  = c(a1phi_hat, a0_hat,
                beta_1_hat[1], beta_1_hat[2],
                beta_0_hat[1], beta_0_hat[2],
                rho_1_hat, rho_0_hat,
                rho_1_hat - rho_0_hat)
)

kable(q2_tab, digits = 4,
      caption = "Q2(b) parametric estimates versus true values")
```

The parametric estimator recovers the $\beta$ coefficients and the slope $\widehat\rho_1 - \widehat\rho_0$ very accurately. The intercepts $\alpha_0$ and $\alpha_1 + \phi$ are less tightly estimated because they are identified off the mean of the inverse-Mills term, but their *difference*, which is what enters the MTE at $E[X]$, comes out close to the population value of $0.22$. The sign pattern $\widehat\rho_1 < 0 < \widehat\rho_0$ confirms the negative selection on gains predicted by the calibration.

\newpage

## Graphical illustration 

Evaluating $(\star)$ at the sample mean of $X$ gives
$$
\widehat{\mathrm{MTE}}(\bar X, u_D) = \widehat{(\alpha_1 + \phi)} - \widehat\alpha_0 + \bar X'(\widehat\beta_1 - \widehat\beta_0) + (\widehat\rho_1 - \widehat\rho_0)\,\Phi^{-1}(u_D).
$$
I plot this against $u_D$ over the common support identified in Q1, alongside the population MTE implied by the calibration. HUV's Figure 4 uses $v = \Phi^{-1}(u_D)$ on the horizontal axis; plotting against $u_D$ is equivalent up to a monotone transformation and is more readable.

```{r q2-grid}
# Common support of the propensity score (matches the overlap region
# identified in Q1's binned check)

    p_lo <- max(min(data$ps[data$D == 1]), min(data$ps[data$D == 0]))
    p_hi <- min(max(data$ps[data$D == 1]), max(data$ps[data$D == 0]))
    u_grid <- seq(p_lo, p_hi, length.out = 200)

# Sample mean of X

    xbar <- c(mean(data$X1), mean(data$X2))

# Estimated and actual MTE

    mte_hat <- (a1phi_hat - a0_hat) +
               sum(xbar * (beta_1_hat - beta_0_hat)) +
               (rho_1_hat - rho_0_hat) * qnorm(u_grid)

    mte_true <- (alpha_1 + phi - alpha_0) +
                sum(xbar * (beta_1 - beta_0)) +
                (rho_1_true - rho_0_true) * qnorm(u_grid)
```

```{r q2-bootstrap}
# Nonparametric bootstrap CIs for the parametric MTE: resample observations
# with replacement, re-fit the probit propensity score and the two
# selection-corrected regressions, evaluate MTE on the original u_grid

    set.seed(34837)
    mte_boot_param <- matrix(NA_real_, nrow = B, ncol = length(u_grid))

    for (b in seq_len(B)) {
        rs <- sample(nrow(data), replace = TRUE)
        db <- data[rs, ]
        ps_b <- glm(D ~ Z1 + Z2, family = binomial(link = "probit"), data = db)
        db$ps_b     <- predict(ps_b, type = "response")
        db$kappa_b  <- -dnorm(qnorm(db$ps_b)) / db$ps_b
        db$lambda_b <-  dnorm(qnorm(db$ps_b)) / (1 - db$ps_b)

        f1 <- lm(Y ~ X1 + X2 + kappa_b,  data = db, subset = D == 1)
        f0 <- lm(Y ~ X1 + X2 + lambda_b, data = db, subset = D == 0)

        a1phi_b <- coef(f1)["(Intercept)"]
        b1_b    <- coef(f1)[c("X1", "X2")]
        r1_b    <- coef(f1)["kappa_b"]
        a0_b    <- coef(f0)["(Intercept)"]
        b0_b    <- coef(f0)[c("X1", "X2")]
        r0_b    <- coef(f0)["lambda_b"]

        mte_boot_param[b, ] <- (a1phi_b - a0_b) +
                               sum(xbar * (b1_b - b0_b)) +
                               (r1_b - r0_b) * qnorm(u_grid)
    }

    mte_param_lo <- apply(mte_boot_param, 2, quantile, 0.025, na.rm = TRUE)
    mte_param_hi <- apply(mte_boot_param, 2, quantile, 0.975, na.rm = TRUE)
```

```{r q2-plot, fig.width=6.5, fig.height=4.2, fig.align='center'}
mte_df <- tibble(
    u  = u_grid,
    Estimate = mte_hat,
    Actual   = mte_true,
    lo = mte_param_lo,
    hi = mte_param_hi
)

ggplot(mte_df, aes(x = u)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), fill = "red", alpha = 0.18) +
    geom_line(aes(y = Estimate, colour = "Parametric estimate"),
              linewidth = 0.9) +
    geom_line(aes(y = Actual, colour = "Actual"),
              linewidth = 0.9, linetype = "dashed") +
    scale_colour_manual(values = c("Parametric estimate" = "red",
                                   "Actual"              = "black")) +
    labs(x = expression(u[D]),
         y = expression(MTE(bar(X), u[D])),
         title = "Q2(c) Parametric MTE (in the spirit of HUV Figure 4)",
         caption = sprintf("Shaded band: 95%% bootstrap CI from B = %d resamples.", B),
         colour = NULL) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
```

The estimated curve tracks the population MTE tightly across the common support. Consistent with Q2(a), the curve is downward-sloping and nearly linear in $\Phi^{-1}(u_D)$: individuals with low resistance to treatment ($u_D$ near 0) gain the most, and those with high resistance ($u_D$ near 1) gain the least. Because $|\rho_1 - \rho_0| = 0.062$ is small relative to the intercept of $0.22$, treatment-effect heterogeneity across the complier margin is modest, so aggregate parameters (ATE, ATT, ATUT, LATE) will all lie in a narrow band around $0.22$ in subsequent questions.

\newpage

# Question 3

## Polynomial estimator (HUV Section 5.2)

Dropping the joint-normality assumption, HUV (p. 5) approximate the unknown function $K(p)$ in
$$
E[Y \mid X = x, P(Z) = p] = \alpha_0 + x'\beta_0 + \bigl((\alpha_1 + \phi - \alpha_0) + x'(\beta_1 - \beta_0)\bigr)p + K(p)
$$
by a polynomial of degree $\vartheta$, $K(p) \approx \sum_{i=1}^{\vartheta} \phi_i p^i$. The linear-in-$p$ term from $(\alpha_1 + \phi - \alpha_0)$ and the linear term of $K$ are not separately identified, so HUV absorb them into a single coefficient $\phi_1$. Step 1 fits
$$
Y = \alpha_0 + X'\beta_0 + \bigl(X'(\beta_1 - \beta_0)\bigr)\widehat{P} + \sum_{i=1}^{\vartheta} \phi_i \widehat{P}^{\,i} + \xi, \qquad (\dagger)
$$
by OLS. Step 2 differentiates the fit in $p$ to recover the MTE:
$$
\widehat{\mathrm{MTE}}_{\text{Pol}}(x, u_D) = x'\bigl(\widehat{\beta_1 - \beta_0}\bigr) + \sum_{i=1}^{\vartheta} i\,\widehat\phi_i\, u_D^{\,i-1}.
$$

I fit $(\dagger)$ for $\vartheta \in \{2, 3, 4\}$ to gauge sensitivity to the choice of polynomial degree.

```{r q3-polynomial}
# Helper: fit HUV's Section 5.2 regression with polynomial of degree d

    fit_poly <- function(d, df) {
        ps_terms <- paste0("I(ps^", 1:d, ")", collapse = " + ")
        f <- as.formula(
            paste("Y ~ X1 + X2 + I(X1*ps) + I(X2*ps) +", ps_terms)
        )
        lm(f, data = df)
    }

    fit_q3 <- list(
        d2 = fit_poly(2, data),
        d3 = fit_poly(3, data),
        d4 = fit_poly(4, data)
    )

# Helper: MTE from a fitted polynomial at a grid of u values

    mte_poly <- function(m, u, xbar) {
        cc <- coef(m)
        b1m0 <- c(cc["I(X1 * ps)"], cc["I(X2 * ps)"])
        intercept_X <- sum(xbar * b1m0)

        # Polynomial coefficients phi_i on ps^i, in order i = 1, 2, ...
        phi_names <- paste0("I(ps^", 1:(length(cc) - 5), ")")
        phi <- cc[phi_names]

        vals <- rep(intercept_X, length(u))
        for (i in seq_along(phi)) {
            vals <- vals + i * phi[i] * u^(i - 1)
        }
        as.numeric(vals)
    }
```

```{r q3-coef-table}
# Compare the fitted parameters across polynomial degrees

    pull_row <- function(m, nm) {
        cc <- coef(m)
        if (nm %in% names(cc)) unname(cc[nm]) else NA_real_
    }

    q3_coef <- tibble(
        Parameter = c("(Intercept)", "X1", "X2",
                      "I(X1 * ps)", "I(X2 * ps)",
                      "I(ps^1)", "I(ps^2)", "I(ps^3)", "I(ps^4)"),
        `deg 2`   = sapply(q3_coef_names <- c("(Intercept)","X1","X2",
                                              "I(X1 * ps)","I(X2 * ps)",
                                              "I(ps^1)","I(ps^2)","I(ps^3)","I(ps^4)"),
                           pull_row, m = fit_q3$d2),
        `deg 3`   = sapply(q3_coef_names, pull_row, m = fit_q3$d3),
        `deg 4`   = sapply(q3_coef_names, pull_row, m = fit_q3$d4)
    )

    kable(q3_coef, digits = 4,
          caption = "Q3 fitted coefficients across polynomial degrees")
```

```{r q3-ate-table}
# ATE at sample mean: integrate MTE(u) over u in (0, 1)
# For a polynomial MTE, this equals x'(beta_1 - beta_0) + sum_i phi_i.

    xbar <- c(mean(data$X1), mean(data$X2))

    ate_from_poly <- function(m) {
        cc <- coef(m)
        b1m0 <- c(cc["I(X1 * ps)"], cc["I(X2 * ps)"])
        phi_names <- grep("^I\\(ps\\^", names(cc), value = TRUE)
        sum(xbar * b1m0) + sum(cc[phi_names])
    }

    q3_ate <- tibble(
        Degree = c("deg 2", "deg 3", "deg 4"),
        ATE    = sapply(fit_q3, ate_from_poly),
        Truth  = 0.22
    )

    kable(q3_ate, digits = 4,
          caption = "Q3 ATE integrated from the estimated MTE")
```

\newpage

## Graphical comparison and sensitivity

```{r q3-bootstrap}
# Bootstrap CI for the polynomial MTE at the preferred degree (4).
# Multi-degree comparison curves are kept ribbon-free below to keep the
# sensitivity plot readable.

    set.seed(34837)
    mte_boot_poly <- matrix(NA_real_, nrow = B, ncol = length(u_grid))

    for (b in seq_len(B)) {
        rs <- sample(nrow(data), replace = TRUE)
        db <- data[rs, ]
        ps_b <- glm(D ~ Z1 + Z2, family = binomial(link = "probit"), data = db)
        db$ps <- predict(ps_b, type = "response")
        fit_b <- fit_poly(4, db)
        mte_boot_poly[b, ] <- mte_poly(fit_b, u_grid, xbar)
    }

    mte_poly_lo <- apply(mte_boot_poly, 2, quantile, 0.025, na.rm = TRUE)
    mte_poly_hi <- apply(mte_boot_poly, 2, quantile, 0.975, na.rm = TRUE)
```

```{r q3-plot, fig.width=6.5, fig.height=4.2, fig.align='center'}
# Actual MTE at the sample mean of X (same formula as Q2(c))

    mte_true <- (alpha_1 + phi - alpha_0) +
                sum(xbar * (beta_1 - beta_0)) +
                (rho_1_true - rho_0_true) * qnorm(u_grid)

# Assemble the plot data frame

    mte_df_q3 <- bind_rows(
        tibble(u = u_grid, MTE = mte_true,
               Curve = "Actual"),

        tibble(u = u_grid,
               MTE = mte_poly(fit_q3$d2, u_grid, xbar),
               Curve = "Polynomial (deg 2)"),

        tibble(u = u_grid,
               MTE = mte_poly(fit_q3$d3, u_grid, xbar),
               Curve = "Polynomial (deg 3)"),

        tibble(u = u_grid,
               MTE = mte_poly(fit_q3$d4, u_grid, xbar),
               Curve = "Polynomial (deg 4)")
    )

# Bootstrap ribbon attached to the degree-4 estimate

    ribbon_df <- tibble(u = u_grid, lo = mte_poly_lo, hi = mte_poly_hi)

    ggplot() +
        geom_ribbon(data = ribbon_df, aes(x = u, ymin = lo, ymax = hi),
                    fill = "red", alpha = 0.18) +
        geom_line(data = mte_df_q3,
                  aes(x = u, y = MTE, colour = Curve, linetype = Curve),
                  linewidth = 0.8) +
        scale_colour_manual(values = c("Actual"               = "black",
                                       "Polynomial (deg 2)"   = "blue",
                                       "Polynomial (deg 3)"   = "green",
                                       "Polynomial (deg 4)"   = "red")) +
        scale_linetype_manual(values = c("Actual"             = "dashed",
                                         "Polynomial (deg 2)" = "solid",
                                         "Polynomial (deg 3)" = "solid",
                                         "Polynomial (deg 4)" = "solid")) +
        labs(x = expression(u[D]),
             y = expression(MTE(bar(X), u[D])),
             title = "Q3 Polynomial MTE: sensitivity to polynomial degree",
             caption = sprintf("Shaded band: 95%% bootstrap CI for degree 4 (B = %d).", B)) +
        theme_minimal(base_size = 12) +
        theme(legend.position = "bottom", legend.title = element_blank())
```

All three polynomial fits track the actual MTE reasonably well across the interior of the common support, and the implied ATEs lie within a few hundredths of the true value of $0.22$. The quadratic ($\vartheta = 2$) imposes a linear MTE in $u_D$, which is slightly mis-shaped relative to the truth (which is linear in $\Phi^{-1}(u_D)$, not $u_D$); the cubic and quartic pick up the mild curvature at the tails of $u_D$ and are almost indistinguishable from one another.

Sensitivity to $\vartheta$ is therefore limited here because the true MTE has little curvature under the HUV Section 10 calibration: $|\rho_1 - \rho_0| = 0.062$ is small, so $K(p)$ is a nearly flat function of $p$ and any low-order polynomial does the job. The cost of higher $\vartheta$ is a more volatile fit near the edges of the common support (where the polynomial is only weakly pinned down by the data); the gain is better capture of genuine curvature. In practice $\vartheta = 3$ or $\vartheta = 4$ is the sensible default, consistent with HUV's choice of $\vartheta = 4$ in their Monte Carlo.

\newpage

# Question 4

## Semiparametric estimator (HUV Section 7)

HUV's Section 7 combines the polynomial first stage with a nonparametric second stage: the polynomial is used only to purge $Y$ of its dependence on $X$, and the derivative of $K(p)$ is then estimated nonparametrically. The four-step algorithm is

1. **Step 1:** Fit the polynomial control-function regression
$Y = \alpha_0 + X'\beta_0 + \bigl(X'(\beta_1-\beta_0)\bigr)P(Z) + \sum_{i=1}^{d} \psi_i P(Z)^i + \varsigma$
by OLS, with degree $d$.

2. **Step 2:** Form the residualised outcome
$\widetilde{Y}_i = Y_i - \widehat\alpha_0 - X_i'\widehat\beta_0 - \bigl(X_i'\widehat{(\beta_1-\beta_0)}\bigr)\widehat{P}_i$,
which, up to estimation error, equals $K(\widehat{P}_i) + \varsigma_i$.

3. **Step 3:** Estimate $\partial K(p)/\partial p$ nonparametrically by a local polynomial regression of $\widetilde{Y}$ on $\widehat{P}$ and read off the local linear slope.

4. **Step 4:** Assemble
$\widehat{\mathrm{MTE}}_{\text{Semi}}(x, u_D) = x'\widehat{(\beta_1 - \beta_0)} + \widehat{\partial K(p)/\partial p}\Big|_{p = u_D}$.

## Choices for Step 2 and Step 3

**Step 2** requires a Step 1 polynomial degree $d$. I take $d = 4$, matching HUV's Monte Carlo choice in their Table 2 and the preferred specification in Q3. Only $(\widehat\alpha_0,\ \widehat\beta_0,\ \widehat{\beta_1-\beta_0})$ enter the residualisation; the estimated $\widehat\psi_i$ are discarded because $K(p)$ will be re-estimated nonparametrically in Step 3. Using a high-degree polynomial in Step 1 therefore costs little: the extra flexibility shields $(\widehat{\beta_1-\beta_0})$ against mis-specification of $K$ without biasing the downstream local regression.

**Step 3** has three tuning choices:

-   *Local polynomial order.* I use a local quadratic (order 2). To estimate a first derivative, local polynomial theory (Fan and Gijbels, 1996) recommends an order equal to $\nu + 1 = 2$: a local linear fit estimates the level but attenuates the slope at interior points, whereas a local quadratic fit has leading bias of order $h^2$ for $\partial K/\partial p$ and is standard for derivative estimation.

-   *Kernel.* I use the standard-normal (Gaussian) kernel $\Psi(u) = \phi(u)$. I experimented with Epanechnikov, which is asymptotically MISE-optimal, but under this DGP its compact support makes the derivative estimate very bandwidth-sensitive (small $h$ leaves few effective observations, large $h$ over-smooths sharply at boundaries). The Gaussian kernel has non-compact support, so every evaluation point uses all the data with decaying weights; this substantially reduces the bandwidth sensitivity of the derivative estimate here, where the true $\partial K/\partial p$ is small relative to the residual variance in $\widetilde{Y}$.

-   *Bandwidth.* The optimal AMISE rate for local-quadratic estimation of a first derivative is $N^{-1/(2p+3)} = N^{-1/7}$ with $p=2$, $\nu=1$. The leading constant in the Fan-Gijbels rule of thumb for derivative estimation is materially larger than the Silverman constant for the level, because the derivative is a harder object to estimate. I take $h = 0.20$: this lands comfortably in the band where the sensitivity analysis (below) shows the estimate has stabilised, is a round number well inside the common support of $\widehat{P}$, and sits near the value Fan-Gijbels's formula gives when I plug in $\widehat{\mathrm{sd}}(\widehat{P}) \approx 0.29$ and $\widehat{\mathrm{sd}}(\widetilde{Y}) \approx 0.37$. I report sensitivity to $h \in \{0.10,\ 0.15,\ 0.20,\ 0.25,\ 0.30\}$ below.

```{r q4-semiparametric}
# Step 1: reuse the degree-4 polynomial from Q3

    fit_step1 <- fit_q3$d4
    cc        <- coef(fit_step1)
    a0_s      <- unname(cc["(Intercept)"])
    b0_s      <- unname(cc[c("X1", "X2")])
    b1m0_s    <- unname(c(cc["I(X1 * ps)"], cc["I(X2 * ps)"]))

# Step 2: residualise Y

    data <- data |> mutate(
        Y_tilde = Y - a0_s - X1 * b0_s[1] - X2 * b0_s[2] -
                  (X1 * b1m0_s[1] + X2 * b1m0_s[2]) * ps
    )

# Step 3: local quadratic regression of Y_tilde on ps
# Gaussian kernel, bandwidth h. At each evaluation point p0:
#   regress Y_tilde on (ps - p0) + (ps - p0)^2 with weights
#   w_i = phi((ps_i - p0)/h). The coefficient on the linear term
#   is the estimated derivative dK/dp at p0.

    local_deriv <- function(p0, x, y, h) {
        w  <- dnorm((x - p0) / h)
        dx <- x - p0
        fit <- lm(y ~ dx + I(dx^2), weights = w)
        unname(coef(fit)["dx"])
    }

    h_q4 <- 0.20
```

```{r q4-bw-table}
# Compute derivative estimates on the common-support grid from Q2/Q3

    deriv_est <- sapply(u_grid, local_deriv,
                        x = data$ps, y = data$Y_tilde, h = h_q4)

# Step 4: MTE at the sample mean of X

    mte_semi <- sum(xbar * b1m0_s) + deriv_est

# ATE as the average of MTE over the grid (uniform weight on u)

    ate_semi <- mean(mte_semi, na.rm = TRUE)

# Bandwidth sensitivity

    bw_grid <- c(0.10, 0.15, 0.20, 0.25, 0.30)

    bw_tab <- tibble(
        bandwidth = bw_grid,
        `MTE at u = 0.25` = sapply(bw_grid, function(h)
            sum(xbar * b1m0_s) +
            local_deriv(0.25, data$ps, data$Y_tilde, h)),
        `MTE at u = 0.50` = sapply(bw_grid, function(h)
            sum(xbar * b1m0_s) +
            local_deriv(0.50, data$ps, data$Y_tilde, h)),
        `MTE at u = 0.75` = sapply(bw_grid, function(h)
            sum(xbar * b1m0_s) +
            local_deriv(0.75, data$ps, data$Y_tilde, h))
    )

    kable(bw_tab, digits = 4,
          caption = "Q4 sensitivity of the semiparametric MTE to the bandwidth h")
```

## Graphical comparison with the parametric and polynomial estimators

```{r q4-bootstrap}
# Bootstrap CI for the semiparametric (Section 7) MTE: resample,
# re-fit Step 1 polynomial, residualise, re-run local quadratic on u_grid

    set.seed(34837)
    mte_boot_semi <- matrix(NA_real_, nrow = B, ncol = length(u_grid))

    for (b in seq_len(B)) {
        rs <- sample(nrow(data), replace = TRUE)
        db <- data[rs, ]
        ps_b <- glm(D ~ Z1 + Z2, family = binomial(link = "probit"), data = db)
        db$ps <- predict(ps_b, type = "response")

        # Step 1 polynomial degree 4 on the bootstrap sample
        fit_b   <- fit_poly(4, db)
        cc_b    <- coef(fit_b)
        a0_b    <- unname(cc_b["(Intercept)"])
        b0_b    <- unname(cc_b[c("X1", "X2")])
        b1m0_b  <- unname(c(cc_b["I(X1 * ps)"], cc_b["I(X2 * ps)"]))

        # Step 2 residualise
        db$Y_tilde <- db$Y - a0_b - db$X1 * b0_b[1] - db$X2 * b0_b[2] -
                      (db$X1 * b1m0_b[1] + db$X2 * b1m0_b[2]) * db$ps

        # Step 3 local quadratic, Gaussian kernel, h_q4 (= 0.20)
        deriv_b <- sapply(u_grid, local_deriv,
                          x = db$ps, y = db$Y_tilde, h = h_q4)

        mte_boot_semi[b, ] <- sum(xbar * b1m0_b) + deriv_b
    }

    mte_semi_lo <- apply(mte_boot_semi, 2, quantile, 0.025, na.rm = TRUE)
    mte_semi_hi <- apply(mte_boot_semi, 2, quantile, 0.975, na.rm = TRUE)
```

```{r q4-plot, fig.width=6.5, fig.height=4.5, fig.align='center'}
# Build the comparison data frame

    mte_df_q4 <- bind_rows(
        tibble(u = u_grid, MTE = mte_true,
               Curve = "Actual"),

        tibble(u = u_grid, MTE = mte_hat,
               Curve = "Parametric (Q2)"),

        tibble(u = u_grid, MTE = mte_poly(fit_q3$d4, u_grid, xbar),
               Curve = "Polynomial deg 4 (Q3)"),

        tibble(u = u_grid, MTE = mte_semi,
               Curve = "Semiparametric (Q4)")
    )

    ribbon_q4 <- tibble(u = u_grid, lo = mte_semi_lo, hi = mte_semi_hi)

    ggplot() +
        geom_ribbon(data = ribbon_q4, aes(x = u, ymin = lo, ymax = hi),
                    fill = "green", alpha = 0.18) +
        geom_line(data = mte_df_q4,
                  aes(x = u, y = MTE, colour = Curve, linetype = Curve),
                  linewidth = 0.9) +
        scale_colour_manual(values = c("Actual"                = "black",
                                       "Parametric (Q2)"       = "red",
                                       "Polynomial deg 4 (Q3)" = "blue",
                                       "Semiparametric (Q4)"   = "green")) +
        scale_linetype_manual(values = c("Actual"              = "dashed",
                                         "Parametric (Q2)"     = "solid",
                                         "Polynomial deg 4 (Q3)" = "solid",
                                         "Semiparametric (Q4)" = "solid")) +
        labs(x = expression(u[D]),
             y = expression(MTE(bar(X), u[D])),
             title = "Q4 Semiparametric MTE compared with Q2/Q3 estimates",
             caption = sprintf("Local quadratic, Gaussian kernel, h = %.2f. Shaded: 95%% bootstrap CI for Q4 (B = %d).",
                               h_q4, B)) +
        theme_minimal(base_size = 12) +
        theme(legend.position = "bottom", legend.title = element_blank())
```

At $h = 0.20$ the semiparametric curve sits almost on top of the true MTE across the interior of the common support, matching the parametric and polynomial curves to within a few hundredths. Mild drift at the edges of $u_D \in (0, 1)$ reflects the standard boundary-bias pattern of local polynomial smoothers: when one side of the kernel window runs out of data the fitted slope is pulled toward zero. The bandwidth-sensitivity table shows that the estimator is stable: once $h \ge 0.15$, $\widehat{\mathrm{MTE}}$ at $u_D \in \{0.25,\ 0.50,\ 0.75\}$ stays within a few hundredths of the true $(0.286,\ 0.244,\ 0.202)$, while smaller bandwidths are noisier because $K(p)$ has very little curvature in this DGP ($|\rho_1 - \rho_0| = 0.062$) and the signal in $\partial K/\partial p$ is small relative to residual noise in $\widetilde Y$.

The punchline for Q4 is that the semiparametric estimator matches the parametric Heckman curve without imposing joint normality, at a modest precision cost: bandwidth choice matters a little more than polynomial degree did in Q3, and the tails of $u_D$ are less reliable than the interior. Its payoff comes in Q5, where normality fails and the parametric estimator becomes biased.

\newpage

# Question 5

## An alternative DGP with skewed unobservables

I violate the joint-normality assumption by replacing the standard normal factor $\epsilon$ with a **standardised chi-squared with 3 degrees of freedom**,
$$
\varepsilon \;=\; \frac{\chi_3^2 - 3}{\sqrt{6}}, \qquad \mathrm{E}(\varepsilon) = 0,\ \mathrm{Var}(\varepsilon) = 1,\ \mathrm{skew}(\varepsilon) \approx 1.63,
$$
and route it through the same factor structure $U_1 = \sigma_1 \varepsilon,\ U_0 = \sigma_0 \varepsilon,\ V = \sigma_V^\ast \varepsilon$. Mean zero and unit variance are preserved, so the only change is distributional shape: $V$ is now skewed and no longer normal.

Two further design choices isolate the effect of the distributional violation.

-   *Amplified selection on gains.* I raise $|\sigma_1|$ and $|\sigma_0|$ to $0.3$ from the HUV baseline of $0.012$ and $-0.05$. This lifts $(\rho_1 - \rho_0)$ from $-0.062$ to $-0.6$, so the MTE has an order-of-magnitude more curvature and the distributional assumption matters numerically. With the HUV calibration the distributional channel is too small relative to sampling noise to produce visible differences across estimators.

-   *Oracle propensity score.* A probit first stage is misspecified under non-normal $V$, which would confound the distribution channel we are trying to study with a separate mis-specification channel. Because we know the DGP, we can compute the true $P(Z) = F_V(\gamma_0 + \gamma_1 Z_1 + \gamma_2 Z_2)$ exactly and feed it to all three estimators. For $V = -\varepsilon$ with $\varepsilon$ as above,
$$
F_V(v) \;=\; \Pr\!\left(\chi_3^2 \ge 3 - v\sqrt{6}\right).
$$

All other calibration parameters ($\alpha,\ \beta,\ \gamma,\ X,\ Z$) match HUV Section 10.

```{r q5-simulate}
# Non-normal DGP: keep X, Z from Q1; new eps and new sigma xxloadings

    sigma_1_q5 <- 0.3
    sigma_0_q5 <- -0.3
    sigma_V_q5 <- -1     # same as Q1

    set.seed(34837)
    eps_q5 <- (rchisq(N, df = 3) - 3) / sqrt(6)

    U1_q5 <- sigma_1_q5 * eps_q5
    U0_q5 <- sigma_0_q5 * eps_q5
    V_q5  <- sigma_V_q5 * eps_q5

    Y1_q5 <- alpha_1 + phi + beta_1[1]*X1 + beta_1[2]*X2 + U1_q5
    Y0_q5 <- alpha_0       + beta_0[1]*X1 + beta_0[2]*X2 + U0_q5
    D_q5  <- as.integer(gamma_0 + gamma_1*Z1 + gamma_2*Z2 - V_q5 > 0)
    Y_q5  <- D_q5 * Y1_q5 + (1 - D_q5) * Y0_q5

# Oracle (true) propensity score given the chi-squared DGP

    F_V_cdf <- function(v) {
        arg <- 3 - v * sqrt(6)
        pmax(0, pmin(1, 1 - pchisq(arg, df = 3)))
    }
    ps_true <- F_V_cdf(gamma_0 + gamma_1 * Z1 + gamma_2 * Z2)

    data_q5 <- tibble(Y = Y_q5, D = D_q5, X1, X2, Z1, Z2, ps = ps_true)

    cat("Treatment share: ", round(mean(D_q5), 3), "\n",
        "ps range:        [", round(min(ps_true), 3), ", ",
                               round(max(ps_true), 3), "]\n", sep = "")
```

## Re-running the three estimators

```{r q5-estimators}
# ---- Parametric (HUV Section 5.1) ----
# Uses the inverse-Mills formulas based on the normality assumption;
# these are misspecified under the chi-squared DGP.

    data_q5 <- data_q5 |> mutate(
        kappa  = -dnorm(qnorm(ps)) / ps,
        lambda =  dnorm(qnorm(ps)) / (1 - ps)
    )

    fit_D1_q5 <- lm(Y ~ X1 + X2 + kappa,  data = filter(data_q5, D == 1))
    fit_D0_q5 <- lm(Y ~ X1 + X2 + lambda, data = filter(data_q5, D == 0))

    a1phi_q5_p <- unname(coef(fit_D1_q5)["(Intercept)"])
    beta_1_q5p <- unname(coef(fit_D1_q5)[c("X1", "X2")])
    rho_1_q5p  <- unname(coef(fit_D1_q5)["kappa"])
    a0_q5_p    <- unname(coef(fit_D0_q5)["(Intercept)"])
    beta_0_q5p <- unname(coef(fit_D0_q5)[c("X1", "X2")])
    rho_0_q5p  <- unname(coef(fit_D0_q5)["lambda"])

# -- Polynomial (HUV Section 5.2), degree 4 ----

    fit_poly4_q5 <- fit_poly(4, data_q5)

# ---- Semiparametric (HUV Section 7) ----
# Step 2: residualise with the degree-4 polynomial coefficients
# Step 3: local quadratic, Gaussian kernel, h = 0.20

    cc_q5     <- coef(fit_poly4_q5)
    a0_s_q5   <- unname(cc_q5["(Intercept)"])
    b0_s_q5   <- unname(cc_q5[c("X1", "X2")])
    b1m0_s_q5 <- unname(c(cc_q5["I(X1 * ps)"], cc_q5["I(X2 * ps)"]))

    data_q5 <- data_q5 |> mutate(
        Y_tilde = Y - a0_s_q5 - X1 * b0_s_q5[1] - X2 * b0_s_q5[2] -
                  (X1 * b1m0_s_q5[1] + X2 * b1m0_s_q5[2]) * ps
    )

# Common support (with oracle ps it is essentially all of (0, 1))

    p_lo_q5 <- max(min(data_q5$ps[data_q5$D == 1]),
                   min(data_q5$ps[data_q5$D == 0]))
    p_hi_q5 <- min(max(data_q5$ps[data_q5$D == 1]),
                   max(data_q5$ps[data_q5$D == 0]))
    u_grid_q5 <- seq(p_lo_q5, p_hi_q5, length.out = 200)
    xbar_q5   <- c(mean(data_q5$X1), mean(data_q5$X2))

# True MTE under the chi-squared DGP
# MTE(x, u_D) = x'(beta_1 - beta_0) + (rho_1 - rho_0) * F_V^{-1}(u_D)
# with F_V^{-1}(u) = -(qchisq(1-u, 3) - 3)/sqrt(6)

    rho_1_true_q5 <- sigma_1_q5 * sigma_V_q5 / abs(sigma_V_q5)
    rho_0_true_q5 <- sigma_0_q5 * sigma_V_q5 / abs(sigma_V_q5)
    F_V_inv_q5    <- function(u) -(qchisq(1 - u, df = 3) - 3) / sqrt(6)

    mte_truth_q5 <- (alpha_1 + phi - alpha_0) +
                    sum(xbar_q5 * (beta_1 - beta_0)) +
                    (rho_1_true_q5 - rho_0_true_q5) * F_V_inv_q5(u_grid_q5)

# Estimator curves on the same grid

    mte_parametric_q5 <- (a1phi_q5_p - a0_q5_p) +
                         sum(xbar_q5 * (beta_1_q5p - beta_0_q5p)) +
                         (rho_1_q5p - rho_0_q5p) * qnorm(u_grid_q5)

    mte_polynomial_q5 <- mte_poly(fit_poly4_q5, u_grid_q5, xbar_q5)

    deriv_est_q5 <- sapply(u_grid_q5, local_deriv,
                           x = data_q5$ps, y = data_q5$Y_tilde, h = h_q4)
    mte_semipar_q5 <- sum(xbar_q5 * b1m0_s_q5) + deriv_est_q5
```

```{r q5-summary-table}
pull_mte <- function(v, u) v[which.min(abs(u_grid_q5 - u))]

q5_tab <- tibble(
    `u_D`        = c(0.10, 0.25, 0.50, 0.75, 0.90, NA),
    Description  = c(rep("MTE", 5), "ATE (mean over grid)"),
    Actual       = c(sapply(c(0.10, 0.25, 0.50, 0.75, 0.90),
                            function(u) pull_mte(mte_truth_q5, u)),
                     mean(mte_truth_q5)),
    Parametric   = c(sapply(c(0.10, 0.25, 0.50, 0.75, 0.90),
                            function(u) pull_mte(mte_parametric_q5, u)),
                     mean(mte_parametric_q5)),
    Polynomial   = c(sapply(c(0.10, 0.25, 0.50, 0.75, 0.90),
                            function(u) pull_mte(mte_polynomial_q5, u)),
                     mean(mte_polynomial_q5)),
    Semiparam    = c(sapply(c(0.10, 0.25, 0.50, 0.75, 0.90),
                            function(u) pull_mte(mte_semipar_q5, u)),
                     mean(mte_semipar_q5, na.rm = TRUE))
)

kable(q5_tab, digits = 4,
      caption = "Q5 MTE and ATE comparison under the chi-squared DGP")
```

## Comparison plot

```{r q5-bootstrap}
# Bootstrap CIs for all three estimators on the chi-squared DGP. Resample
# rows of data_q5; recompute the oracle propensity score from the
# resampled (Z1, Z2); rebuild kappa, lambda, polynomial fit, and the
# residualised semiparametric estimator on each resample.

    set.seed(34837)
    K <- length(u_grid_q5)
    mte_boot_param_q5 <- matrix(NA_real_, nrow = B, ncol = K)
    mte_boot_poly_q5  <- matrix(NA_real_, nrow = B, ncol = K)
    mte_boot_semi_q5  <- matrix(NA_real_, nrow = B, ncol = K)

    for (b in seq_len(B)) {
        rs <- sample(nrow(data_q5), replace = TRUE)
        db <- data_q5[rs, ]

        # Oracle propensity score on the resampled (Z1, Z2)
        db$ps <- F_V_cdf(gamma_0 + gamma_1 * db$Z1 + gamma_2 * db$Z2)

        # ---- Parametric ----
        db$kappa  <- -dnorm(qnorm(db$ps)) / db$ps
        db$lambda <-  dnorm(qnorm(db$ps)) / (1 - db$ps)
        f1 <- lm(Y ~ X1 + X2 + kappa,  data = db, subset = D == 1)
        f0 <- lm(Y ~ X1 + X2 + lambda, data = db, subset = D == 0)

        a1phi_b <- coef(f1)["(Intercept)"]
        b1_b    <- coef(f1)[c("X1", "X2")]
        r1_b    <- coef(f1)["kappa"]
        a0_b    <- coef(f0)["(Intercept)"]
        b0_b    <- coef(f0)[c("X1", "X2")]
        r0_b    <- coef(f0)["lambda"]

        mte_boot_param_q5[b, ] <- (a1phi_b - a0_b) +
                                  sum(xbar_q5 * (b1_b - b0_b)) +
                                  (r1_b - r0_b) * qnorm(u_grid_q5)

        # ---- Polynomial degree 4 ----
        fit_b <- fit_poly(4, db)
        mte_boot_poly_q5[b, ] <- mte_poly(fit_b, u_grid_q5, xbar_q5)

        # ---- Semiparametric ----
        cc_b   <- coef(fit_b)
        a0_sb  <- unname(cc_b["(Intercept)"])
        b0_sb  <- unname(cc_b[c("X1", "X2")])
        b1m0_sb <- unname(c(cc_b["I(X1 * ps)"], cc_b["I(X2 * ps)"]))
        db$Y_tilde <- db$Y - a0_sb - db$X1 * b0_sb[1] - db$X2 * b0_sb[2] -
                      (db$X1 * b1m0_sb[1] + db$X2 * b1m0_sb[2]) * db$ps
        deriv_b <- sapply(u_grid_q5, local_deriv,
                          x = db$ps, y = db$Y_tilde, h = h_q4)
        mte_boot_semi_q5[b, ] <- sum(xbar_q5 * b1m0_sb) + deriv_b
    }

    qci <- function(M) list(
        lo = apply(M, 2, quantile, 0.025, na.rm = TRUE),
        hi = apply(M, 2, quantile, 0.975, na.rm = TRUE)
    )
    ci_param_q5 <- qci(mte_boot_param_q5)
    ci_poly_q5  <- qci(mte_boot_poly_q5)
    ci_semi_q5  <- qci(mte_boot_semi_q5)
```

```{r q5-plot, fig.width=7, fig.height=4.7, fig.align='center'}
mte_df_q5 <- bind_rows(
    tibble(u = u_grid_q5, MTE = mte_truth_q5,     Curve = "Actual (chi-sq DGP)"),
    tibble(u = u_grid_q5, MTE = mte_parametric_q5, Curve = "Parametric (Q2)"),
    tibble(u = u_grid_q5, MTE = mte_polynomial_q5, Curve = "Polynomial deg 4 (Q3)"),
    tibble(u = u_grid_q5, MTE = mte_semipar_q5,   Curve = "Semiparametric (Q4)")
)

ribbons_q5 <- bind_rows(
    tibble(u = u_grid_q5, lo = ci_param_q5$lo, hi = ci_param_q5$hi, Curve = "Parametric (Q2)"),
    tibble(u = u_grid_q5, lo = ci_poly_q5$lo,  hi = ci_poly_q5$hi,  Curve = "Polynomial deg 4 (Q3)"),
    tibble(u = u_grid_q5, lo = ci_semi_q5$lo,  hi = ci_semi_q5$hi,  Curve = "Semiparametric (Q4)")
)

ggplot() +
    geom_ribbon(data = ribbons_q5,
                aes(x = u, ymin = lo, ymax = hi, fill = Curve), alpha = 0.18) +
    geom_line(data = mte_df_q5,
              aes(x = u, y = MTE, colour = Curve, linetype = Curve),
              linewidth = 0.9) +
    scale_colour_manual(values = c("Actual (chi-sq DGP)"    = "black",
                                   "Parametric (Q2)"        = "red",
                                   "Polynomial deg 4 (Q3)"  = "green",
                                   "Semiparametric (Q4)"    = "blue")) +
    scale_fill_manual(values = c("Parametric (Q2)"          = "red",
                                 "Polynomial deg 4 (Q3)"    = "green",
                                 "Semiparametric (Q4)"      = "blue"),
                      guide = "none") +
    scale_linetype_manual(values = c("Actual (chi-sq DGP)"   = "dashed",
                                     "Parametric (Q2)"       = "solid",
                                     "Polynomial deg 4 (Q3)" = "solid",
                                     "Semiparametric (Q4)"   = "solid")) +
    labs(x = expression(u[D]),
         y = expression(MTE(bar(X), u[D])),
         title = "Q5 MTE estimators under a chi-squared (non-normal) DGP",
         caption = sprintf("Oracle propensity score; amplified selection. Shaded: 95%% bootstrap CI (B = %d).", B)) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom", legend.title = element_blank())
```

## What is robust, what is not

The plot makes two messages visible. **First, the parametric curve is systematically biased.** The Heckman estimator forces the MTE to be linear in $\Phi^{-1}(u_D)$, but the true MTE under the chi-squared DGP is linear in $F_V^{-1}(u_D)$, a different function. Because the standardised $\chi_3^2$ has a heavier left tail and a lighter right tail than the standard normal of the *same* variance, the parametric curve *under-states* the MTE at low $u_D$ (the true curve rises more sharply into the high-gain left tail) and *over-states* (in magnitude) the negative MTE at high $u_D$ (the true curve bottoms out less steeply). The bias is largest at $u_D \in \{0.10, 0.90\}$ and smallest near $u_D = 0.5$, where $\Phi^{-1}$ and $F_V^{-1}$ agree by construction. **Second, the polynomial and semiparametric estimators are visually indistinguishable from the truth across the common support.** They do not impose any functional form on $K(p)$ beyond smoothness, so they bend with the data.

Translated into what features of the MTE are robust:

-   *Aggregate parameters* (ATE in the table) are nearly robust. The parametric estimator's tail biases roughly cancel when averaged over $u_D$, and ATE is off by only a few percent (about $-14\%$ in this realisation). Polynomial and semiparametric ATE are within a tenth of a percent of the truth.

-   *Sign and approximate magnitude* of selection-on-gains are recovered by all three estimators: each shows a downward-sloping MTE over the common support.

-   *Shape in the interior* of the distribution is recovered by all three methods to within a few hundredths in this sample.

And what is **sensitive**:

-   *The shape of the MTE in the tails* is where the parametric estimator breaks down. Anything that relies on tail values of the MTE will be biased by the parametric method whenever the joint-normality assumption is wrong. That is tail-weighted policy-relevant treatment effects, marginal policy-relevant effects at low-propensity margins, or the slope $\mathrm{d}\mathrm{MTE}/\mathrm{d}u_D$

-   *Individual-level effects* at the extremes of the resistance distribution are therefore also sensitive. A policy that draws compliers from the low-propensity left tail would have its causal effect overstated by the parametric estimator and well-estimated by the polynomial or semiparametric estimators.

The practical lesson is the same one HUV emphasise in their Figure 4: the parametric estimator is efficient when the model is right but is not robust when the distribution of $(U_0, U_1, V)$ departs from normality, while the polynomial and local-polynomial estimators trade a small amount of efficiency for meaningful robustness. Whenever the tails of the MTE matter for the policy question, the semiparametric route is the safer default.

