---
title: "Advanced Empirical Research Methods Hilary 2025/26"
author: 'Candidate ID: 1091806'
date: "2026-04-24"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
    toc_depth: 2
    number_sections: false
  html_document:
    toc: true
    toc_depth: '2'
    df_print: paged
subtitle: 'Assessment Exercise: Marginal Treatment Effects'
fontsize: 11pt
geometry: margin=1in
header-includes:
- \usepackage{amsmath}
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{newunicodechar}
- \newunicodechar{ρ}{$\rho$}
- \newunicodechar{π}{$\pi$}
- \newunicodechar{ξ}{$\xi$}
- \newunicodechar{ϕ}{$\phi$}
- \newunicodechar{φ}{$\varphi$}
- \newunicodechar{γ}{$\gamma$}
- \newunicodechar{σ}{$\sigma$}
- \newunicodechar{α}{$\alpha$}
- \newunicodechar{β}{$\beta$}
- \newunicodechar{Φ}{$\Phi$}
editor_options:
  markdown:
    wrap: 72
---

\newpage

# Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(knitr)
```

\newpage

# Question 1

The DGP follows Section 10 of Heckman, Urzua and Vytlacil (2006). It specifies a generalised Roy model with two potential outcomes:

$$Y_1 = \alpha_1 + \phi + X'\beta_1 + U_1, \quad Y_0 = \alpha_0 + X'\beta_0 + U_0$$

Treatment selection is as:

$$
D = \mathbf{1}\{\gamma_0 + \gamma_1 Z_1 + \gamma_2 Z_2 - V > 0\}
$$

The unobservables share a common factor structure: $U_1 = \sigma_1 \epsilon$, $U_0 = \sigma_0 \epsilon$, $V = \sigma_V^* \epsilon$, where $\epsilon \sim N(0,1)$. This ensures joint normality,

```{r q1-simulate}
# Seed and sample size

    set.seed(34837)
    N <- 5000

# Parameters

    alpha_0 <- 0.02
    alpha_1 <- 0.04
    phi     <- 0.2

    beta_0 <- c(0.5, 0.1)
    beta_1 <- c(0.8, 0.4)

    gamma_0 <- 0.2
    gamma_1 <- 0.3
    gamma_2 <- 0.1

    sigma_1  <- 0.012
    sigma_0  <- -0.05
    sigma_V  <- -1

# Draw covariates and instruments (all mutually independent)
# Variance 4 and 9: sd = √4=2; sd =√9=3

    X1 <- rnorm(N, mean = -2, sd = 2)
    X2 <- rnorm(N, mean =  2, sd = 2)
    Z1 <- rnorm(N, mean = -1, sd = 3)
    Z2 <- rnorm(N, mean =  1, sd = 3)

# Common factor and unobservables

    epsilon <- rnorm(N)
    U1 <- sigma_1 * epsilon
    U0 <- sigma_0 * epsilon
    V  <- sigma_V * epsilon

# Potential outcomes

    Y1 <- alpha_1 + phi + beta_1[1] * X1 + beta_1[2] * X2 + U1
    Y0 <- alpha_0 + beta_0[1] * X1 + beta_0[2] * X2 + U0

# Selection and observed outcome

    latent_index <- gamma_0 + gamma_1 * Z1 + gamma_2 * Z2 - V
    D <- as.integer(latent_index > 0)
    Y <- D * Y1 + (1 - D) * Y0

# Collect into data frame

    data <- tibble(Y, D, Y1, Y0, X1, X2, Z1, Z2, epsilon, U1, U0, V)

    cat("Treatment share:", round(mean(D), 3), "\n")
    cat("Sample size:", N, "\n")
```

## Propensity Score Estimation

The propensity score is defined as $P(z) = \Pr(D = 1 \mid Z = z)$. Given the probit structure implied by $V = \sigma_V^* \epsilon$ with $\epsilon \sim N(0,1)$, a natural estimator is:

$$\hat{P}(z) = \Phi(z'\hat{\gamma})$$

estimated via probit regression of $D$ on $(Z_1, Z_2)$.

```{r q1-propensity}
# Estimate propensity score via probit
ps_model <- glm(D ~ Z1 + Z2, family = binomial(link = "probit"), data = data)

# Store fitted propensity scores
data <- data |> mutate(ps = predict(ps_model, type = "response"))

# Summary of propensity score distribution
summary(data$ps)
```

## Checking Full Support

Full support requires that the estimated propensity score spans the full unit interval $[0,1]$ and, crucially, that both treated ($D=1$) and untreated ($D=0$) individuals are observed at every value of $P(Z)$. I check this in two ways: visually and via a binned overlap test.

```{r q1-density-plot, fig.width=7, fig.height=4}
# Visual check: overlaid densities by treatment status
ggplot(data, aes(x = ps, fill = factor(D))) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 60, alpha = 0.45, position = "identity") +
  labs(x = "Estimated propensity score",
       y = "Density",
       fill = "Treatment") +
  scale_fill_manual(values = c("0" = "steelblue", "1" = "tomato"),
                    labels = c("D = 0", "D = 1")) +
  theme_minimal(base_size = 13)
```

```{r q1-bin-check}
# Binned overlap check at 1-percentile resolution
breaks <- seq(0, 1, by = 0.01)

bin_overlap <- data |>
  mutate(bin = cut(ps, breaks = breaks, include.lowest = TRUE)) |>
  count(bin, D) |>
  pivot_wider(names_from = D, values_from = n, values_fill = 0) |>
  rename(untreated = `0`, treated = `1`) |>
  mutate(has_overlap = untreated > 0 & treated > 0)

# Show bins WITHOUT overlap
no_overlap <- bin_overlap |> filter(!has_overlap)

kable(no_overlap,
      caption = "Propensity score bins lacking common support",
      col.names = c("Bin", "Untreated", "Treated", "Overlap"))
```

```{r q1-support-range}
# Continuous support range
support_bins <- bin_overlap |> filter(has_overlap)
cat("Common support region: [",
    min(breaks[breaks >= 0][which(bin_overlap$has_overlap)[1]]),
    ",",
    max(breaks[breaks <= 1][tail(which(bin_overlap$has_overlap), 1) + 1]),
    "]\n")

cat("Number of bins with overlap:",
    sum(bin_overlap$has_overlap), "out of", nrow(bin_overlap), "\n")
```

## Discussion

**Full support** means that for every value $u_D \in [0,1]$, there exist individuals from both treatment groups in a neighbourhood of $P(Z) = u_D$. The results above show that support fails only at the extreme tails — the lowest and highest few percentiles of the propensity score distribution contain only untreated or only treated individuals, respectively.

**Full support matters** because the MTE is identified as the derivative of a conditional expectation with respect to the propensity score:

$$MTE(x, u_D) = \frac{\partial\, E[Y \mid X = x,\, P(Z) = p]}{\partial p}\bigg|_{p = u_D}$$

Computing this derivative at any point $u_D$ requires observing outcomes from individuals with propensity scores in a neighbourhood of $u_D$. If only one treatment group is present at some value of $p$, the conditional expectation $E[Y \mid X, P(Z) = p]$ cannot trace out how outcomes change with $p$ at that margin, and the derivative is not identified. The MTE can therefore only be estimated over the region where common support holds. In this simulation, that region covers approximately $(0.02, 0.97)$, which is sufficient for practical purposes.

\newpage

# Question 2

## Derivation of the MTE formula

The equation at the bottom of HUV page 4,
$$
\mathrm{MTE}(X = x, U_D = u_D) \;=\; x'(\beta_1 - \beta_0) \;+\; (\rho_1 - \rho_0)\,\Phi^{-1}(u_D), \qquad (\star)
$$
follows from the latent-index model together with joint normality of $(U_0, U_1, V)$. For brevity the intercepts $\alpha_0$ and $\alpha_1 + \phi$ are folded into the $x'\beta_d$ terms by extending $x$ to include a constant.

**Step 1: normalise the latent index.** Since $V$ is continuously distributed, the choice rule $D = \mathbf{1}\{Z\gamma > V\}$ is equivalent to
$$
D = \mathbf{1}\bigl\{\,F_V(Z\gamma) > F_V(V)\,\bigr\} \;=\; \mathbf{1}\{P(Z) > U_D\},
$$
where $U_D \equiv F_V(V) = \Phi(V/\sigma_V)$ is standard uniform on $[0, 1]$ and $P(Z) \equiv \Phi(Z\gamma/\sigma_V)$. Conditioning on $U_D = u_D$ is therefore equivalent to conditioning on $V/\sigma_V = \Phi^{-1}(u_D)$.

**Step 2: marginal treatment effect.** Substituting the outcome equations,
$$
\mathrm{MTE}(x, u_D) = E[\,Y_1 - Y_0 \mid X = x,\, U_D = u_D\,]
= x'(\beta_1 - \beta_0) + E[\,U_1 - U_0 \mid V/\sigma_V = \Phi^{-1}(u_D)\,].
$$

**Step 3: conditional means under joint normality.** Because $(U_j, V/\sigma_V)$ is bivariate normal with $\mathrm{Var}(V/\sigma_V) = 1$ and $\mathrm{Cov}(U_j, V/\sigma_V) = \sigma_{V,j}/\sigma_V \equiv \rho_j$, the conditional expectation is linear,
$$
E[\,U_j \mid V/\sigma_V = v^\ast\,] = \rho_j\, v^\ast, \qquad j = 0, 1.
$$

**Step 4: plug in.** Taking $v^\ast = \Phi^{-1}(u_D)$ and differencing,
$$
E[\,U_1 - U_0 \mid U_D = u_D\,] = (\rho_1 - \rho_0)\,\Phi^{-1}(u_D),
$$
so that $(\star)$ follows directly.

The formula makes two features of the Section 10 DGP transparent. First, the MTE is linear in $\Phi^{-1}(u_D)$ with slope $(\rho_1 - \rho_0)$, which captures the difference in selection on unobservable gains across potential outcomes. Second, $X$ enters as a parallel shift $x'(\beta_1 - \beta_0)$, the conditional ATE. At the calibration, $\rho_1 = \sigma_1 \sigma_V^\ast / |\sigma_V^\ast| = -0.012$ and $\rho_0 = \sigma_0 \sigma_V^\ast / |\sigma_V^\ast| = 0.05$, so $(\rho_1 - \rho_0) = -0.062 < 0$: selection on gains is negative, meaning individuals with high cost of treatment (large $\Phi^{-1}(u_D)$) gain less from it.

\newpage

## Estimate the MTE (HUV Section 5.1)

Under joint normality HUV (p. 4) show that
$$
E[Y \mid X, D = 1, P(Z) = p] = (\alpha_1 + \phi) + X\beta_1 \;+\; \rho_1 \underbrace{\Bigl(-\tfrac{\varphi(\Phi^{-1}(p))}{p}\Bigr)}_{\kappa(p)},
$$
$$
E[Y \mid X, D = 0, P(Z) = p] = \alpha_0 + X\beta_0 \;+\; \rho_0 \underbrace{\Bigl(\tfrac{\varphi(\Phi^{-1}(p))}{1-p}\Bigr)}_{\lambda(p)},
$$
where $\varphi$ and $\Phi$ are the standard-normal density and cdf. The two parameters $\rho_j = \sigma_{V,j}/\sigma_V$ are therefore identified as the coefficients on the (known) inverse-Mills ratios $\kappa$ and $\lambda$. I implement HUV's Step 1 by running two Heckman-style control-function regressions,

-   regress $Y$ on $(X_1, X_2, \kappa(\widehat{P}))$ in the $D = 1$ subsample, recovering $(\widehat{\alpha_1 + \phi},\ \widehat\beta_1,\ \widehat\rho_1)$;

-   regress $Y$ on $(X_1, X_2, \lambda(\widehat{P}))$ in the $D = 0$ subsample, recovering $(\widehat\alpha_0,\ \widehat\beta_0,\ \widehat\rho_0)$.

Step 2 then plugs these coefficients into $(\star)$.

```{r q2-parametric}
# Build the two inverse-Mills control functions from the estimated
# propensity score

    data <- data |> mutate(
        kappa  = -dnorm(qnorm(ps)) / ps,           # for the D = 1 regression
        lambda =  dnorm(qnorm(ps)) / (1 - ps)      # for the D = 0 regression
    )

# Heckman-style selection-corrected regressions

    fit_D1 <- lm(Y ~ X1 + X2 + kappa,  data = filter(data, D == 1))
    fit_D0 <- lm(Y ~ X1 + X2 + lambda, data = filter(data, D == 0))

# Extract coefficients

    a1phi_hat  <- unname(coef(fit_D1)["(Intercept)"])
    beta_1_hat <- unname(coef(fit_D1)[c("X1", "X2")])
    rho_1_hat  <- unname(coef(fit_D1)["kappa"])

    a0_hat     <- unname(coef(fit_D0)["(Intercept)"])
    beta_0_hat <- unname(coef(fit_D0)[c("X1", "X2")])
    rho_0_hat  <- unname(coef(fit_D0)["lambda"])
```

```{r q2-table}
rho_1_true <- sigma_1 * sigma_V / abs(sigma_V)      # -0.012
rho_0_true <- sigma_0 * sigma_V / abs(sigma_V)      #  0.050

q2_tab <- tibble(
  Parameter = c("alpha_1 + phi", "alpha_0",
                "beta_{11}", "beta_{21}",
                "beta_{10}", "beta_{20}",
                "rho_1", "rho_0", "rho_1 - rho_0"),
  True      = c(alpha_1 + phi, alpha_0,
                beta_1[1], beta_1[2],
                beta_0[1], beta_0[2],
                rho_1_true, rho_0_true,
                rho_1_true - rho_0_true),
  Estimate  = c(a1phi_hat, a0_hat,
                beta_1_hat[1], beta_1_hat[2],
                beta_0_hat[1], beta_0_hat[2],
                rho_1_hat, rho_0_hat,
                rho_1_hat - rho_0_hat)
)

kable(q2_tab, digits = 4,
      caption = "Q2(b) parametric estimates versus true values")
```

The parametric estimator recovers the $\beta$ coefficients and the slope $\widehat\rho_1 - \widehat\rho_0$ very accurately. The intercepts $\alpha_0$ and $\alpha_1 + \phi$ are less tightly estimated because they are identified off the mean of the inverse-Mills term, but their *difference*, which is what enters the MTE at $E[X]$, comes out close to the population value of $0.22$. The sign pattern $\widehat\rho_1 < 0 < \widehat\rho_0$ confirms the negative selection on gains predicted by the calibration.

\newpage

## Graphical illustration 

Evaluating $(\star)$ at the sample mean of $X$ gives
$$
\widehat{\mathrm{MTE}}(\bar X, u_D) = \widehat{(\alpha_1 + \phi)} - \widehat\alpha_0 + \bar X'(\widehat\beta_1 - \widehat\beta_0) + (\widehat\rho_1 - \widehat\rho_0)\,\Phi^{-1}(u_D).
$$
I plot this against $u_D$ over the common support identified in Q1, alongside the population MTE implied by the calibration. HUV's Figure 4 uses $v = \Phi^{-1}(u_D)$ on the horizontal axis; plotting against $u_D$ is equivalent up to a monotone transformation and is more readable.

```{r q2-plot, fig.width=6.5, fig.height=4.2, fig.align='center'}
# Common support of the propensity score (matches the overlap region
# identified in Q1's binned check)

    p_lo <- max(min(data$ps[data$D == 1]), min(data$ps[data$D == 0]))
    p_hi <- min(max(data$ps[data$D == 1]), max(data$ps[data$D == 0]))
    u_grid <- seq(p_lo, p_hi, length.out = 200)

# Sample mean of X

    xbar <- c(mean(data$X1), mean(data$X2))

# Estimated and actual MTE

    mte_hat <- (a1phi_hat - a0_hat) +
               sum(xbar * (beta_1_hat - beta_0_hat)) +
               (rho_1_hat - rho_0_hat) * qnorm(u_grid)

    mte_true <- (alpha_1 + phi - alpha_0) +
                sum(xbar * (beta_1 - beta_0)) +
                (rho_1_true - rho_0_true) * qnorm(u_grid)

# Plot

    mte_df <- bind_rows(
        tibble(u = u_grid, MTE = mte_hat,  Curve = "Parametric estimate"),
        tibble(u = u_grid, MTE = mte_true, Curve = "Actual")
    )

    ggplot(mte_df, aes(x = u, y = MTE, colour = Curve, linetype = Curve)) +
        geom_line(linewidth = 0.9) +
        scale_colour_manual(values = c("Parametric estimate" = "firebrick",
                                       "Actual"              = "black")) +
        scale_linetype_manual(values = c("Parametric estimate" = "solid",
                                         "Actual"              = "dashed")) +
        labs(x = expression(u[D]),
             y = expression(MTE(bar(X), u[D])),
             title = "Q2(c) Parametric MTE (in the spirit of HUV Figure 4)",
             caption = "Evaluated at the sample mean of X over the common support of P(Z).") +
        theme_minimal(base_size = 12) +
        theme(legend.position = "bottom", legend.title = element_blank())
```

The estimated curve tracks the population MTE tightly across the common support. Consistent with Q2(a), the curve is downward-sloping and nearly linear in $\Phi^{-1}(u_D)$: individuals with low resistance to treatment ($u_D$ near 0) gain the most, and those with high resistance ($u_D$ near 1) gain the least. Because $|\rho_1 - \rho_0| = 0.062$ is small relative to the intercept of $0.22$, treatment-effect heterogeneity across the complier margin is modest, so aggregate parameters (ATE, ATT, ATUT, LATE) will all lie in a narrow band around $0.22$ in subsequent questions.
