---
title: "AERM Exam Outputs and Results"
output:
  pdf_document: default
  html_document: default
date: "2026-03-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

Use Julia code

# Exercise 1: Solve Standard Consumption-Saving Problem for Different Values of $\rho$

## Question 1

Income follows a two-state Markov chain with symmetric persistence $\rho_L = \rho_H = \rho$ and transition matrix:

$$P = \begin{pmatrix} \rho & 1-\rho \\ 1-\rho & \rho \end{pmatrix}$$

Starting from $P(y = y_L) = 0.2$, the figure below shows the share of low-income households over a 100-year horizon for each value of $\rho$.

```{r, echo=FALSE, fig.align='center', out.width='85%'}
knitr::include_graphics("/Users/caisuyan/Documents/GitHub/AERM/Advanced ERM Problem Set/MT25 Assignment 1/exercise 1/results/fig_ex1_q1.pdf")
```

*Explanation*:

For all $\rho \in (0,1)$, the share of low-income households converges to 0.5, which is the unique stationary distribution of this symmetric Markov chain, but higher persistence slows the rate of convergence.

The two edge cases behave differently: $\rho = 0$ causes full state-switching each period (so the share oscillates before settling), while $\rho = 1$ makes both states absorbing, so the initial distribution $P(y_L) = 0.2$ is permanent.

## Question 2

| $\rho$ | Var($\Delta \log y$) |
|--------|---------------------|
| 0.00   | 0.48045             |
| 0.05   | 0.45643             |
| 0.20   | 0.38436             |
| 0.50   | 0.24023             |
| 0.70   | 0.14414             |
| 0.95   | 0.02402             |
| 1.00   | 0.00000             |

Labour income volatility, measured as $\text{Var}(\Delta \log y)$, is maximised at $\rho = 0$ since households switch between income states every period, making changes in income most frequent. However, $\rho = 0$ does not generate the most risk: when income alternates deterministically, households can anticipate future income and smooth consumption via bond savings, whereas for intermediate or high values of $\rho$, income shocks are harder to insure against because the current state provides less (or too much persistent) information about future income.

## Question 3

```{r, echo=FALSE, fig.align='center', out.width='85%'}
knitr::include_graphics("/Users/caisuyan/Documents/GitHub/AERM/Advanced ERM Problem Set/MT25 Assignment 1/exercise 1/results/fig_ex1_q3.pdf")
```

The simulated share of high-income types closely tracks the continuum prediction but does NOT match it exactly, because a finite sample of 1000 households is subject to sampling variation whereas the continuum result holds by the law of large numbers over a unit mass of agents.

## Question 4

Using the simulated panel data, I regress current income on lagged income via OLS and fixed effects (implemented by demeaning), using only linear algebra as instructed.

| Model       | $\hat{\beta}$ |
|-------------|---------------|
| OLS         | 0.8983        |
| FE          | 0.5690        |
| True $\rho$ | 0.95          |

OLS is closer to the true $\rho = 0.95$ because the data-generating process contains no individual fixed effects — all households face the same Markov chain — so OLS is consistent. The FE estimator suffers from Nickell bias: with only $T = 8$ periods, demeaning mechanically creates a negative correlation between the demeaned lagged variable and the demeaned error, biasing $\hat{\beta}_{FE}$ substantially downward. This bias is systematic and would shrink only as $T \to \infty$.
