---
title: "Advanced Empirical Research Methods"
subtitle: "MT25 Assignment 1"
author: "Candidate Number: XXXXXXX"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
    toc_depth: 2
    number_sections: false
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{amsmath}
  - \usepackage{newunicodechar}
  - \newunicodechar{ρ}{$\rho$}
  - \newunicodechar{π}{$\pi$}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

# Exercise 1:

## Question 1

Income follows a two-state Markov chain with symmetric persistence $\rho_L = \rho_H = \rho$ and transition matrix:

$$P = \begin{pmatrix} \rho & 1-\rho \\ 1-\rho & \rho \end{pmatrix}$$

Starting from $P(y = y_L) = 0.2$, the figure below shows the share of low-income households over a 100-year horizon for each value of $\rho$.

```{r, echo=FALSE, fig.align='center', out.width='85%'}
knitr::include_graphics("/Users/caisuyan/Documents/GitHub/AERM/Advanced ERM Problem Set/MT25 Assignment 1/exercise 1/results/fig_ex1_q1.pdf")
```

*Explanation*:

For all $\rho \in (0,1)$, the share of low-income households converges to 0.5, which is the unique stationary distribution of this symmetric Markov chain, but higher persistence slows the rate of convergence.

The two edge cases behave differently: $\rho = 0$ causes full state-switching each period (so the share oscillates before settling), while $\rho = 1$ makes both states absorbing, so the initial distribution $P(y_L) = 0.2$ is permanent.

\newpage

## Question 2

| $\rho$ | Var($\Delta \log y$) |
|--------|---------------------|
| 0.00   | 0.48045             |
| 0.05   | 0.45643             |
| 0.20   | 0.38436             |
| 0.50   | 0.24023             |
| 0.70   | 0.14414             |
| 0.95   | 0.02402             |
| 1.00   | 0.00000             |

Labour income volatility, measured as $\text{Var}(\Delta \log y)$, is maximised at $\rho = 0$ since households switch between income states every period, making changes in income most frequent. However, $\rho = 0$ does not generate the most risk: when income alternates deterministically, households can anticipate future income and smooth consumption via bond savings, whereas for intermediate or high values of $\rho$, income shocks are harder to insure against because the current state provides less (or too much persistent) information about future income.

## Question 3

```{r, echo=FALSE, fig.align='center', out.width='85%'}
knitr::include_graphics("/Users/caisuyan/Documents/GitHub/AERM/Advanced ERM Problem Set/MT25 Assignment 1/exercise 1/results/fig_ex1_q3.pdf")
```

The simulated share of high-income types closely tracks the continuum prediction but does NOT match it exactly, because a finite sample of 1000 households is subject to sampling variation whereas the continuum result holds by the law of large numbers over a unit mass of agents.

## Question 4

Using the simulated panel data from Q3, I regress current income on lagged income via pooled OLS and fixed effects (implemented via within-transformation demeaning), using only linear algebra as instructed.

$$y_{i,t} = \alpha_{OLS} + \beta_{OLS} y_{i,t-1} + u_{i,t}$$

$$\tilde{y}_{i,t} = \beta_{FE} \tilde{y}_{i,t-1} + \tilde{u}_{i,t}, \quad \tilde{x}_{i,t} = x_{i,t} - \bar{x}_i$$

| Coefficient | Estimate | 
|-------------|----------|
| $\alpha_{OLS}$ | 0.1535 |
| $\beta_{FE}$   | 0.5690 |

The true population regression slope is $\beta = 2\rho - 1 = 0.9$, since $E[y_{t+1}|y_t]$ is linear in $y_t$ with slope $2\rho - 1$ when $y \in \{1,2\}$. OLS is close to this true value because the DGP contains no individual heterogeneity, so $E[y_{i,t-1} u_{i,t}] = 0$ and OLS is consistent. The FE estimator suffers from Nickell bias: demeaning over a short panel ($T=8$) induces a mechanical negative correlation between $\tilde{y}_{i,t-1}$ and $\tilde{u}_{i,t}$, biasing $\hat{\beta}_{FE}$ substantially downward. As the Monte Carlo simulations below confirm, this downward bias increases systematically with $\rho$.

```{r, echo=FALSE, fig.align='center', out.width='40%'}
knitr::include_graphics("../results/fig_ex1_q4_rho_0_0.pdf")
knitr::include_graphics("../results/fig_ex1_q4_rho_0_05.pdf")
knitr::include_graphics("../results/fig_ex1_q4_rho_0_2.pdf")
knitr::include_graphics("../results/fig_ex1_q4_rho_0_5.pdf")
knitr::include_graphics("../results/fig_ex1_q4_rho_0_7.pdf")
knitr::include_graphics("../results/fig_ex1_q4_rho_0_95.pdf")

```

\newpage

# Exercise 2:

## Question 1

| $\rho$ | Asset Supply |
|--------|-------------|
| 0.00   | -0.275510   |
| 0.05   | -0.258511   |
| 0.20   | -0.185725   |
| 0.50   | 0.027987    |
| 0.70   | 0.132477    |
| 0.95   | -0.443872   |
| 1.00   | -0.500000   |

For $\rho = 1$ the transition matrix is the identity, making both income states absorbing. There is no unique stationary distribution, so the iterative procedure fails to converge. However, since $\beta(1+r) = 0.85 \times 1.1 = 0.935 < 1$, households are impatient and face a declining consumption path. Without income risk, there is no precautionary savings motive, so both types eventually borrow to the constraint, giving $B = -0.5$.


## Question 2a

Table of counterfactual aggregate asset supply:

| | $B(\rho=0, c=c_{.5})$ | $B(\rho=.95, c=c_{.5})$ |
|---|---|---|
| Asset Supply | -0.11499999999999998 | 0.5655346767345566 |

## Question 2b

Table of decompositions:

| $\rho$ | $T_{total}$ | $T_1$ (%) | $T_2$ (%) |
|--------|------------|-----------|-----------|
| 0.00   | -0.3035   | 52.9  | 47.1  |
| 0.95   | -0.4719   | 213.9  | -113.9  |


## Question 2c

Although the stationary *marginal* distribution over income is always 50/50 regardless of $\rho$, changing $\rho$ alters the *joint* dynamics of how households move between income states, which affects how long they spend in consecutive high- or low-income spells and thus where they sit on the asset grid. With higher persistence, high-income households accumulate more assets over longer spells before switching, while low-income households remain borrowing-constrained for longer — shifting the stationary distribution over $(b, y)$ and hence aggregate asset supply, even with the same consumption policy.



\newpage


# Appendix

## Part 1 Exercise 1, Julia code

```julia
# Exam MT 2025 Assignment 1
# Part 1 Exercise 1 

# Load packages
using Pkg
using Plots
using Random
using Statistics
using LinearAlgebra

# ==============================================================================
# Question 1
# ==============================================================================

# === Set-up ===
# Income states

    yL = 1.0
    
    yH = 2.0

# Values of persistence to consider

    ρ_vals = [0.0, 0.05, 0.2, 0.5, 0.7, 0.95, 1.0]

# Time horizon
    T = 100

# Initial distribution: P(y = yL) = 0.2
    π0 = [0.2, 0.8]  # [P(yL), P(yH)]using Pkg

# Transition matrix for symmetric persistence ρ:
#       yL      yH
# yL [  ρ    1-ρ  ]
# yH [ 1-ρ    ρ   ]
    function transition_matrix(ρ)
        return [ρ (1-ρ); (1-ρ) ρ]
    end

# Evolve distribution forward: π(t+1) = πt' * P
# (row vector times transition matrix)
    function evolve_distribution(π0, P, T)
        shares_low = zeros(T + 1)
        shares_low[1] = π0[1]
        πt = copy(π0)
        for t in 1:T
            πt = πt' * P |> vec
            shares_low[t + 1] = πt[1]
        end
        return shares_low
    end

# Compute and plot
    Pkg.add("Plots")
    plt = plot(
        xlabel = "Year",
        ylabel = "Share of low-income households",
        title = "Evolution of low-income share by persistence ρ",
        legend = :right,
        size = (800, 500),
        linewidth = 2
    )

    for ρ in ρ_vals
        P = transition_matrix(ρ)
        shares = evolve_distribution(π0, P, T)
        plot!(plt, 0:T, shares, label = "ρ = $ρ")
    end

# Add reference line at 0.5

    hline!(plt, [0.5], color = :black, linestyle = :dash, label = "0.5", alpha = 0.5)

# ==============================================================================
# Question 2
# ==============================================================================

# At the stationary distribution (50/50), Pr(switch) = 1-ρ for each type
# Δlog(y) ∈ {0, log(2), -log(2)}

    Δ = log(2)

    println("ρ \t\t Var(Δlog y)")

    println("—"^30)

    for ρ in ρ_vals
        # Stationary shares
            p_L = (ρ == 1.0) ? 0.5 : 0.5  # symmetric chain

            p_H = 1 - p_L
        
        # Joint probabilities of transitions

            p_LL = p_L * ρ

            p_LH = p_L * (1 - ρ)
            
            p_HL = p_H * (1 - ρ)
            
            p_HH = p_H * ρ
            
        # E[Δlog y] and E[(Δlog y)²]

            mean_Δ = p_LH * Δ + p_HL * (-Δ)

            mean_Δ_sq = p_LH * Δ^2 + p_HL * Δ^2
            
            var_Δ = mean_Δ_sq - mean_Δ^2

        println("ρ = $ρ \t\t $(round(var_Δ, digits=5))")
    end


# ==============================================================================
# Question 3
# ==============================================================================

# Set seed

    Random.seed!(1776)

# Generate random numbers for 1000 households and 8 years

    rand_mat = rand(1000, 8)

# Set parameters

    N = 1000
    T_sim = 8
    ρ_sim = 0.95
    

# Transition matrix for ρ = 0.95

    P_sim = transition_matrix(ρ_sim)

# Allocate state matrix: 1 = low income, 2 = high income
# Use column 1 of rand_mat to assign initial condition:
# ~20% start in low state, ~80% start in high state
    states = fill(2, N, T_sim)
    
    states[rand_mat[:, 1] .< 0.2, 1] .= 1

# Simulate forward using columns 2:T_sim of rand_mat
# For each household, compare rand draw to cumulative probability of staying in current state
# If rand < P[s,1] → go to state 1 (low), otherwise → state 2 (high)

    for t in 2:T_sim
        for i in 1:N

            # Get current state
            s = states[i, t-1]

            # Update state based on random draw and transition probabilities
            states[i, t] = rand_mat[i, t] < P_sim[s, 1] ? 1 : 2
        end
    end

# Compute share of high types in simulation at each period
   
    share_high_sim = vec(mean(states .== 2, dims=1))

# Compute share of high types from the continuum (theoretical) exercise
# Reuse the evolve_distribution function from Q1 with ρ = 0.95
    
    P95 = transition_matrix(0.95)
    π_cont = [0.2, 0.8]  # initial distribution [P(low), P(high)]
    shares_low_cont = evolve_distribution(π_cont, P95, T_sim - 1)
    share_high_cont = 1.0 .- shares_low_cont  # convert to high-type share

# Plot comparison

    plt3 = plot(1:T_sim, share_high_sim,

        marker=:circle, linewidth=2, label="1000 households",
        xlabel="Year", ylabel="Share of high-income households",
        title="Simulated vs. Continuum: Share of High Types (ρ=0.95)",
        size=(800, 500))

    plot!(plt3, 1:T_sim, share_high_cont,
        marker=:circle, linestyle=:dash, linewidth=2, label="Continuum")
        

# ==============================================================================
# Question 4
# ==============================================================================

# Convert states to income levels (1 = yL = 1, 2 = yH = 2)

    Y_panel = float.(states)
    Y_panel[states .== 1] .= yL
    Y_panel[states .== 2] .= yH

# Dependent variable y_{i,t} for t=2,...,8 and lag y_{i,t-1} for t=1,...,7
    
    Y_curr = Y_panel[:, 2:end]
    Y_lag  = Y_panel[:, 1:end-1]

# Stack into vectors

    y_vec     = vec(Y_curr)
    y_lag_vec = vec(Y_lag)

# --- Model 1: Pooled OLS ---

    X_ols = hcat(ones(length(y_lag_vec)), y_lag_vec)
    β_ols = X_ols \ y_vec
    
    println("OLS:  α = $(round(β_ols[1], digits=4)),  β = $(round(β_ols[2], digits=4))")

# --- Model 2: Fixed Effects (within transformation) ---

    y_curr_mean = mean(Y_curr, dims=2)
    y_lag_mean  = mean(Y_lag, dims=2)

    Y_curr_dm = Y_curr .- y_curr_mean
    Y_lag_dm  = Y_lag  .- y_lag_mean

    y_dm     = vec(Y_curr_dm)
    y_lag_dm = vec(Y_lag_dm)

    β_fe = dot(y_lag_dm, y_dm) / dot(y_lag_dm, y_lag_dm)

    println("FE:   β = $(round(β_fe, digits=4))")
    println("True β = 2ρ - 1 = $(2*0.95 - 1)")

# ===== Monte Carlo simulation: sampling distribution of β̂ across ρ values =====

    function monte_carlo_beta(ρ; N=1000, T=8, reps=5000)

        β_ols_vec = zeros(reps)
        β_fe_vec  = zeros(reps)
        
        P_mc = transition_matrix(ρ)
        
        for r in 1:reps
            # Draw random matrix
            rm = rand(N, T)
            
            # Assign initial states using column 1
            s = fill(2, N, T)
            s[rm[:, 1] .< 0.2, 1] .= 1
            
            # Simulate forward
            for t in 2:T
                for i in 1:N
                    s[i, t] = rm[i, t] < P_mc[s[i, t-1], 1] ? 1 : 2
                end
            end
            
            # Convert to income
            y = float.(s)
            y[s .== 1] .= yL
            y[s .== 2] .= yH
            
            # Current and lag
            Yc = y[:, 2:end]
            Yl = y[:, 1:end-1]
            
            # OLS
            yc_v = vec(Yc)
            yl_v = vec(Yl)
            X = hcat(ones(length(yl_v)), yl_v)
            β_ols_vec[r] = (X \ yc_v)[2]
            
            # FE (demeaning)
            Yc_dm = Yc .- mean(Yc, dims=2)
            Yl_dm = Yl .- mean(Yl, dims=2)
            yc_dm = vec(Yc_dm)
            yl_dm = vec(Yl_dm)
            denom = dot(yl_dm, yl_dm)
            β_fe_vec[r] = denom ≈ 0.0 ? NaN : dot(yl_dm, yc_dm) / denom
        end
        
        return β_ols_vec, β_fe_vec
    end

# ===== Plotting parameters =====

# Run Monte Carlo for each ρ and plot histograms
for ρ in ρ_vals
    β_ols_mc, β_fe_mc = monte_carlo_beta(ρ)
    β_fe_mc = β_fe_mc[isfinite.(β_fe_mc)]
    β_true = 2ρ - 1

    # Combine all finite estimates to set axis range
    all_β = vcat(β_ols_mc, β_fe_mc)
    bmin, bmax = extrema(all_β)
    margin = max((bmax - bmin) * 0.1, 0.01)
    xl = (bmin - margin, bmax + margin)
    edges = range(xl[1], xl[2], length=60)

    plt_mc = histogram(β_ols_mc, bins=edges, alpha=0.6, normalize=true, label="OLS",
        xlabel="β̂", ylabel="Density",
        title="Sampling distribution of β̂ (ρ = $ρ)",
        size=(700, 450), xlims=xl)

    if length(β_fe_mc) > 0
        histogram!(plt_mc, β_fe_mc, bins=edges, alpha=0.6, normalize=true, label="FE")
    end

    vline!(plt_mc, [β_true], color=:red, linestyle=:dash, linewidth=2,
        label="True β = $(round(β_true, digits=2))")

    ρ_str = replace(string(ρ), "." => "_")
    savefig(plt_mc, joinpath(RESULTS_DIR, "fig_ex1_q4_ρ_$(ρ_str).pdf"))
    display(plt_mc)
end
```







\newpage
## Part 1 Exercise 2, Julia code

## - Exercise 2 Main

```julia
# Exam MT 2025 Assignment 1
# Part 1 Exercise 2

# === Set up ===

    # Load packages

        using LinearAlgebra
        using Statistics
        using Plots
        using Printf

    # Set up paths

        const RESULTS_DIR = joinpath(@__DIR__, "..", "results")

    # Load code files for functions

        include("ex2_parameters.jl")
        include("ex2_vfi.jl")
        include("ex2_kfe.jl")

    # Set up parameters and ρ values    

        p = Params()
        ρ_vals = [0.0, 0.05, 0.2, 0.5, 0.7, 0.95, 1.0]

# ============================================================
# Question 1
# ============================================================

asset_supply = Dict{Float64, Float64}()
solutions = Dict{Float64, Any}()

for ρ in ρ_vals
    println("\n=== Solving for ρ = $ρ ===")
    sol = solve_household(p, ρ)
    solutions[ρ] = sol

    if ρ == 1.0
        # ρ=1: absorbing states, no unique stationary dist
        # Under 50/50 split, both types hit borrowing constraint → B = -0.5
        asset_supply[ρ] = -0.5
        println("ρ=1.0: absorbing states → B = -0.5 (by argument)")
    else
        μ = stationary_dist(sol, ρ, p)
        B = agg_asset_supply(μ, sol)
        asset_supply[ρ] = B
        println("Asset supply: B = $B")
    end
end

# Print table
println("\n===== Q1: Asset Supply by ρ =====")
println("ρ \t\t Asset Supply")
println("—"^35)
for ρ in ρ_vals
    @printf("%.2f \t\t %.6f\n", ρ, asset_supply[ρ])
end



# ============================================================
# Question 2a: Counterfactual asset supply
# Keep c = c^{0.5} but use income process for ρ=0 and ρ=0.95
# ============================================================

# Benchmark solution: ρ = 0.5
sol_benchmark = solutions[0.5]

# Counterfactual: use benchmark policy but different ρ for income transitions
B_cf = Dict{Float64, Float64}()
for ρ_cf in [0.0, 0.95]
    # Use the benchmark solution's policy_idx, but compute stationary dist under ρ_cf
    μ_cf = stationary_dist(sol_benchmark, ρ_cf, p)
    B_cf[ρ_cf] = agg_asset_supply(μ_cf, sol_benchmark)
    println("B(ρ=$ρ_cf, c=c^0.5) = $(B_cf[ρ_cf])")
end



# ============================================================
# Question 2b: Decomposition table
# ============================================================

B_benchmark = asset_supply[0.5]

# ρ = 0 decomposition
T_total_0 = asset_supply[0.0] - B_benchmark
T1_0 = asset_supply[0.0] - B_cf[0.0]
T2_0 = B_cf[0.0] - B_benchmark

# ρ = 0.95 decomposition
T_total_95 = asset_supply[0.95] - B_benchmark
T1_95 = asset_supply[0.95] - B_cf[0.95]
T2_95 = B_cf[0.95] - B_benchmark

# Print decomposition table
println("\n===== Q2b: Decomposition =====")
println("ρ \t T_total \t T1 (%) \t\t T2 (%)")
println("—"^55)
@printf("0.00 \t %.4f \t\t %.1f \t\t %.1f\n", T_total_0, 100*T1_0/T_total_0, 100*T2_0/T_total_0)
@printf("0.95 \t %.4f \t\t %.1f \t\t %.1f\n", T_total_95, 100*T1_95/T_total_95, 100*T2_95/T_total_95)
```


\newpage
## - Exercise 2 Parameters
```julia
# Part 1, Exercise 2: Parameters

# ex2_parameters.jl
using Parameters

Params = @with_kw (
    nb = 50,
    b_min = -0.5,
    b_max = 5.0,
    yL = 1.0,
    yH = 2.0,
    β = 0.85,
    γ = 1.5,
    r = 0.10,
    tol_vfi = 1e-6,
    tol_kfe = 1e-6,
    max_iter = 500
)

trans_matrix(ρ) = [ρ (1-ρ); (1-ρ) ρ]
```

\newpage
## - Exercise 2 Value Function Iteration
```julia
# Part 1, Exercise 2: Value Function Iteration

# ex2_vfi.jl

function solve_household(p, ρ::Float64)
    
    # Unpack

        nb = p.nb
        b_grid = collect(LinRange(p.b_min, p.b_max, nb))
        y_grid = [p.yL, p.yH]
        ny = 2
        Π = trans_matrix(ρ)

    # CRRA utility

        function u(c)
            c > 0 ? (c^(1 - p.γ)) / (1 - p.γ) : -1e10
        end

    # Pre-compute consumption for all (b, y, b') triples
    # c(b, y, b') = (1+r)*b + y - b'
    # Store as matrix: rows = (b,y) states, cols = b' choices

        n_states = nb * ny
        c_mat = zeros(n_states, nb)
        for iy in 1:ny
            for ib in 1:nb
                idx = (iy - 1) * nb + ib
                for ib_next in 1:nb
                    c_mat[idx, ib_next] = (1 + p.r) * b_grid[ib] + y_grid[iy] - b_grid[ib_next]
                end
            end
        end

    # Utility matrix

        u_mat = u.(c_mat)

    # Initialise value function

        V = -ones(n_states)
        V_new = similar(V)

        # stores index of optimal b'
        policy_idx = zeros(Int, n_states) 

    # VFI loop

    converged = false
    for iter in 1:p.max_iter
        # Compute expected continuation value: EV(b') = Σ_y' Π(y,y') V(b',y')
        # V is ordered as [V(b1,y1), V(b2,y1),...,V(bN,y1), V(b1,y2),...,V(bN,y2)]
        
        # nb × ny
            V_mat = reshape(V, nb, ny)          
        
        # nb × ny (expected V for each b' and current y)
            EV_mat = V_mat * Π'                 
        
        # back to vector, same ordering as V
            EV = vec(EV_mat)                    

        # Maximise over b' for each state (b, y)
        
        for iy in 1:ny

            # EV(b', y) for current y
                EV_slice = EV[(iy-1)*nb+1 : iy*nb]  
                for ib in 1:nb
                    idx = (iy - 1) * nb + ib
                    best_val = -Inf
                    best_ib = 1
                    for ib_next in 1:nb
                        val = u_mat[idx, ib_next] + p.β * EV_slice[ib_next]
                        if val > best_val
                            best_val = val
                            best_ib = ib_next
                        end
                    end
                    V_new[idx] = best_val
                    policy_idx[idx] = best_ib
                end
        end

    # Check convergence
        
        if maximum(abs.(V_new - V)) < p.tol_vfi
            V .= V_new
            converged = true
            println("VFI converged after $iter iterations (ρ=$ρ)")
            break
        end
        V .= V_new
    end

    if !converged
        println("VFI did NOT converge for ρ=$ρ")
    end

    # Extract policy functions
        b_next = [b_grid[policy_idx[i]] for i in 1:n_states]
        c_policy = [(1 + p.r) * b_grid[mod1(i, nb)] + y_grid[div(i-1, nb)+1] - b_next[i] for i in 1:n_states]

    return (V=V, policy_idx=policy_idx, b_next=b_next, c_policy=c_policy,
            b_grid=b_grid, y_grid=y_grid, nb=nb, ny=ny, converged=converged)
end
```

\newpage
## - Exercise 2 Kolmogorov-Forward Equation
```julia
# Part 1, Exercise 2: Kolmogorov-Forward Equation
# kfe.jl — Compute stationary distribution over (b, y) states

function stationary_dist(sol, ρ::Float64, p)
    nb = sol.nb
    ny = sol.ny
    n_states = nb * ny
    Π = trans_matrix(ρ)

    # Build full transition matrix over (b, y) → (b', y')
    # For state i = (b, y): policy says go to b' = b_grid[policy_idx[i]]
    # Then y transitions according to Π
    T = zeros(n_states, n_states)

    for iy in 1:ny
        for ib in 1:nb
            idx_from = (iy - 1) * nb + ib
            ib_next = sol.policy_idx[idx_from]
            for iy_next in 1:ny
                idx_to = (iy_next - 1) * nb + ib_next
                T[idx_from, idx_to] = Π[iy, iy_next]
            end
        end
    end

    # Iterate to find stationary distribution: μ' = T' * μ
    μ = ones(n_states) ./ n_states  # uniform initial guess

    for k in 1:p.max_iter
        μ_old = copy(μ)
        μ = T' * μ
        if maximum(abs.(μ - μ_old)) < p.tol_kfe
            println("KFE converged after $k iterations")
            break
        end
    end

    return μ
end

# Compute aggregate asset supply: B = Σ μ(b,y) * b
function agg_asset_supply(μ, sol)
    b_grid_large = vcat(sol.b_grid, sol.b_grid)  # repeat for each y
    return dot(μ, b_grid_large)
end
```


