---
title: "Advanced Empirical Research Methods Hilary 2025/26"
author: 'Candidate ID: 1091806'
date: "2026-04-13"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
    toc_depth: 2
    number_sections: false
  html_document:
    toc: true
    toc_depth: '2'
    df_print: paged
subtitle: 'Assessment Exercise: Production Functions'
fontsize: 11pt
geometry: margin=1in
header-includes:
- \usepackage{amsmath}
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{amsmath}
- \usepackage{newunicodechar}
- \newunicodechar{ρ}{$\rho$}
- \newunicodechar{π}{$\pi$}
- \newunicodechar{ξ}{$\xi$}
- \newunicodechar{ϕ}{$\phi$}
---

\newpage
# Set up

## Model

We consider a two-factor Cobb-Douglas value-added production function in logs:

$$v_{it} \;=\; \beta_0 \,+\, \beta_L \, l_{it} \,+\, \beta_K \, k_{it} \,+\, a_{it}, \qquad \beta_L + \beta_K < 1$$

The price of value-added is normalised to one. Labour $L_{it}$ is perfectly flexible
and chosen in period $t$ after the firm observes total factor productivity
$A_{it} = \exp(a_{it})$ and the wage $W_{it}$. Capital $K_{it}$ is predetermined and
chosen in period $t-1$ to maximise expected period-$t$ profit. There are no
adjustment costs for either input, and the underlying gross-output technology is
Leontief in value-added and a flexible intermediate input.

## Stochastic processes

Total factor productivity follows an AR(1):
$$a_{it} \;=\; \rho \, a_{i,t-1} \,+\, \xi_{it}, \qquad \xi_{it} \sim \mathrm{i.i.d.}\, N(0, \sigma_\xi^2), \qquad 0 < \rho < 1.$$

The log wage follows an MA(1):
$$w_{it} \;=\; w \,+\, c_{it} \,+\, \phi \, c_{i,t-1}, \qquad c_{it} \sim \mathrm{i.i.d.}\, N(0, \sigma_c^2).$$

The rental rate of capital $U$ is **fixed and common across firms and time**.

The observed log of value-added is contaminated with classical measurement error:
$$\tilde v_{it} \;=\; v_{it} \,+\, \varepsilon_{it}, \qquad \varepsilon_{it} \sim \mathrm{i.i.d.}\, N(0, \sigma_\varepsilon^2).$$

## Information sets

The firm's information set at $t-1$ includes $a_{i,t-1}$, $w_{i,t-1}$, $U$, and the
fixed value-added price; it excludes $a_{it}$ and $w_{it}$. So $K_{it}$ is
uncorrelated with $\xi_{it}$ by construction, while $L_{it}$ — chosen after
$\xi_{it}$ is realised — is not.

## Observed data

Only the input and output quantities $\bigl(\tilde V_{it} = e^{\tilde v_{it}},\, L_{it},\, K_{it}\bigr)$ are observed.
Wages, the rental rate, intermediate inputs, investment, and TFP are all unobserved.
This rules out the GNR labour-share estimator (no $W$), the LP/ACF intermediate-input
proxy (no $M$), the OP investment proxy (no $I$), and the wage instrument (no $W$).

\newpage

## Calibration for the simulations

| Parameter                       | Symbol            | Value                       |
|---------------------------------|-------------------|-----------------------------|
| Output elasticity of labour     | $\beta_L$         | 0.6                         |
| Output elasticity of capital    | $\beta_K$         | 0.3                         |
| AR(1) persistence of TFP        | $\rho$            | 0.5                         |
| Variance of TFP innovation      | $\sigma_\xi^2$    | 0.05                        |
| Mean log wage                   | $w$               | $\log(1) = 0$               |
| Variance of wage innovation     | $\sigma_c^2$      | 0.05                        |
| Log rental rate (fixed)         | $u$               | $\log(0.1)$                 |
| Variance of measurement error   | $\sigma_\varepsilon^2$ | 0.1                    |
| MA(1) wage parameter            | $\phi$            | 0 (Q1, Q2b, Q2d); 1 (Q2c)   |
| Number of periods               | $T$               | 100,001                     |

## R Set-up

```{r r-setup, message=FALSE, warning=FALSE}
# Load libraries

  library(dplyr)    # data manipulation
  library(psych)    # geometric.mean()
  library(AER)      # ivreg(), linearHypothesis()
  library(gmm)      # gmm()
  library(knitr)    # kable()

# Set seed

  set.seed(1234)
```


## Reusable simulation function

To keep the rest of the document concise, we wrap the DGP in a single function.
It takes the two switches we toggle across questions ($\phi$ and `urndvar`) plus
the standard parameters, and returns a data frame with the variables we will use:
$\tilde v$, $l$, $k$, plus their lags and (for diagnostic purposes) the latent
$a$ and $w$.

```{r sim-function}
simulate_pf <- function(obs   = 100001,
                        bL    = 0.6,
                        bK    = 0.3,
                        rho   = 0.5,
                        shkvar = 0.05,
                        wfix   = log(1),
                        wrndvar = 0.05,
                        phi    = 0,
                        ufix   = log(0.1),
                        urndvar = 0,
                        merrvar = 0.1) {

  
  # AR(1) TFP
  
    shk  <- rnorm(obs, mean = 0, sd = sqrt(shkvar))
    stat <- rnorm(obs, mean = 0, sd = sqrt(shkvar / (1 - rho^2)))
    a    <- stat
    for (i in 2:obs) a[i] <- rho * a[i - 1] + shk[i]

  # Rental rate (fixed if urndvar = 0)
    
    u <- rnorm(obs, mean = ufix, sd = sqrt(urndvar))

  # MA(1) log wage

    wrnd <- rnorm(obs, mean = 0, sd = sqrt(wrndvar))
    w    <- wfix + wrnd + phi * lag(wrnd)

  # Capital decision rule (predetermined)
      
    rat     <- bK / (1 - bL)
    fac     <- 1 / (1 - rat)
    pow     <- 1 / (1 - bL)
    tau     <- bL^(bL * pow) - bL^pow
    logetau <- log(tau) - bL * pow * wfix - phi * bL * pow * lag(wrnd) +
               ((bL * pow)^2) * wrndvar / 2
    k <- fac * (log(rat) + logetau - u + pow * rho * lag(a) +
                0.5 * shkvar * pow^2)

  # True log value-added (without measurement error)
 
    v_true <- bL * pow * (log(bL) - w) + rat * k + pow * a

  # Flexible labour decision rule
  
    l <- pow * (log(bL) - w + bK * k + a)

  # Add measurement error to obtain observed log value-added
  
    merr <- rnorm(obs, mean = 0, sd = sqrt(merrvar))
    vt   <- v_true + merr
  
    data.frame(t = seq_len(obs), vt = vt, l = l, k = k,
               a = a, w = w, v_true = v_true)
}
```

A quick sanity check that the simulation matches the assumed DGP (the
infeasible OLS regression of $\tilde v$ on $l$, $k$, and the true $a$ should
recover $(\beta_0, \beta_L, \beta_K)$ up to sampling error) while feasible OLS
omitting $a$ should not.

```{r sim-check}
dat0 <- simulate_pf(phi = 0)

# Infeasible OLS — should give bL = 0.6, bK = 0.3

  summary(lm(vt ~ l + k + a, data = dat0))$coefficients

# Feasible OLS omitting a — biased (transmission bias)

  summary(lm(vt ~ l + k,    data = dat0))$coefficients
```

If the infeasible OLS coefficients land near $(\beta_L, \beta_K) = (0.6, 0.3)$
and the feasible OLS does not, the DGP is correctly implemented.

\newpage
# Question 1 

Throughout Q1 we impose $\phi = 0$, that means the log wage is i.i.d. across firms and time.

## Q1(a) Two-stage procedure

**Capital as proxy for past TFP.** With $U$ fixed, $\phi = 0$, and no adjustment costs, the FOC for capital combined with the AR(1) for $a$ gives an *exactly* log-linear capital decision rule:

$$k_{it} \;=\; \kappa_0 \,+\, \gamma\, a_{i,t-1}, \qquad \gamma \;=\; \frac{\rho}{1-\beta_L-\beta_K}.$$

The map is invertible, so $a_{i,t-1} = (k_{it}-\kappa_0)/\gamma \equiv g(k_{it})$, capital alone is a sufficient statistic for past productivity.

**First stage: Olley–Pakes style ($\beta_L$)**

Replacing $a_{it} = \rho\, g(k_{it}) + \xi_{it}$ in the production function gives
$$\tilde v_{it} \;=\; \beta_L\, l_{it} \,+\, \varphi(k_{it}) \,+\, \xi_{it} \,+\, \varepsilon_{it}, \qquad \varphi(k_{it}) \equiv \beta_0 + \beta_K k_{it} + \rho\, g(k_{it}).$$
We estimate by OLS regression of $\tilde v_{it}$ on $l_{it}$ and a polynomial in $k_{it}$, and read $\hat\beta_L$ off the coefficient on $l_{it}$. Following the OP (1996) argument: conditional on $k_{it}$, identifying variation in $l_{it}$ comes from the i.i.d. wage shock $w_{it}$, which by construction is independent of $\xi_{it}$ and $\varepsilon_{it}$. The "serially uncorrelated wage" condition that OP requires is delivered here by $\phi = 0$, and the "no firm-level variation in the user cost" condition is delivered by $U$ being fixed.

**Second stage: Blundell–Bond dynamic GMM ($\beta_K$)**

Form the cleaned dependent variable $y_{it} \equiv \tilde v_{it} - \hat\beta_L\, l_{it}$, which is consistent for $\beta_0 + \beta_K k_{it} + a_{it} + \varepsilon_{it}$. Quasi-differencing using the AR(1) yields
$$y_{it} \,-\, \rho\, y_{i,t-1} \;=\; \beta_0(1-\rho) \,+\, \beta_K\bigl(k_{it} - \rho\, k_{i,t-1}\bigr) \,+\, \xi_{it} \,+\, \varepsilon_{it} - \rho\,\varepsilon_{i,t-1}.$$

Estimate $(\beta_K, \rho)$ jointly by non-linear GMM with instruments
$$z_{it} \;=\; (1,\, k_{it},\, k_{i,t-1},\, k_{i,t-2},\, y_{i,t-2}).$$

- *Validity:* $k_{i,t-s}$ for $s \ge 0$ lies in the firm's information set at $t-1$ and is therefore orthogonal to $\xi_{it}$; $k_{i,t-2}$ and $y_{i,t-2}$ are dated before $t-1$ and so also orthogonal to $\varepsilon_{i,t-1}$. 

- *Relevance:* the AR(1) persistence in $a_{it}$ induces serial correlation in $k$ and $y$, so the instruments are correlated with the endogenous regressors.


\newpage
## Q1(b) Simulation and two-stage estimation

Simulate under the Q1 DGP ($\phi = 0$, $U$ fixed) and build the leads and lags
needed for both stages.

```{r q1b-setup}
dat_q1 <- simulate_pf(phi = 0, urndvar = 0)

dat_q1 <- dat_q1 %>%
  mutate(
    k_p1 = lead(k),   # k_{i,t+1} — acts as the investment-proxy
    k_m1 = lag(k),
    k_m2 = lag(k, 2))
```

**Stage 1**

Run the OP first-stage regression to recover $\beta_L$. We use a second-order polynomial in
$(k_{it}, k_{i,t+1})$, which is general enough to handle nonlinear
$\varphi$ in principle; under this DGP the higher-order terms should come out
near zero.
 
The model is:

$$\tilde v_{it} = \beta_L * l_{it} + \varphi(k_{it}, k_{t+1}) + \epsilon_{it}$$

Under this DGP varphi is affine, but we include quadratic and interaction terms as a hedge, they should come out near zero, which is the point we will exploit in Q1(c).

```{r q1b-stage1}
# Fit the first-stage regression of log value-added on labour plus a
# full second-order polynomial in (k_t, k_{t+1})

  stage1 <- lm(vt ~ l + k + k_p1 + I(k^2) + I(k_p1^2) + I(k * k_p1),
               data = dat_q1, na.action = na.exclude)

# Report the coefficient estimates and standard errors

  round(summary(stage1)$coefficients[, 1:2], 4)

# Extract the coefficient on l — this is our Stage 1 estimate of beta_L

  bL_hat <- coef(stage1)["l"]
  bL_hat
```

**Stage 2**

Form the cleaned dependent variable $y_{it} = \tilde v_{it} - \hat\beta_L\, l_{it}$ and run Blundell–Bond dynamic GMM to recover $\beta_K$ and $\rho$.

The moment function:

$$u_{it} \;=\; y_{it} - \rho\, y_{i,t-1} - \beta_K (k_{it} - \rho\, k_{i,t-1}) - \beta_0(1 - \rho),$$

and the instruments are $(k_{it},\, k_{i,t-1},\, k_{i,t-2},\, y_{i,t-2},\, 1)$, five moments for three parameters, so the system is over-identified.

```{r q1b-stage2}
# ---------------------------------------------------------------------
# Quasi-differenced equation (after subtracting bL_hat * l from vt):
#   y_t - rho * y_{t-1} = beta_0*(1-rho) + beta_K*(k_t - rho*k_{t-1})
#                       + xi_t + eps_t - rho*eps_{t-1}

# Instruments: (k_t, k_{t-1}, k_{t-2}, y_{t-2}, 1) — dated at t-1 or
# earlier, so orthogonal to xi_t and to eps_{t-1}
# ---------------------------------------------------------------------

# Build the cleaned dependent variable y_t and its lags
  
  dat_q1 <- dat_q1 |> 
    mutate(
      y    = vt - bL_hat * l,
      y_m1 = lag(y),
      y_m2 = lag(y, 2)
    )

# Stack the six columns needed by the moment function and drop rows
# with NAs from lagging (first two periods)

  X_gmm <- with(dat_q1, cbind(y, y_m1, k, k_m1, y_m2, k_m2)) |>  na.omit()


# Moment function: returns a matrix whose columns are the five moment
# conditions evaluated at each observation. GMM will minimise the
# quadratic form of their sample means
  
  moms_stage2 <- function(theta, x) {
    
    # theta = (rho, bK, b0)
    u <- x[,1] - theta[1]*x[,2] -
         theta[2]*(x[,3] - theta[1]*x[,4]) -
         theta[3]*(1 - theta[1])
    cbind(x[,3]*u,   # k_t   instrument
          x[,4]*u,   # k_{t-1}
          x[,6]*u,   # k_{t-2}
          x[,5]*u,   # y_{t-2}
          u)         # constant
  }

# Run GMM with starting values (rho, bK, b0) = (0.3, 0.3, 0)
# With 5 moments and 3 parameters the system is over-identified
  
  stage2 <- gmm(moms_stage2, X_gmm,
                t0 = c(rho = 0.3, bK = 0.3, b0 = 0))

  summary(stage2)
```

\newpage

**Results**

```{r q1b-results}
results_q1 <- data.frame(
  Parameter = c("beta_L", "beta_K", "rho"),
  True      = c(0.6, 0.3, 0.5),
  Estimate  = c(as.numeric(bL_hat),
                coef(stage2)["bK"],
                coef(stage2)["rho"])
)

kable(results_q1, digits = 3,
      caption = "Q1(b) — Two-stage estimates vs. true values")
```

With $T = 100{,}001$ all three estimates should be within two decimal places of
the truth. The Hansen $J$ statistic reported by `gmm()` should not reject the
over-identifying restrictions.


\newpage
## Q1(c) Is the second-stage needed?

**No.** Under this DGP the first-stage function $\varphi(k_{it}, k_{i,t+1})$ is
not a generic non-linear object, it is exactly affine:
$$\varphi(k_{it}, k_{i,t+1}) \;=\; \beta_K\, k_{it} \,+\, \tfrac{1}{\gamma}\, k_{i,t+1} \,+\, \text{const}.$$

So a *linear* first stage — OLS of $\tilde v_{it}$ on $l_{it}$, $k_{it}$, and
$k_{i,t+1}$ — identifies $\beta_L$ *and* $\beta_K$ in a single regression, as
the coefficients on $l_{it}$ and $k_{it}$ respectively. The second stage is
redundant.

Three features of the DGP combine to make $\varphi$ affine rather than
genuinely non-linear:

1. **No adjustment costs for capital.** The capital Euler equation does not
   involve lagged capital as an additional state, so the policy function for
   $k_{i,t+1}$ is log-linear in $a_{it}$ rather than a non-trivial function of
   multiple states.

2. **User cost $U$ fixed across firms.** No firm-level $u_{it}$ enters the
   capital rule, so $k_{i,t+1}$ depends on $a_{it}$ alone — not on the pair
   $(a_{it}, u_{it})$.

3. **$\phi = 0$.** Wage innovations are i.i.d., so lagged $c_{i,t-1}$ does not
   forecast $w_{it}$ and therefore does not feed into the period-$t$ choice of
   $k_{i,t+1}$ via rational expectations of next period's wage.

Together these deliver $k_{i,t+1} = \kappa_0 + \gamma\, a_{it}$, exactly
invertible in $k_{i,t+1}$. Substituting back into the production function
yields the affine $\varphi$ above, with $\beta_K$ appearing cleanly as the
coefficient on $k_{it}$ (its direct structural role) and $1/\gamma$ as the
coefficient on $k_{i,t+1}$ (the inverse of the capital rule's sensitivity to
$a_{it}$).

**A sharper observation.** Not only is the second stage unnecessary — the
*standard* OP second stage is under-identified here. In the OP formulation,
one would regress $y_{it} = \tilde v_{it} - \hat\beta_L l_{it}$ on $k_{it}$
and a flexible function $h(k_{i,t-1}, k_{it})$ capturing $\rho\, a_{i,t-1}$.
Under the exact log-linear capital rule, $h$ collapses to a linear function
of $k_{it}$ (and $k_{i,t-1}$), and the OP second stage loses rank: it only
identifies the sum $\beta_K + \rho/\gamma = 1 - \beta_L$, not $\beta_K$ and
$\rho$ separately. This is the Bond–Söderbom (2005) non-identification
result. The BB-style second stage used in Q1(b) avoids the collapse by
exploiting over-identifying moments from $k_{i,t-2}$ and $y_{i,t-2}$, which is
why it continues to work — but at that point we are no longer using the OP
second stage in any recognisable form.

### Numerical confirmation

*Linear first stage:* Under this DGP varphi is affine, so no polynomial terms are needed: a simple OLS of vt on (`l`, `k`, `k_{t+1}`) delivers $\beta_L$ and $\beta_K$ (and $1/\gamma$ as a bonus) in one shot.

```{r q1c-linear-first-stage}
# Fit the linear first stage with just labour, capital, and next-period capital

  stage1_lin <- lm(vt ~ l + k + k_p1,
                   data = dat_q1,
                   na.action = na.exclude)

# Report the four coefficient estimates and their standard errors

  round(summary(stage1_lin)$coefficients, 4)
```

Expected against the calibration: the coefficient on `l` $\approx \beta_L = 0.6$;
the coefficient on `k` $\approx \beta_K = 0.3$ (read directly from Stage 1,
matching the Q1(b) BB-GMM estimate); the coefficient on `k_p1`
$\approx 1/\gamma = 0.2$, since $\gamma = \rho/(1-\beta_L-\beta_K) = 5$.

*Sanity check:* the coefficient on $k_{t+1}$ in the linear first stage should numerically equal $1/\gamma$ computed from the structural parameters. This confirms the affine decomposition of varphi.

```{r q1c-gamma-check}
# Calibration values, taken from the Setup section

  bL_cal  <- 0.6
  bK_cal  <- 0.3
  rho_cal <- 0.5

# gamma is the sensitivity of k_{t+1} to current TFP under the capital rule
  
  gamma_true <- rho_cal / (1 - bL_cal - bK_cal)   # = 0.5 / 0.1 = 5
  inv_gamma  <- 1 / gamma_true                     # = 0.2

# Side-by-side comparison
  
  c(`1/gamma (from calibration)` = inv_gamma,
    `k_p1 coefficient (OLS)`     = as.numeric(coef(stage1_lin)["k_p1"]))
```

*Final check*: in the polynomial first stage from Q1(b), the quadratic and interaction terms should have coefficients numerically indistinguishable from zero,  i.e. varphi has no curvature in this DGP.

```{r q1c-polynomial-terms}
# Extract the estimated coefficients on k^2, k_{t+1}^2, and k * k_{t+1}
# from the polynomial first stage (`stage1` from Q1(b))

  round(summary(stage1)$coefficients[c("I(k^2)",
                                       "I(k_p1^2)",
                                       "I(k * k_p1)"), 1:2], 5)
```

If these three coefficients are (a) small in magnitude and (b) statistically
indistinguishable from zero, we have a direct numerical confirmation that
$\varphi$ is affine under this DGP — and hence that the first-stage
specification alone identifies both $\beta_L$ and $\beta_K$ consistently,
with no second stage required.

\newpage
# Question 2



## Q2 (a)

\newpage
## Q2 (b)

\newpage
## Q2 (c)







