---
title: "Advanced Empirical Research Methods Hilary 2025/26"
author: 'Candidate ID: 1091806'
date: "2026-04-13"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
    toc_depth: 2
    number_sections: false
  html_document:
    toc: true
    toc_depth: '2'
    df_print: paged
subtitle: 'Assessment Exercise: Production Functions'
fontsize: 11pt
geometry: margin=1in
header-includes:
- \usepackage{amsmath}
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{amsmath}
- \usepackage{newunicodechar}
- \newunicodechar{ρ}{$\rho$}
- \newunicodechar{π}{$\pi$}
- \newunicodechar{ξ}{$\xi$}
- \newunicodechar{ϕ}{$\phi$}
editor_options: 
  markdown: 
    wrap: 72
---

\newpage

# Set up

## Model

We consider a two-factor Cobb-Douglas value-added production function in
logs:

$$v_{it} \;=\; \beta_0 \,+\, \beta_L \, l_{it} \,+\, \beta_K \, k_{it} \,+\, a_{it}, \qquad \beta_L + \beta_K < 1$$

The price of value-added is normalised to one. Labour $L_{it}$ is
perfectly flexible and chosen in period $t$ after the firm observes
total factor productivity $A_{it} = \exp(a_{it})$ and the wage $W_{it}$.
Capital $K_{it}$ is predetermined and chosen in period $t-1$ to maximise
expected period-$t$ profit. There are no adjustment costs for either
input, and the underlying gross-output technology is Leontief in
value-added and a flexible intermediate input.

## Stochastic processes

Total factor productivity follows an AR(1):
$$a_{it} \;=\; \rho \, a_{i,t-1} \,+\, \xi_{it}, \qquad \xi_{it} \sim \mathrm{i.i.d.}\, N(0, \sigma_\xi^2), \qquad 0 < \rho < 1.$$

The log wage follows an MA(1):
$$w_{it} \;=\; w \,+\, c_{it} \,+\, \phi \, c_{i,t-1}, \qquad c_{it} \sim \mathrm{i.i.d.}\, N(0, \sigma_c^2).$$

The rental rate of capital $U$ is **fixed and common across firms and
time**.

The observed log of value-added is contaminated with measurement error:
$$\tilde v_{it} \;=\; v_{it} \,+\, \varepsilon_{it}, \qquad \varepsilon_{it} \sim \mathrm{i.i.d.}\, N(0, \sigma_\varepsilon^2).$$

## Information sets

The firm's information set at $t-1$ includes $a_{i,t-1}$, $w_{i,t-1}$,
$U$, and the fixed value-added price. It excludes $a_{it}$ and $w_{it}$.
So $K_{it}$ is uncorrelated with $\xi_{it}$ by construction, while
$L_{it}$ (chosen after $\xi_{it}$ is realised) is not.

## Observed data

Only the input and output quantities
$\bigl(\tilde V_{it} = e^{\tilde v_{it}},\, L_{it},\, K_{it}\bigr)$ are
observed. Wages, the rental rate, intermediate inputs, investment, and
TFP are all unobserved. This rules out the GNR labour-share estimator
(no $W$), the LP/ACF intermediate-input proxy (no $M$), the OP
investment proxy (no $I$), and the wage instrument (no $W$).

\newpage

## Calibration for the simulations

| Parameter | Symbol | Value |
|-----------------------------|------------------|-------------------------|
| Output elasticity of labour | $\beta_L$ | 0.6 |
| Output elasticity of capital | $\beta_K$ | 0.3 |
| AR(1) persistence of TFP | $\rho$ | 0.5 |
| Variance of TFP innovation | $\sigma_\xi^2$ | 0.05 |
| Mean log wage | $w$ | $\log(1) = 0$ |
| Variance of wage innovation | $\sigma_c^2$ | 0.05 |
| Log rental rate (fixed) | $u$ | $\log(0.1)$ |
| Variance of measurement error | $\sigma_\varepsilon^2$ | 0.1 |
| MA(1) wage parameter | $\phi$ | 0 (Q1, Q2b, Q2d); 1 (Q2c) |
| Number of periods | $T$ | 100,001 |

## R Set-up

```{r r-setup, message=FALSE, warning=FALSE}
# Load libraries

  library(dplyr)    # data manipulation
  library(psych)    # geometric.mean()
  library(AER)      # ivreg(), linearHypothesis()
  library(gmm)      # gmm()
  library(knitr)    # kable()

# Set seed

  set.seed(1234)
```

## Reusable simulation function

To keep the rest of the document concise, we wrap the DGP in a single
function. It takes the two switches we toggle across questions ($\phi$
and `urndvar`) plus the standard parameters, and returns a data frame
with the variables we will use: $\tilde v$, $l$, $k$, plus their lags
and (for diagnostic purposes) the latent $a$ and $w$.

```{r sim-function}
simulate_pf <- function(obs   = 100001,
                        bL    = 0.6,
                        bK    = 0.3,
                        rho   = 0.5,
                        shkvar = 0.05,
                        wfix   = log(1),
                        wrndvar = 0.05,
                        phi    = 0,
                        ufix   = log(0.1),
                        urndvar = 0,
                        merrvar = 0.1) {

  
  # AR(1) TFP
  
    shk  <- rnorm(obs, mean = 0, sd = sqrt(shkvar))
    stat <- rnorm(obs, mean = 0, sd = sqrt(shkvar / (1 - rho^2)))
    a    <- stat
    for (i in 2:obs) a[i] <- rho * a[i - 1] + shk[i]

  # Rental rate (fixed if urndvar = 0)
    
    u <- rnorm(obs, mean = ufix, sd = sqrt(urndvar))

  # MA(1) log wage

    wrnd <- rnorm(obs, mean = 0, sd = sqrt(wrndvar))
    w    <- wfix + wrnd + phi * lag(wrnd)

  # Capital decision rule (predetermined)
      
    rat     <- bK / (1 - bL)
    fac     <- 1 / (1 - rat)
    pow     <- 1 / (1 - bL)
    tau     <- bL^(bL * pow) - bL^pow
    logetau <- log(tau) - bL * pow * wfix - phi * bL * pow * lag(wrnd) +
               ((bL * pow)^2) * wrndvar / 2
    k <- fac * (log(rat) + logetau - u + pow * rho * lag(a) +
                0.5 * shkvar * pow^2)

  # True log value-added (without measurement error)
 
    v_true <- bL * pow * (log(bL) - w) + rat * k + pow * a

  # Flexible labour decision rule
  
    l <- pow * (log(bL) - w + bK * k + a)

  # Add measurement error to obtain observed log value-added
  
    merr <- rnorm(obs, mean = 0, sd = sqrt(merrvar))
    vt   <- v_true + merr
  
    data.frame(t = seq_len(obs), vt = vt, l = l, k = k,
               a = a, w = w, v_true = v_true)
}
```

A quick sanity check that the simulation matches the assumed DGP (the
infeasible OLS regression of $\tilde v$ on $l$, $k$, and the true $a$
should recover $(\beta_0, \beta_L, \beta_K)$ up to sampling error) while
feasible OLS omitting $a$ should not.

```{r sim-check}
dat0 <- simulate_pf(phi = 0)

# Infeasible OLS: should give bL = 0.6, bK = 0.3

  summary(lm(vt ~ l + k + a, data = dat0))$coefficients

# Feasible OLS omitting a: biased (transmission bias)

  summary(lm(vt ~ l + k,    data = dat0))$coefficients
```

If the infeasible OLS coefficients land near
$(\beta_L, \beta_K) = (0.6, 0.3)$ and the feasible OLS does not, the DGP
is correctly implemented.

\newpage

# Question 1

Throughout Q1 we impose $\phi = 0$, that means the log wage is i.i.d.
across firms and time.

## Q1(a) Two-stage procedure

With $U$ fixed, $\phi = 0$, and no
adjustment costs, the FOC for capital combined with the AR(1) for $a$
gives an *exactly* log-linear capital decision rule:

$$k_{it} \;=\; \kappa_0 \,+\, \gamma\, a_{i,t-1}, \qquad \gamma \;=\; \frac{\rho}{1-\beta_L-\beta_K}.$$

The map is invertible, so
$a_{i,t-1} = (k_{it}-\kappa_0)/\gamma \equiv g(k_{it})$, capital alone
is a sufficient statistic for past productivity.

**First stage: Olley–Pakes style** ($\beta_L$)

Replacing $a_{it} = \rho\, g(k_{it}) + \xi_{it}$ in the production
function gives
$$\tilde v_{it} \;=\; \beta_L\, l_{it} \,+\, \varphi(k_{it}) \,+\, \xi_{it} \,+\, \varepsilon_{it}, \qquad \varphi(k_{it}) \equiv \beta_0 + \beta_K k_{it} + \rho\, g(k_{it}).$$
We estimate by OLS regression of $\tilde v_{it}$ on $l_{it}$ and a
polynomial in $k_{it}$, and read $\hat\beta_L$ off the coefficient on
$l_{it}$. Following the OP (1996) argument: conditional on $k_{it}$,
identifying variation in $l_{it}$ comes from the i.i.d. wage shock
$w_{it}$, which by construction is independent of $\xi_{it}$ and
$\varepsilon_{it}$. The "serially uncorrelated wage" condition that OP
requires is delivered here by $\phi = 0$, and the "no firm-level
variation in the user cost" condition is delivered by $U$ being fixed.

**Second stage: Blundell–Bond dynamic GMM (**$\beta_K$)

Form the cleaned dependent variable
$y_{it} \equiv \tilde v_{it} - \hat\beta_L\, l_{it}$, which is
consistent for $\beta_0 + \beta_K k_{it} + a_{it} + \varepsilon_{it}$.
Quasi-differencing using the AR(1) yields
$$y_{it} \,-\, \rho\, y_{i,t-1} \;=\; \beta_0(1-\rho) \,+\, \beta_K\bigl(k_{it} - \rho\, k_{i,t-1}\bigr) \,+\, \xi_{it} \,+\, \varepsilon_{it} - \rho\,\varepsilon_{i,t-1}.$$

Estimate $(\beta_K, \rho)$ jointly by non-linear GMM with instruments
$$z_{it} \;=\; (1,\, k_{it},\, k_{i,t-1},\, k_{i,t-2},\, y_{i,t-2}).$$

-   *Validity:* $k_{i,t-s}$ for $s \ge 0$ lies in the firm's information
    set at $t-1$ and is therefore orthogonal to $\xi_{it}$; $k_{i,t-2}$
    and $y_{i,t-2}$ are dated before $t-1$ and so also orthogonal to
    $\varepsilon_{i,t-1}$.

-   *Relevance:* the AR(1) persistence in $a_{it}$ induces serial
    correlation in $k$ and $y$, so the instruments are correlated with
    the endogenous regressors.

\newpage

## Q1(b) Simulation and two-stage estimation

Simulate under the Q1 DGP ($\phi = 0$, $U$ fixed) and build the leads
and lags needed for both stages.

```{r q1b-setup}
dat_q1 <- simulate_pf(phi = 0, urndvar = 0)

dat_q1 <- dat_q1 |> 
  mutate(
    k_p1 = lead(k),   # k_{i,t+1} acts as the investment-proxy
    k_m1 = lag(k),
    k_m2 = lag(k, 2))
```

**Stage 1**

Run the OP first-stage regression to recover $\beta_L$. We use a
second-order polynomial in $(k_{it}, k_{i,t+1})$, which is general
enough to handle nonlinear $\varphi$ in principle; under this DGP the
higher-order terms should come out near zero.

The model is:

$$\tilde v_{it} = \beta_L * l_{it} + \varphi(k_{it}, k_{t+1}) + \epsilon_{it}$$

Under this DGP varphi is affine, but we include quadratic and
interaction terms as a hedge, they should come out near zero.

```{r q1b-stage1}
# Fit the first-stage regression of log value-added on labour plus a
# full second-order polynomial in (k_t, k_{t+1})

  stage1 <- lm(vt ~ l + k + k_p1 + I(k^2) + I(k_p1^2) + I(k * k_p1),
               data = dat_q1, na.action = na.exclude)

# Report the coefficient estimates and standard errors

  round(summary(stage1)$coefficients[, 1:2], 4)

# Extract the coefficient on l, this is our Stage 1 estimate of beta_L

  bL_hat <- coef(stage1)["l"]
  bL_hat
```

\newpage

**Stage 2**

Form the cleaned dependent variable
$y_{it} = \tilde v_{it} - \hat\beta_L\, l_{it}$ and run Blundell–Bond
dynamic GMM to recover $\beta_K$ and $\rho$.

The moment function:

$$u_{it} \;=\; y_{it} - \rho\, y_{i,t-1} - \beta_K (k_{it} - \rho\, k_{i,t-1}) - \beta_0(1 - \rho),$$

and the instruments are
$(k_{it},\, k_{i,t-1},\, k_{i,t-2},\, y_{i,t-2},\, 1)$, five moments for
three parameters, so the system is over-identified.

```{r q1b-stage2}
# ---------------------------------------------------------------------
# Quasi-differenced equation (after subtracting bL_hat * l from vt):
#   y_t - rho * y_{t-1} = beta_0*(1-rho) + beta_K*(k_t - rho*k_{t-1})
#                       + xi_t + eps_t - rho*eps_{t-1}

# Instruments: (k_t, k_{t-1}, k_{t-2}, y_{t-2}, 1), dated at t-1 or
# earlier, so orthogonal to xi_t and to eps_{t-1}
# ---------------------------------------------------------------------

# Build the cleaned dependent variable y_t and its lags
  
  dat_q1 <- dat_q1 |> 
    mutate(
      y    = vt - bL_hat * l,
      y_m1 = lag(y),
      y_m2 = lag(y, 2)
    )

# Stack the six columns needed by the moment function and drop rows
# with NAs from lagging (first two periods)

  X_gmm <- with(dat_q1, cbind(y, y_m1, k, k_m1, y_m2, k_m2)) |>  na.omit()


# Moment function: returns a matrix whose columns are the five moment
# conditions evaluated at each observation. GMM will minimise the
# quadratic form of their sample means
  
  moms_stage2 <- function(theta, x) {
    
    # theta = (rho, bK, b0)
    u <- x[,1] - theta[1]*x[,2] -
         theta[2]*(x[,3] - theta[1]*x[,4]) -
         theta[3]*(1 - theta[1])
    cbind(x[,3]*u,   # k_t   instrument
          x[,4]*u,   # k_{t-1}
          x[,6]*u,   # k_{t-2}
          x[,5]*u,   # y_{t-2}
          u)         # constant
  }

# Run GMM with starting values (rho, bK, b0) = (0.3, 0.3, 0)
# With 5 moments and 3 parameters the system is over-identified
  
  stage2 <- gmm(moms_stage2, X_gmm,
                t0 = c(rho = 0.3, bK = 0.3, b0 = 0))

  summary(stage2)
```

\newpage

**Results**

```{r q1b-results}
results_q1 <- data.frame(
  Parameter = c("beta_L", "beta_K", "rho"),
  True      = c(0.6, 0.3, 0.5),
  Estimate  = c(as.numeric(bL_hat),
                coef(stage2)["bK"],
                coef(stage2)["rho"])
)

kable(results_q1, digits = 3,
      caption = "Q1(b) Two-stage estimates vs. true values")
```

With $T = 100{,}001$ all three estimates should be within two decimal
places of the truth. The Hansen $J$ statistic reported by `gmm()` should
not reject the over-identifying restrictions.

\newpage

## Q1(c) Is the second-stage needed?

**No.** Under this DGP the first-stage function
$\varphi(k_{it}, k_{i,t+1})$ is not a generic non-linear object, it is
exactly affine:
$$\varphi(k_{it}, k_{i,t+1}) \;=\; \beta_K\, k_{it} \,+\, \tfrac{1}{\gamma}\, k_{i,t+1} \,+\, \text{const}.$$

So a *linear* first stage (OLS of $\tilde v_{it}$ on $l_{it}$, $k_{it}$,
and $k_{i,t+1}$ ) identifies $\beta_L$ *and* $\beta_K$ in a single
regression, as the coefficients on $l_{it}$ and $k_{it}$ respectively.
The second stage is redundant.

Three features of the DGP combine to make $\varphi$ affine rather than
genuinely non-linear:

1.  **No adjustment costs for capital.** The capital Euler equation does
    not involve lagged capital as an additional state, so the policy
    function for $k_{i,t+1}$ is log-linear in $a_{it}$ rather than a
    non-trivial function of multiple states.

2.  **User cost** $U$ fixed across firms. No firm-level $u_{it}$ enters
    the capital rule, so $k_{i,t+1}$ depends on $a_{it}$ alone, not on
    the pair $(a_{it}, u_{it})$.

3.  $\phi = 0$. Wage innovations are i.i.d., so lagged $c_{i,t-1}$ does
    not forecast $w_{it}$ and therefore does not feed into the
    period-$t$ choice of $k_{i,t+1}$ via rational expectations of next
    period's wage.

Together these deliver $k_{i,t+1} = \kappa_0 + \gamma\, a_{it}$, exactly
invertible in $k_{i,t+1}$. Substituting back into the production
function yields the affine $\varphi$ above, with $\beta_K$ appearing
cleanly as the coefficient on $k_{it}$ (its direct structural role) and
$1/\gamma$ as the coefficient on $k_{i,t+1}$ (the inverse of the capital
rule's sensitivity to $a_{it}$).

**A sharper observation.** Not only is the second stage unnecessary, the
*standard* OP second stage is under-identified here. In the OP
formulation, one would regress
$y_{it} = \tilde v_{it} - \hat\beta_L l_{it}$ on $k_{it}$ and a flexible
function $h(k_{i,t-1}, k_{it})$ capturing $\rho\, a_{i,t-1}$. Under the
exact log-linear capital rule, $h$ collapses to a linear function of
$k_{it}$ (and $k_{i,t-1}$), and the OP second stage loses rank: it only
identifies the sum $\beta_K + \rho/\gamma = 1 - \beta_L$, not $\beta_K$
and $\rho$ separately. This is the Bond–Söderbom (2005)
non-identification result. The BB-style second stage used in Q1(b)
avoids the collapse by exploiting over-identifying moments from
$k_{i,t-2}$ and $y_{i,t-2}$, which is why it continues to work, but at
that point we are no longer using the OP second stage in any
recognisable form.

### Numerical confirmation

*Linear first stage:* Under this DGP varphi is affine, so no polynomial
terms are needed: a simple OLS of vt on (`l`, `k`, `k_{t+1}`) delivers
$\beta_L$ and $\beta_K$ (and $1/\gamma$ as a bonus) in one shot.

```{r q1c-linear-first-stage}
# Fit the linear first stage with just labour, capital, and next-period capital

  stage1_lin <- lm(vt ~ l + k + k_p1,
                   data = dat_q1,
                   na.action = na.exclude)

# Report the four coefficient estimates and their standard errors

  round(summary(stage1_lin)$coefficients, 4)
```

Expected against the calibration: the coefficient on `l`
$\approx \beta_L = 0.6$; the coefficient on `k` $\approx \beta_K = 0.3$
(read directly from Stage 1, matching the Q1(b) BB-GMM estimate); the
coefficient on `k_p1` $\approx 1/\gamma = 0.2$, since
$\gamma = \rho/(1-\beta_L-\beta_K) = 5$.

*Sanity check:* the coefficient on $k_{t+1}$ in the linear first stage
should numerically equal $1/\gamma$ computed from the structural
parameters. This confirms the affine decomposition of varphi.

```{r q1c-gamma-check}
# Calibration values, taken from the Setup section

  bL_cal  <- 0.6
  bK_cal  <- 0.3
  rho_cal <- 0.5

# gamma is the sensitivity of k_{t+1} to current TFP under the capital rule
  
  gamma_true <- rho_cal / (1 - bL_cal - bK_cal)   # = 0.5 / 0.1 = 5
  inv_gamma  <- 1 / gamma_true                     # = 0.2

# Side-by-side comparison
  
  c(`1/gamma (from calibration)` = inv_gamma,
    `k_p1 coefficient (OLS)`     = as.numeric(coef(stage1_lin)["k_p1"]))
```

*Final check*: in the polynomial first stage from Q1(b), the quadratic
and interaction terms should have coefficients numerically
indistinguishable from zero, i.e. varphi has no curvature in this DGP.

```{r q1c-polynomial-terms}
# Extract the estimated coefficients on k^2, k_{t+1}^2, and k * k_{t+1}
# from the polynomial first stage (`stage1` from Q1(b))

  round(summary(stage1)$coefficients[c("I(k^2)",
                                       "I(k_p1^2)",
                                       "I(k * k_p1)"), 1:2], 5)
```

If these three coefficients are (a) small in magnitude and (b)
statistically indistinguishable from zero, we have a direct numerical
confirmation that $\varphi$ is affine under this DGP , and hence that
the first-stage specification alone identifies both $\beta_L$ and
$\beta_K$ consistently, with no second stage required.

\newpage

# Question 2

The researcher substitutes $a_{it} = \rho\, a_{i,t-1} + \xi_{it}$ into
the production function and treats $a_{i,t-1}$ with a control function
$h(k_{it}, k_{i,t-1})$, giving the estimating equation

$$\tilde v_{it} \;=\; \beta_0 \,+\, \beta_L\, l_{it} \,+\, \beta_K\, k_{it} \,+\, \rho\, h(k_{it}, k_{i,t-1}) \,+\, \xi_{it} \,+\, \varepsilon_{it}, \tag{$\star$}$$

and estimates $(\beta_0, \beta_L, \beta_K)$ jointly by 2SLS, treating
$l_{it}$ as the single endogenous regressor and using $l_{i,t-1}$ as the
outside instrument.

## Q2(a) Why is an outside instrument needed for $l_{it}$?

Labour is chosen at $t$ after the firm has observed $a_{it}$ (and the
wage shock $w_{it}$). The log-labour FOC in the setup gives

$$l_{it} \;=\; \text{pow}\cdot\bigl[\log\beta_L \,+\, \beta_0 \,+\, \beta_K k_{it} \,+\, a_{it} \,-\, w_{it}\bigr], \qquad \text{pow} \equiv \frac{1}{1-\beta_L}.$$

Decomposing $a_{it} = \rho\, a_{i,t-1} + \xi_{it}$, this becomes

$$l_{it} \;=\; \text{pow}\cdot\bigl[\log\beta_L \,+\, \beta_0 \,+\, \beta_K k_{it} \,+\, \rho\, a_{i,t-1} \,-\, w_{it}\bigr] \,+\, \text{pow}\cdot \xi_{it}.$$
So $l_{it}$ is a linear function of the period-$t$ TFP innovation
$\xi_{it}$ with coefficient $\text{pow} > 0$.

In equation $(\star)$, the composite disturbance is
$\xi_{it} + \varepsilon_{it}$, and
$$\mathrm{Cov}(l_{it},\, \xi_{it} + \varepsilon_{it} \mid k_{it}, k_{i,t-1}) \;=\; \text{pow}\cdot \sigma_\xi^2 \;>\; 0.$$

OLS on $(\star)$ is therefore inconsistent for $\beta_L$ . Controlling
for $h(k_{it}, k_{i,t-1})$ does not fix the problem because $h$ only
absorbs $a_{i,t-1}$; the period-$t$ innovation $\xi_{it}$ remains in the
residual and is positively correlated with $l_{it}$. An outside
instrument, which is something correlated with $l_{it}$ but orthogonal
to $\xi_{it}$ and $\varepsilon_{it}$, is therefore needed to recover
$\beta_L$.

\newpage

## Q2(b) Is $l_{i,t-1}$ a good IV when $\phi = 0$?

**Valid but irrelevant. The 2SLS estimator is inconsistent for**
$\beta_L$

**1. Exogeneity**

$l_{i,t-1}$ is chosen at date $t-1$ and is therefore a function of the
information set
$\mathcal{I}_{i,t-1} = \{a_{i,t-1}, w_{i,t-1}, k_{i,t-1}\}$. Both
$\xi_{it}$ (a period-$t$ TFP innovation) and $\varepsilon_{it}$ (i.i.d.
measurement error) are independent of $\mathcal{I}_{i,t-1}$, so
$$\mathrm{Cov}\bigl(l_{i,t-1},\, \xi_{it} + \varepsilon_{it}\bigr) \;=\; 0.$$
The exclusion restriction holds.

**2. Relevance**

Take the labour FOCs at $t$ and $t-1$ and subtract off everything held
fixed by the 2SLS controls $(k_{it}, k_{i,t-1})$. Under the exact
log-linear capital rule, $k_{it} = \kappa_0 + \gamma\, a_{i,t-1}$, so
conditioning on $k_{it}$ already pins down $a_{i,t-1}$. Residual
variation in $l_{i,t-1}$ given the controls is therefore
$$l_{i,t-1} \,-\, \mathbb{E}[l_{i,t-1} \mid k_{it}, k_{i,t-1}] \;=\; -\,\text{pow}\cdot w_{i,t-1},$$
while residual variation in $l_{it}$ is

$$l_{it} \,-\, \mathbb{E}[l_{it} \mid k_{it}, k_{i,t-1}] \;=\; \text{pow}\cdot(\xi_{it} - w_{it})$$
So the relevant covariance is

$$\mathrm{Cov}(l_{i,t-1}, l_{it} \mid k_{it}, k_{i,t-1}) \;=\; \text{pow}^2\cdot \mathrm{Cov}(-w_{i,t-1},\, \xi_{it} - w_{it}) \;=\; 0,$$

because under $\phi = 0$ the wage is i.i.d. over time
($w_{i,t-1} \perp w_{it}$) and independent of TFP innovations
($w_{i,t-1} \perp \xi_{it}$). The rank condition fails: $l_{i,t-1}$ is
**irrelevant** given the controls, so 2SLS is inconsistent.

**Simulation**

Simulate under phi = 0 and run the researcher's 2SLS with $l_{t-1}$ as
the outside instrument for $l_t$.

```{r q2b-setup}
# ---------------------------------------------------------------------
# Controls: k_t and k_{t-1} (standing in for the linear collapse of
# beta_K k_t + rho h(k_t, k_{t-1}) under this DGP)

# Endogenous: l_t
# Outside IV: l_{t-1}

# Expectation: partial F-stat on l_{t-1} in the first stage should be
# small (close to 1); 2SLS point estimate of beta_L should be far from
# 0.6 with a large standard error
# ---------------------------------------------------------------------

# Add the lagged labour variable as the outside instrument
dat_q2b <- dat_q1 |> 
  mutate(l_m1 = lag(l))
```

```{r q2b-firststage}
# ---------------------------------------------------------------------
# First-stage OLS: regress l_t on the outside IV plus the controls

# Under phi = 0 the partial coefficient on l_m1 should be
# statistically indistinguishable from zero.
# ---------------------------------------------------------------------

# Fit the first-stage regression
  
  fs_q2b <- lm(l ~ l_m1 + k + k_m1,
               data = dat_q2b, na.action = na.exclude)

# Table 1: coefficient on the excluded instrument l_{t-1}

  kable(
    summary(fs_q2b)$coefficients["l_m1", , drop = FALSE],
    digits = 4,
    caption = "Q2(b) first-stage, coefficient on the outside instrument $l_{i,t-1}$"
  )

# Partial F-statistic on the single excluded instrument
# (with one IV this is simply the t-stat squared)
  
  fs_t_q2b <- summary(fs_q2b)$coefficients["l_m1", "t value"]
  fs_F_q2b <- fs_t_q2b^2

# Table 2: weak-IV diagnostic summary
    
  kable(
    data.frame(
      Quantity = c("Partial F-stat on $l_{i,t-1}$",
                   "Staiger-Stock critical value",
                   "Diagnosis"),
      Value    = c(sprintf("%.3f", fs_F_q2b),
                   "10.000",
                   ifelse(fs_F_q2b < 10, "Weak IV (F < 10)", "Strong IV"))
    ),
    caption = "Q2(b) first-stage weak-instrument diagnostic"
  )
```

```{r q2b-iv}
# ---------------------------------------------------------------------
# Expected: a point estimate of beta_L
# that is (a) far from the true 0.6 and/or (b) has a very wide
# confidence interval
# ---------------------------------------------------------------------
  
  iv_q2b <- ivreg(vt ~ l + k + k_m1 | l_m1 + k + k_m1, data = dat_q2b)
  iv_sum  <- summary(iv_q2b, diagnostics = TRUE)

# Table 1: 2SLS coefficient estimates

  kable(
    iv_sum$coefficients,
    digits = 4,
    caption = "Q2(b) 2SLS coefficient estimates"
  )

# Table 2: diagnostic tests: weak IV, Wu-Hausman endogeneity, Sargan over-id
# (Sargan is NA here because the system is exactly identified: 1 IV for 1
# endogenous variable)
  
  kable(
    iv_sum$diagnostics,
    digits = 4,
    caption = "Q2(b) 2SLS diagnostic tests"
  )
```

Compare the researcher's 2SLS estimate of beta_L to the truth and to the
consistent Q1(b) estimate.

```{r q2b-summary}
results_q2b <- data.frame(
  Estimator = c("Truth",
                "Q1(b) Stage 1 (valid)",
                "Q2 2SLS with l_{t-1} (weak IV)"),
  beta_L    = c(0.6,
                as.numeric(bL_hat),
                as.numeric(coef(iv_q2b)["l"]))
)

kable(results_q2b, digits = 4,
             caption = "Q2(b) Weak-IV failure under phi = 0")
```

The partial $F$ on $l_{i,t-1}$ is 0.358, well below the Staiger-Stock
threshold of 10. This lines up with the theoretical prediction that the
conditional correlation is zero in population. The 2SLS point estimate of
$\beta_L$ is 0.795 with a standard error of 0.747, so the $t$-statistic is
1.06 and the 95% confidence interval covers roughly $[-0.67, 2.26]$. The
estimator delivers a number but no information about $\beta_L$.

\newpage

## Q2(c) Is $l_{i,t-1}$ a good IV when $\phi \neq 0$?

**No. The IV is now relevant but invalid. The 2SLS estimator is
inconsistent for** $\beta_L$

**Relevance**

With $\phi \neq 0$ the wage follows an MA(1):
$w_{it} = w + c_{it} + \phi\, c_{i,t-1}$. Both $w_{i,t-1}$ and $w_{it}$
load on the shock $c_{i,t-1}$ . The former directly, the latter via the
moving-average coefficient, so
$$\mathrm{Cov}(w_{i,t-1}, w_{it}) \;=\; \phi\, \sigma_c^2 \;\neq\; 0.$$

From the labour FOC, $l_{it}$ and $l_{i,t-1}$ inherit this correlation
through the $-\text{pow}\cdot w$ term, and the correlation survives
conditioning on $(k_{it}, k_{i,t-1})$. The first-stage partial $F$
should now be large.

**Invalidity**

When $\phi \neq 0$, the period-$t$ choice of next-period capital depends
on the *expected* future wage,
$$\mathbb{E}[w_{i,t+1} \mid \mathcal{I}_{it}] \;=\; w \,+\, \phi\, c_{it},$$
so the capital rule picks up an extra state:
$$k_{i,t+1} \;=\; \kappa_0 \,+\, \gamma\, a_{it} \,+\, \delta\, c_{it}, \qquad \delta \;=\; -\phi\, \beta_L \cdot \text{pow} \,/\, (1 - \beta_L - \beta_K),$$
with $\delta < 0$ (a higher $c_{it}$ raises the expected wage next
period, depressing labour and therefore the marginal product of
capital). Lagged one period:
$k_{it} = \kappa_0 + \gamma\, a_{i,t-1} + \delta\, c_{i,t-1}$.

The researcher's control function $h(k_{it}, k_{i,t-1})$ is therefore
**misspecified**: one observed capital is a two-state object
($a_{i,t-1}$ and $c_{i,t-1}$), so no function of $(k_{it}, k_{i,t-1})$
alone can recover $a_{i,t-1}$. The best-linear-predictor error
$$\eta_{i,t-1} \;\equiv\; a_{i,t-1} \,-\, h(k_{it}, k_{i,t-1})$$ is not
identically zero, and it is correlated with $c_{i,t-1}$,
because $c_{i,t-1}$ is the piece of $k_{it}$ that the control function
misattributes to $a_{i,t-1}$. The residual in the researcher's equation
therefore contains $\rho\, \eta_{i,t-1}$, giving
$$\mathrm{Cov}(l_{i,t-1},\, \xi_{it} + \varepsilon_{it} + \rho\, \eta_{i,t-1}) \;=\; \rho\,\mathrm{Cov}(l_{i,t-1},\, \eta_{i,t-1}) \;\neq\; 0,$$
because $l_{i,t-1}$ depends on
$w_{i,t-1} = w + c_{i,t-1} + \phi\, c_{i,t-2}$ and so is correlated with
$c_{i,t-1}$. The exclusion restriction fails, so 2SLS is inconsistent.

**Simulation**

Simulate under phi != 0. Pick phi = 0.5 so that wages are moderately
MA(1) but not extreme. The rest of the DGP is unchanged

```{r q2c-setup}
# ---------------------------------------------------------------------
# Expectation:
#  - Partial F on l_{t-1} is LARGE (relevance is now satisfied)
#  - The 2SLS point estimate of beta_L is precisely estimated but
#    biased away from 0.6 (invalidity)
# ---------------------------------------------------------------------

# Simulate a fresh dataset under phi = 0.5
dat_q2c <- simulate_pf(phi = 0.5, urndvar = 0) |> 
  mutate(
    l_m1 = lag(l),   # outside instrument for l_t
    k_m1 = lag(k)    # included exogenous regressor
  )
```

```{r q2c-firststage}
# ---------------------------------------------------------------------
# First-stage OLS: regress l_t on the outside IV plus controls

# Unlike the phi = 0 case, the partial coefficient on l_m1 should be
# large and highly significant
# ---------------------------------------------------------------------

# Fit the first stage
  
  fs_q2c <- lm(l ~ l_m1 + k + k_m1,
               data = dat_q2c, na.action = na.exclude)

# Table 1: coefficient on the excluded instrument

  kable(
    summary(fs_q2c)$coefficients["l_m1", , drop = FALSE],
    digits = 4,
    caption = "Q2(c) first-stage: coefficient on $l_{i,t-1}$ under $\\phi = 0.5$"
  )

# Partial F-statistic on l_m1 (t-stat squared since 1 excluded IV)
  
  fs_t_q2c <- summary(fs_q2c)$coefficients["l_m1", "t value"]
  fs_F_q2c <- fs_t_q2c^2

# Table 2: weak-IV diagnostic summary

  kable(
    data.frame(
      Quantity = c("Partial F-stat on $l_{i,t-1}$",
                   "Staiger-Stock critical value",
                   "Diagnosis"),
      Value    = c(sprintf("%.3f", fs_F_q2c),
                   "10.000",
                   ifelse(fs_F_q2c < 10, "Weak IV (F < 10)", "Strong IV"))
    ),
    caption = "Q2(c) first-stage strength diagnostic"
  )
```

```{r q2c-iv}
# ---------------------------------------------------------------------
# Researcher's 2SLS under phi = 0.5

# Expected: a tight standard error on beta_L (because the IV is now relevant), 
# but a biased point estimate (because the IV is correlated with the residual 
# through c_{t-1})
# ---------------------------------------------------------------------

  iv_q2c  <- ivreg(vt ~ l + k + k_m1 | l_m1 + k + k_m1, data = dat_q2c)
  iv_sum_q2c <- summary(iv_q2c, diagnostics = TRUE)

# Table 1: 2SLS coefficient estimates

  kable(
    iv_sum_q2c$coefficients,
    digits = 4,
    caption = "Q2(c) 2SLS coefficient estimates"
  )

# Table 2: diagnostic tests (Sargan is NA, ie. system is exactly identified)
  
  kable(
    iv_sum_q2c$diagnostics,
    digits = 4,
    caption = "Q2(c) 2SLS diagnostic tests"
  )
```

Compare: true $\beta_L$, the $\phi = 0$ 2SLS (weak-IV failure from
Q2(b)), and the $\phi = 0.5$ 2SLS (invalid-IV failure in this part)

```{r q2c-summary}
  results_q2c <- data.frame(
    Scenario = c("Truth",
                 "Q2(b): phi = 0 (valid IV, weak)",
                 "Q2(c): phi = 0.5 (relevant IV, invalid)"),
    beta_L   = c(0.6,
                 as.numeric(coef(iv_q2b)["l"]),
                 as.numeric(coef(iv_q2c)["l"]))
)
  
kable(results_q2c, digits = 4,
      caption = "Q2 2SLS point estimates across scenarios")
```

The partial $F$ on $l_{i,t-1}$ is 679.9, comfortably above the Staiger-Stock
threshold, so the rank condition is restored when $\phi \neq 0$. But the
2SLS point estimate of $\beta_L$ is 0.0145, far from the true value of 0.6,
and the standard error of 0.034 makes the bias look highly significant.

So the researcher's approach fails in both wage regimes: weakly when $\phi = 0$,
and with precise bias when $\phi \neq 0$.

\newpage

## Q2(d) Can $\beta_K$ be identified even if $\phi = 0$ and $\beta_L$ is known?

**Answer: No. Even granting an oracle value of** $\beta_L$, the
researcher's specification cannot separately identify $\beta_K$ and
$\rho$. This is the Bond-Söderbom (2005) non-identification result.

### Derivation

Assume the best case for the researcher: $\phi = 0$, $U$ fixed, no
adjustment costs. So, the capital rule is exactly log-linear,
$$k_{it} \;=\; \kappa_0 \,+\, \gamma\, a_{i,t-1}, \qquad \gamma = \frac{\rho}{1-\beta_L-\beta_K}.$$
The control function for $a_{i,t-1}$ is then a *linear* function of
$k_{it}$ alone:
$$h(k_{it}, k_{i,t-1}) \;=\; a_{i,t-1} \;=\; \frac{k_{it} - \kappa_0}{\gamma}.$$
Substitute into the researcher's equation, subtract $\beta_L l_{it}$
from both sides (using the oracle $\beta_L$), and collect terms in
$k_{it}$:
$$\tilde v_{it} - \beta_L l_{it} \;=\; \underbrace{\bigl(\beta_0 - \tfrac{\rho \kappa_0}{\gamma}\bigr)}_{\text{constant}} \,+\, \underbrace{\left(\beta_K + \frac{\rho}{\gamma}\right)}_{\pi_K \,\equiv\, \text{reduced-form slope}} k_{it} \,+\, \xi_{it} \,+\, \varepsilon_{it}.$$

The reduced-form slope on $k_{it}$ simplifies dramatically using
$\rho/\gamma = 1 - \beta_L - \beta_K$:
$$\pi_K \;\equiv\; \beta_K + \frac{\rho}{\gamma} \;=\; \beta_K + (1 - \beta_L - \beta_K) \;=\; 1 - \beta_L.$$

### Non-identification

Once $\beta_L$ is known, $\pi_K = 1 - \beta_L = 0.4$ regardless of $\beta_K$
and $\rho$. The regression of $\tilde v_{it} - \beta_L l_{it}$ on $k_{it}$
therefore contains no structural information beyond $\beta_L$. Any pair
$(\beta_K, \rho)$ satisfying
$$\beta_K + \frac{\rho}{1 - \beta_L - \beta_K} \;=\; 1 - \beta_L$$
is observationally equivalent: one equation, two unknowns.

At our calibration: $\beta_K + \rho/\gamma = 0.3 + 0.5/5 = 0.4 = 1-\beta_L$.
A different DGP with $(\beta_K, \rho) = (0.2, 0.4)$ and the same $\beta_L$
gives $\gamma' = 2$ and $\pi_K' = 0.2 + 0.2 = 0.4$, an identical reduced
form indistinguishable from the data.

### Numerical confirmation

```{r q2d-demo}
# Form the cleaned dependent variable using the oracle beta_L

# Since k_t is chosen at t-1 it is orthogonal to (xi_t, eps_t), so
# OLS is consistent for the reduced-form slope

  dat_q2d <- dat_q1 |> 
    mutate(y_tilde = vt - bL_cal * l)   # bL_cal = 0.6, set in q1c-gamma-check

# Regress y_tilde on k_t and k_{t-1}.  The coefficient on k_t is pi_K;
# the coefficient on k_{t-1} should be ~0 because under this DGP h
# does not depend on k_{t-1}

  reg_q2d <- lm(y_tilde ~ k + k_m1,
                data = dat_q2d, na.action = na.exclude)
  
  kable(
    summary(reg_q2d)$coefficients,
    digits = 4,
    caption = "Q2(d) reduced-form regression of $\\tilde v_{it} - \\beta_L l_{it}$ on $(k_{it}, k_{i,t-1})$"
  )

# Compare the estimated pi_K to the theoretical 1 - beta_L = 0.4
  
  kable(
    data.frame(
      Quantity = c("$1 - \\beta_L$ (theoretical)",
                   "$\\beta_K + \\rho/\\gamma$ at calibration",
                   "Estimated slope on $k_t$"),
      Value    = c(1 - bL_cal,
                   bK_cal + rho_cal / gamma_true,
                   as.numeric(coef(reg_q2d)["k"]))
    ),
    digits = 4,
    caption = "Q2(d) the reduced-form slope pins down $1 - \\beta_L$, not $\\beta_K$"
  )
```

The three numbers in the second table should all be $\approx 0.4$ (the
pooled reduced-form coefficient is what the researcher recovers) and no
algebraic manipulation of it can separate $\beta_K$ from $\rho$. The
coefficient on $k_{i,t-1}$ should be numerically indistinguishable from
zero, which can that $h$ has collapsed to a function of $k_{it}$ alone.

### Why Q1(b)'s BB-GMM escapes the collapse

The two-stage procedure from Q1(b) avoids this trap because it uses
**over-identifying moments** from $k_{i,t-2}$ and $y_{i,t-2}$ that the
researcher's specification discards. Under the AR(1) structure, these
twice-lagged variables deliver additional restrictions of the form
$\mathbb{E}[k_{i,t-2}(\xi_{it} + \varepsilon_{it} - \rho\varepsilon_{i,t-1})] = 0$
and similarly for $y_{i,t-2}$, which break the reduced-form collinearity
and separately identify $\beta_K$ and $\rho$. The researcher's 2SLS,
which conditions only on $(k_{it}, k_{i,t-1})$ and uses $l_{i,t-1}$ as
the single excluded instrument, has no analogous source of extra
restrictions, which is exactly why it fails here and why Q1(b)'s dynamic
panel GMM does not.
