---
title: "Advanced Empirical Research Methods Hilary 2025/26"
author: 'Candidate ID: 1091806'
date: "2026-04-13"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
    toc_depth: 2
    number_sections: false
  html_document:
    toc: true
    toc_depth: '2'
    df_print: paged
subtitle: 'Assessment Exercise: Production Functions'
fontsize: 11pt
geometry: margin=1in
header-includes:
- \usepackage{amsmath}
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{amsmath}
- \usepackage{newunicodechar}
- \newunicodechar{ρ}{$\rho$}
- \newunicodechar{π}{$\pi$}
- \newunicodechar{ξ}{$\xi$}
- \newunicodechar{ϕ}{$\phi$}
---

\newpage
# Set up

## Model

We consider a two-factor Cobb-Douglas value-added production function in logs:

$$v_{it} \;=\; \beta_0 \,+\, \beta_L \, l_{it} \,+\, \beta_K \, k_{it} \,+\, a_{it}, \qquad \beta_L + \beta_K < 1$$

The price of value-added is normalised to one. Labour $L_{it}$ is perfectly flexible
and chosen in period $t$ after the firm observes total factor productivity
$A_{it} = \exp(a_{it})$ and the wage $W_{it}$. Capital $K_{it}$ is predetermined and
chosen in period $t-1$ to maximise expected period-$t$ profit. There are no
adjustment costs for either input, and the underlying gross-output technology is
Leontief in value-added and a flexible intermediate input.

## Stochastic processes

Total factor productivity follows an AR(1):
$$a_{it} \;=\; \rho \, a_{i,t-1} \,+\, \xi_{it}, \qquad \xi_{it} \sim \mathrm{i.i.d.}\, N(0, \sigma_\xi^2), \qquad 0 < \rho < 1.$$

The log wage follows an MA(1):
$$w_{it} \;=\; w \,+\, c_{it} \,+\, \phi \, c_{i,t-1}, \qquad c_{it} \sim \mathrm{i.i.d.}\, N(0, \sigma_c^2).$$

The rental rate of capital $U$ is **fixed and common across firms and time**.

The observed log of value-added is contaminated with classical measurement error:
$$\tilde v_{it} \;=\; v_{it} \,+\, \varepsilon_{it}, \qquad \varepsilon_{it} \sim \mathrm{i.i.d.}\, N(0, \sigma_\varepsilon^2).$$

## Information sets

The firm's information set at $t-1$ includes $a_{i,t-1}$, $w_{i,t-1}$, $U$, and the
fixed value-added price; it excludes $a_{it}$ and $w_{it}$. So $K_{it}$ is
uncorrelated with $\xi_{it}$ by construction, while $L_{it}$ — chosen after
$\xi_{it}$ is realised — is not.

## Observed data

Only the input and output quantities $\bigl(\tilde V_{it} = e^{\tilde v_{it}},\, L_{it},\, K_{it}\bigr)$ are observed.
Wages, the rental rate, intermediate inputs, investment, and TFP are all unobserved.
This rules out the GNR labour-share estimator (no $W$), the LP/ACF intermediate-input
proxy (no $M$), the OP investment proxy (no $I$), and the wage instrument (no $W$).

## Differences from the practice exercise

Two assumptions on the DGP differ from `prod_sim_exercise.R` and matter for
identification:

1. The rental rate $U$ has **no firm-level variation** (we set `urndvar = 0`).
2. Question 1 imposes $\phi = 0$, so the log wage is i.i.d.; Question 2(c)
   re-runs the analysis with $\phi \ne 0$.

These two restrictions together imply that capital is an *exact log-linear*
function of past TFP — a key fact we use throughout.

## Calibration for the simulations

We follow the practice-script parameterisation but set the two switches above:

| Parameter                       | Symbol            | Value                       |
|---------------------------------|-------------------|-----------------------------|
| Output elasticity of labour     | $\beta_L$         | 0.6                         |
| Output elasticity of capital    | $\beta_K$         | 0.3                         |
| AR(1) persistence of TFP        | $\rho$            | 0.5                         |
| Variance of TFP innovation      | $\sigma_\xi^2$    | 0.05                        |
| Mean log wage                   | $w$               | $\log(1) = 0$               |
| Variance of wage innovation     | $\sigma_c^2$      | 0.05                        |
| Log rental rate (fixed)         | $u$               | $\log(0.1)$                 |
| Variance of measurement error   | $\sigma_\varepsilon^2$ | 0.1                    |
| MA(1) wage parameter            | $\phi$            | 0 (Q1, Q2b, Q2d); 1 (Q2c)   |
| Number of periods               | $T$               | 100,001                     |

## R Set-up

```{r r-setup, message=FALSE, warning=FALSE}
# Load libraries
  library(dplyr)    # data manipulation
  library(psych)    # geometric.mean()
  library(AER)      # ivreg(), linearHypothesis()
  library(gmm)      # gmm()

# Set seed
  set.seed(1234)
```


## Reusable simulation function

To keep the rest of the document concise, we wrap the DGP in a single function.
It takes the two switches we toggle across questions ($\phi$ and `urndvar`) plus
the standard parameters, and returns a data frame with the variables we will use:
$\tilde v$, $l$, $k$, plus their lags and (for diagnostic purposes) the latent
$a$ and $w$.

```{r sim-function}
simulate_pf <- function(obs   = 100001,
                        bL    = 0.6,
                        bK    = 0.3,
                        rho   = 0.5,
                        shkvar = 0.05,
                        wfix   = log(1),
                        wrndvar = 0.05,
                        phi    = 0,
                        ufix   = log(0.1),
                        urndvar = 0,
                        merrvar = 0.1) {

  # AR(1) TFP
  shk  <- rnorm(obs, mean = 0, sd = sqrt(shkvar))
  stat <- rnorm(obs, mean = 0, sd = sqrt(shkvar / (1 - rho^2)))
  a    <- stat
  for (i in 2:obs) a[i] <- rho * a[i - 1] + shk[i]

  # Rental rate (fixed if urndvar = 0)
  u <- rnorm(obs, mean = ufix, sd = sqrt(urndvar))

  # MA(1) log wage
  wrnd <- rnorm(obs, mean = 0, sd = sqrt(wrndvar))
  w    <- wfix + wrnd + phi * lag(wrnd)

  # Capital decision rule (predetermined)
  rat     <- bK / (1 - bL)
  fac     <- 1 / (1 - rat)
  pow     <- 1 / (1 - bL)
  tau     <- bL^(bL * pow) - bL^pow
  logetau <- log(tau) - bL * pow * wfix - phi * bL * pow * lag(wrnd) +
             ((bL * pow)^2) * wrndvar / 2
  k <- fac * (log(rat) + logetau - u + pow * rho * lag(a) +
              0.5 * shkvar * pow^2)

  # True log value-added (without measurement error)
  v_true <- bL * pow * (log(bL) - w) + rat * k + pow * a

  # Flexible labour decision rule
  l <- pow * (log(bL) - w + bK * k + a)

  # Add measurement error to obtain observed log value-added
  merr <- rnorm(obs, mean = 0, sd = sqrt(merrvar))
  vt   <- v_true + merr

  data.frame(t = seq_len(obs), vt = vt, l = l, k = k,
             a = a, w = w, v_true = v_true)
}
```

A quick sanity check that the simulation matches the assumed DGP — the
infeasible OLS regression of $\tilde v$ on $l$, $k$, and the true $a$ should
recover $(\beta_0, \beta_L, \beta_K)$ up to sampling error, while feasible OLS
omitting $a$ should not.

```{r sim-check}
dat0 <- simulate_pf(phi = 0)

# Infeasible OLS — should give bL = 0.6, bK = 0.3
summary(lm(vt ~ l + k + a, data = dat0))$coefficients

# Feasible OLS omitting a — biased (transmission bias)
summary(lm(vt ~ l + k,    data = dat0))$coefficients
```

If the infeasible OLS coefficients land near $(\beta_L, \beta_K) = (0.6, 0.3)$
and the feasible OLS does not, the DGP is correctly implemented.

\newpage
# Question 1 Assume ϕ=0, 

## Q1 (a) Two-stage procedure




\newpage
## Q1 (b)

\newpage
## Q1 (c)



\newpage
# Question 2

## Q2 (a)

\newpage
## Q2 (b)

\newpage
## Q2 (c)







